\part{What About Real World Cases?}
\label{ch.modnine}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}
\input{statusofabortion}
\chapter{Part 20: What is an Abortion?}

This question is the start of a very heated debate which can get nasty, so please keep that in mind. I have used other heated examples before, so you should know how to handle this. For this chunk, we are just looking at what it is, not the moral status of it. Here, we are looking at the metaphysical question concerning abortion, what is it? Later, once this is settled, we will look at the ethical question, is it OK to have one? 

One essential feature to an abortion, it would seem, is that there needs to be the ending of a pregnancy. But, that certainly can't be it. Take this case as an example:

\begin{earg}
    \item[] Birth is the ending of a pregnancy. 
    \item[] By our definition, the ending of a pregnancy is an abortion. 
    \item[] Therefore, birth is an abortion.
\end{earg}

So, we could add in something about the pre-mature nature of the termination, but that would make pre-mature births abortions, which also seems just as wrong (as in misfitting). Glossing over some of the more graphic examples I could give, the core, missing feature which makes an act an abortion and not birth or some crime against another person seems to be that it needs to be voluntary. The woman needs to, with informed consent, want to terminate the pregnancy early, without resulting in a child. There can be interesting cases, worth thinking about, where the woman gives consent, but not informed consent (she may have been misinformed about what exactly it entails, which might make her not want it).   For this module, we will be defining an abortion as one of these two things (could be both, but that's a weird case), these two features fit for both the pro-choice side and the pro-life side of the debate, we will be covering both:

\begin{enumerate}
\item A woman voluntarily terminating her own pregnancy.  	
\item A woman allowing another to terminate her pregnancy (referring to the subject).
\end{enumerate}
If you only have the first one, then you will not get cases of, say, doctor assisting the woman in terminating her pregnancy. If you only have the second one, then you will not get cases of self-administered abortions.  To avoid the cases where birth could be defined as an abortion, we need to say that `terminating a pregnancy' does involve the ending of a fetus. The moral status of that fetus is where the debate is.  

\subsection{Is it morally permissible to have an abortion?}

I get that this is a hot issue, and if I have not already, I guess that I will need to be far more active in the comments in the discussion for this one than I already have been, please remember to be civil. For Warren, the moral status of abortion hinges on the answer to the following question:

\begin{center}
The Fetus Question: Is a fetus a person, in the morally relevant sense?
\end{center}

The main tie-in, and one which you will read me say several times, is that if a fetus is a person, then abortion is wrong, if a fetus is not a person, then abortion is permissible. The Fetus Question moves the ethical debate regarding abortion, ``is abortion murder?" to a metaphysical debate regarding person-hood. Typically, our moral intuitions are gut responses, which come from mental shortcuts, when we analyse that shortcut, we can get down to a metaphysical question which we have grounds to prove or disprove. The author of the reading is arguing that it is possible to show that a fetus is not a person, making abortion permissible.\footnote{Some students, especially those who use translation dictionaries, will have issues with two different words, these words are ``person" and ``human". These do not mean the same thing, and we will see how these come apart later. For now, a human is a member of our species and a person is a being with moral worth equal to you or me. We will see later how there can be non-human persons.} 

\subsection{Framing the Problem}

Of course, while some philosophers and others would deny the possibility of a proof that a fetus is not a person, claiming that to do so would be to prove a contradiction or that it's not possible to prove either way (this would lead to skepticism about fetal person-hood). By the same token, others will claim that there's no need for a proof. These people claim that the moral status of a fetus, its person-hood, is too obvious to need justification. But, both sides of the debate, pro-choice and the pro-life, take their evidence and reasoning to be obvious, to an equal degree. This is much like a belief in God. Some Atheists claim that the non-existence of God is clear and obvious, while at the same time, Theists claim that the existence of God is equally obvious. This disagreement means that we can't trust our gut instincts on this issue, we need to use logic to show that one of the sides is faulty.

Through this module, we will see the best arguments on both sides of the abortion debate. But, to start us off, we will look at the commonplace, pro-choice arguments. Though Warren agrees with their conclusions, she disagrees with how they got there. There are glaring issues in their reasoning. 


\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
Argument A&Argument B\\
    \begin{earg}
    \item[1] Restrictive Laws regarding abortion cause more harm than the lack of those laws.
    \item[2] Causing more harm than otherwise is always wrong.
    \item[3] Therefore, restrictive laws regarding abortions are wrong.
\end{earg}&\begin{earg}
    \item[1] Restrictive laws regarding abortion deny women the ability to control their reproduction.
    \item[2] Denying women the right to control their reproduction violates their right to control their body.
    \item[3] Violating their right to control their body is always wrong.
    \item[4] Therefore, restrictive laws regarding abortions are wrong.
\end{earg}\\
\end{tabular}
Pro-choicers (I made that term up) have never adequately come to grips with the conceptual issues surrounding abortion. As a result, their arguments miss the mark when they try to attack the pro-life side. Their arguments avoid the fight rather than engaging in it. You can think of it as the pro-lifers are in a castle, and the pro-choicers are attacking it, but their catapults always miss.  Most, if not all, of the arguments which they give in favor of legal abortions fail to refute or even weaken the traditional pro-life argument. This is that a fetus is a human being and, therefore, abortion is murder.

The pro-choice arguments tend to fall into two different categories. First, the arguments use consequentialist/utilitarian style reasoning to show that having abortions be illegal is wrong. This is exemplified by Argument A. For example, these arguments point to the terrible results of having the restrictive laws. These include things like the deaths caused by unsafe abortions, the fact that they unfairly result in harder hardship on poorer women, the fact that the lack of access results in emotional hardship, and so on. But, the pro-life side has an easy reply to this. They can claim that the tragic side effects don't, by themselves, show that the laws aren't justified (they can still be justified even with the results). This is were we can get these argument: 


\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
The Pro-Life Response to Argument A:&The Pro-Life Response to Argument B:\\
\begin{earg}
    \item[1 ] A fetus is a being worthy of our moral consideration, same as you or me.
    \item[2 ] If a fetus is a being worthy of our moral consideration, then abortion is murder.
    \item[3 ] Murder is always wrong (regardless of the consequences).
    \item[4 ] So, from (1) and (2), abortion is murder.
    \item[5 ] Therefore, from (3) and (4), abortion is always wrong (regardless of the consequences). 
\end{earg}&
\begin{earg}
    \item[1 ] Violating a person's rights is always wrong.
    \item[2 ] A person's rights extend so far as they do not violate another's stronger right.
    \item[3 ] A person's right to life is stronger than another person's bodily rights.
    \item[4 ] Having an abortion (a claimed case of bodily right) is violating a fetus' right to life (as a fetus is a person).
    \item[5 ] Therefore, having an abortion is always wrong.
\end{earg}
\end{tabular}

The Pro-Life Response to Argument A , in their eyes, takes out the reasoning for the second line of Argument A. This is basically showing that there are some cases where the moral status of some behavior is not determined by the consequences. This can be supported by the very definition of murder, which is wrongful intentional killing. This kind of argument falls into the non-consequentialist style thinking and the basis there is that something are wrong regardless of the consequences. This will show up several times, the pro-life side of the debate tends to give reasons rooted in non-consequentialist style thinking. There is another pro-choice reply to this, which debates the idea that the abortion is murder in this case, further deepening the divide between the intuitions (because the reasonable response is more strongly consequentialist) 

The second argument given by the pro-choicers is exemplified by Argument B. This one uses a more non-consequentialist/Kantian style reasoning, pointing out that denying a woman the ability is have an abortion is to deprive her of some manner of bodily right. For example, the right to choose when and how one bears young. This one also falls short. The Pro-Life Response to Argument B shows where this falls short. They are basically saying there that while a person has rights, those rights can't imply that it's OK to violate another's rights. For example, take property rights. It seems clear that a person has the right to remove another from their property and has the right to defend their property, but how far does that right extend? Take this example, which really did happen, but I have exaggerated for our purposes:

\thoughtex{Piano Wire on the Trail}{A man owns a large piece of property, land, and there's a group of young hooligans who ride their motor bikes on the trails through his land. So, one day, he put up a line of piano wire across one of the trails, at a particular height. As one of the young people rides through the trail, the wire, tight, catches them on the neck and decapitates them.}{pianowire.jpg}

This was wrong of him, or so many have argued, and the pro-lifers can use the sort of reasoning here to support themselves. In putting up the piano wire and killing the hooligan, the man violated that person's right to life. This right is stronger than the property rights which the man would have otherwise had, meaning that in this case, he did not have the right to protect his property in this way.  From this line of thought, the pro-lifer can say that, because a person's right to life is stronger than a person's bodily rights, abortion is still wrong.  Since we have these strong competing intuitions and because the pro-lifer, the serious ones at least, won't give an inch for the consequentialist considerations, we are going to need to approach the abortion debate from the non-consequentialist perspective and show that the system, in fact, allows for abortion.
\subsection{The Fundamental Question}

The most basic question which we need to answer is not about the results of having legal abortions but rather about what it takes to be a person. This is the Fetus Question. The pro-life responses to the commonplace pro-choice arguments show that they take a fetus to be a person. So, there are two routes we can take. First, we could show that there are some cases such that, even if we assume that a fetus is a person, abortion is still permissible (this is the first part of the pro-choice response). Second, we could show that a fetus is not a person, thereby making abortion permissible (before a certain point). So, for that second half, we need to show what features it takes to be a person. If a fetus has the features, then it is a person and abortion is murder, if it does not, then it is not. Warren's case is that there are certain cases where it's permissible to have an abortion, even if a fetus is a person, and then she moves on to show that a fetus is not a person, which entails that abortion is permissible.

\chapter{Part 21: The Abortion Debate (Pro-Choice)}
\section{Assume That a Fetus is a Person}
As I mentioned at the end of the previous page, Warren has two parts to her paper. First, she will assume, just for the sake of argument, that a fetus is a person, and then from that show that if a fetus is a person, there are cases where killing it is permissible. The second section is where she shows that a fetus is not a person, so it's not entitled to the same moral rights as you or me, which means that abortion is permissible. This page is the start of that first section. If there's a way to get that at least some abortions are OK in this case, then you can't have the all out ban on them which is some times proposed. Rather, morally speaking, you would need to have some exceptions. It is worth noting, and we will return to this point, that the morality of an abortion, proved in this section, assuming that a fetus is a person, is limited to a very select range of cases. This range of cases is defined, roughly, by how much consent the woman had in the actions resulting in pregnancy.  This is where Warren, through Judith Thomson, gets the Violinist Thought Experiment:\autocite{Thomson1}

    \thoughtex{The Violinist}{Imagine that you have been kidnapped, and your bloodstream hooked up to that of the violinist, who happens to have an ailment that will certainly kill him unless he is permitted to share your kidneys for a period of nine months. You are a human dialysis machine. No one else can save him, since you alone have the right type of blood. He will be unconscious all that time, and you will have to stay in bed with him, but after the nine months are over he may be unplugged, completely cured, that is provided that you have cooperated. The violinist themselves had no knowledge that this would happen to them.}{violinist.jpg}

A common point stated about this thought experiment is that it says that you are stuck in bed for the 9-months. While this is not true in most cases of pregnancy (as in the woman can move around), there are plenty of cases of pregnancy where this is the case (as in they are stuck in bed for most of it), especially if the woman is quite small (my mom, for example, is 4' 10'') and the father is quite large (my father, for example, is 6' 2'').  But, moving on,  if the person on the pro-life side of this debate is consistent in their beliefs, if they don't have any contradictions in their reasoning, then they will need to say that you would need to go to term and be there for the full 9-months.  Despite you being forced into the situation, you will need to keep the violinist alive. They come to this from the following reasoning, which is much like their response to Argument B in the previous part:


\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
The Pro-Life Response to Argument B:&The Pro-Life Case to Stay Plugged In:\\
\begin{earg}
    \item[0 ] (hidden line) A fetus is a person.
    \item[1 ] Violating a person's rights is always wrong.
    \item[2 ] A person's rights extend so far as they do not violate another's stronger right.
    \item[3 ] A person's right to life is stronger than another person's bodily rights.
    \item[4 ] Having an abortion (a claimed case of bodily right) is violating a fetus' right to life (as a fetus is a person).
    \item[5 ] Therefore, having an abortion is always wrong.
\end{earg}&\begin{earg}
    \item[0 ] (hidden line) A violinist is a person.
    \item[1 ] Violating a person's rights is always wrong.
    \item[2 ] A person's rights extend so far as they do not violate another's stronger right.
    \item[3 ] A person's right to life is stronger than another person's bodily rights.
    \item[4 ] Unplugging from the violinist is violating the violinist's right to life.
    \item[5 ] Therefore, unplugging from the violinist is always wrong.
\end{earg}\end{tabular}
This, as we have seen before, is very strongly non-consequentialist style thinking. But, the vast majority of people would think that it's outrageous to think that there's the moral obligation here to keep the violinist alive. The claim here, roughly, boils down to the idea that the rights of another can't force a person to go above and beyond, take extreme measures, to ensure it. The violinist case shows that there are some cases where a person's bodily rights are stronger than a person's right to life. If there are cases like this for pregnancy, then the pro-lifer, morally, can't hold their position absolutely. It's really good of a person to agree to take on such a sacrifice, especially it was thrust upon them like this, but it seems wrong to say that your refusal is murder. Though he certainly has the right to life, something about this case must be off, removing the obligation. His right to life, in this case, does not force you, morally speaking, to to keep him alive by what ever means necessary; nor does it justify anyone else forcing you to do so. A law that required you to stay in bed with the violinist would clearly be an unjust law, since it is no proper function of the law to force unwilling people to make huge sacrifice for the sake of other people toward whom they have no such prior obligation. The key feature, for the Violinist case, is that you did not give informed consent to be plugged into the violinist. 
\subsection{What does this case get us? }

Well, it does get us something to get started on. There are a few similarities and differences between this case and pregnancies. First, we have a person (assuming that a fetus is a person) who is dependent on another for survival. In both cases, if the aware party does certain actions, then the other will die. The other aspect is that the dependency causes a drain on the aware party. In this case, there's a sense in which the other does not have a moral obligation to keep them alive. But, what removed that obligation? Some would say, as Warren does, that the key feature which negates the obligation is the kidnapping aspect. The person did not knowingly and voluntarily enter into this arrangement. They did not consent to taking on the risk.  If a woman does not enter into this arrangement willingly or without knowing the risks (without informed consent), then her situation is sufficiently similar to the violinist's case.

\subsection{The Results of the Violinist Case and the Problems}
The Violinist Thought-Experiment is initially quite plausible. It gives us a grounding to have that we aren't always obligated to keep people alive by any means necessary. If there are cases where a pregnancy is sufficiently like this case, then we can get the permissibility of abortion in those cases. But, for cases where they aren't relevantly similar to the Violinist case, we don't get the windy-side of morality, necessarily. The only real cases which are sufficiently like the violinist case are cases of pregnancy due to rape. In those cases, the woman did not voluntarily take on the risks. But, there could be some vagueness on how much give the Violinist Case gives us. If we extend it too far from the bounds of the Violinist case, we could run into wrinkles. For example, take this case: 

    \thoughtex{A Trip to Europe}{A woman, who is 7-months pregnant, finds herself unable to travel to Europe because of the pregnancy. She really does not want to postpone the trip. But, if she has an abortion, then the trip will go off without a hitch. Is it permissible for her to have an abortion?}{europetrip.jpg}

There seems to be a relevant difference between this case and the violinist case. For this one, people will often make a few different claims. First, some will claim that the woman is too far into the game to quit now, saying that the time for the abortion has passed. Others might claim that the trip to Europe is not a good enough reason to want an abortion, the case just doesn't make her bodily right strong enough. And others still will say that (assuming this is not a case of rape) that she entered into this knowing the risk and has the obligation.

In the case of pregnancies not caused by rape, there are other things which the woman could have done to prevent her In other cases, there are some things which the woman could have done. These are Warren's examples, so if they aren't correct or in some way off, be mad at her.  First, she could have remained chaste. In other words, she could have denied her partner the relations. If the partner acted anyway, this would be rape and fall into the violinist case. Remember, informed consent is absolutely key. Her second option, if she chooses to have sex, is to have taken her pills more faithfully. I know from the life experiences of friends, family, and former students, that this is hardly a 100\% sure-fire way to prevent pregnancy as it's often believed to be.  Personally, I am in favor of the development of the male-birth control, as it's better to take the bullets out of the gun than put on a bullet proof vest. The third option, if all else fails, is for her to abstain on dangerous days. But this option, also, is not reliable as some might claim. I encourage all people to research these options, but only get your research from non-religious, non-abstinence only, scientific sources.  

\subsubsection{The Pro-Life Response}

Consequently, there is room for the antiabortionist to argue that in the normal case of unwanted pregnancy a woman has, by her own actions, assumed responsibility of the fetus.
\factoidbox{If x behaves in a way which he could have avoided, and which he knows involves a 1\% chance of bringing into existence a human being, with a right to life, and does so knowing that if this should happen then that human being will perish unless x does certain things to keep him alive, then, when it does happen, x is not free of any obligation to what he knew in advance would he required to keep that human being alive.}

 To make this into a case, something which we can imagine and use for the analogy, I have created this thought-experiment, based on the Violinist Case, Violinist Cult Thought Experiment:

    \thoughtex{The Violinist Cult}{Suppose that you are a member of a cult along with 99 other people. I know that cult has a negative stigma to it, but bear with me. All of you have voluntarily and with full reasonable consent, entered into a lottery where one of you will be chosen at random to take on the role of being this violinist's human dialysis machine. Imagine that your name is drawn and you have the violinist hooked up. What's your obligation like now?}{violinistcult.jpg}

Most people, from my experience, say that in this case, there is the obligation to keep the violinist alive. You signed up knowing the risks and you lost the lottery, so to speak. So, what's the difference between the violinist case and the violinist cult case? The first seems to remove the obligation, but the second seems to have it. 

\subsubsection{Restricting the Outcome}

The plausibility of such an argument is enough to show that the Violinist analogy can provide a clear and persuasive defense of a woman’s right to obtain an abortion only when the woman is in no way responsible for her pregnancy. In all other cases, we would almost certainly conclude that it was necessary to look carefully at the particular circumstances in order to determine the extent of the woman’s responsibility and hence the extent of her obligation.

\section{Prove That a Fetus is Not a Person}
\begin{center}Is a fetus a person?\end{center}

As I have mentioned before, the second section of this paper concerns whether or not a fetus is a person. The point of the previous section was to show that there are some cases where abortion is OK, even if a fetus is a person. The point here is to show that a fetus is not a person, which means that abortion isn't murder, and therefore is not wrong. This is where the Fetus Question comes in very strong, this is why I also noted that we need to distinguish between `person' and `human'. Questions regarding personhood are metaphysical questions, does a thing have certain features? Similarly, questions regarding `human-hood' are metaphysical questions. From this, as I have mentioned, we are able to move from an ethical question to a metaphysical one. So, let's look at the standard, non-religious, argument from the pro-life side and another argument, which doesn't look similar, but I will explain how these relate:


\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
The Standard Pro-Life Argument&The Cheese Sandwich Fallacy\\
\begin{earg}
    \item[] It is wrong to kill innocent human beings
    \item[] Fetuses are innocent human beings
    \item[] Therefore, it is wrong to kill fetuses
\end{earg}&
\begin{earg}
    \item[] Nothing is better than God.
    \item[] A cheese sandwich is better than nothing.
    \item[] Therefore, a cheese sandwich is better than God. 
\end{earg}
\end{tabular}
These two arguments might look completely different, but both of them fall into the same logical fallacy, equivocation. Remember, I mentioned that there's a distinction between `human' and `person'. With that in mind, it becomes clear that there's something wrong with the Standard Pro-Life Argument. Glossing over that distinction leads to the equivocation. The Cheese Sandwich Fallacy is a great example of this fallacy. An equivocation is where a person uses the same word in two different ways in an argument, this is meant to mislead the reader.  The Standard Pro-Life Argument has the same error as this one. Now, we aren't the Dark Brotherhood, so the equivocation is not in the word `innocent'. Rather, the phrase ``human being" is being used in two different ways.

In the sentence ``fetuses are innocent human beings", the term `human being' is being used to talk about a member of our species, and up until this point, I have been very consistent in the use of the term `human' to talk about members of our species, things with the same sort of genetic make up as us. If we use this species interpretation of human, which has a long and solid history, we see that this is correct, fetuses are humans.

The other time we encounter this term is in the first line ``it's wrong to kill innocent human beings." If we use the case we use the genetic, or species, interpretation of the phrase `human being', we can quickly find cases where this is wrong, even by the non-consequentialist's lights. For example, if a brain dead human has a living will saying that they should pull the plug after a few days. But, if this was said without any context, we would likely accept it, so what makes it different? Well, the use of `human being' in this sentence, normally, means, in the most generous interpretation, where the line makes sense,  “a full-fledged member of the moral community.” We may call this the moral sense of “human” and I have been using the term ‘person’ to demark this.\footnote{There's a certain idiosyncratic grammar which I have the habit of using in the case of the word `person'. In English, there are, it would seem, two acceptable plurals, `people' and `persons'. How I use them, `people' refers to a collection of persons and `persons' refers to beings with person-hood individually. `People' is a collective or group sense of the plural and `persons' is a more individualistic sense.} In general, even when we are dealing with certain interesting legal cases, this distinction between `human' and `person' is overlooked. For example, a person has rights, but a human may or may not have rights. Though I don't like this example personally, but corporate person-hood is an example of this. We have a non-human entity, a corporation, seen as a person. Now, I would argue that `corporations are people' is a legal fiction, it's not actual. But, it's certainly possible, as we will see later, that there are actual, non-fictitious, non-human persons. Having a clear distinction between these in our everyday speech makes many thorny moral questions disappear, or, at the very least, makes them more intelligible. If we remove the equivocations in both of the arguments we get the following:


\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
The Standard Pro-Life Argument&The Cheese Sandwich Fallacy\\
\begin{earg}
    \item[] It is wrong to kill innocent persons
    \item[] Fetuses are innocent human beings
    \item[] Therefore, it is wrong to kill fetuses
\end{earg}&
\begin{earg}
    \item[] No existing thing is better than God.
    \item[] A cheese sandwich is better than not having anything at all.
    \item[] Therefore, a cheese sandwich is better than God. 
\end{earg}
\end{tabular}

As you can see, this doesn’t work. But, if we have that all humans are persons, as in that being a generic human is enough to be a person, the argument would work. This would be to say that all humans are persons. On the other hand, however, if there are cases where a human is not a person, then the argument fails. If we can show that all fetuses (before a certain point of development) and not persons, then we have that abortion is permissible (before a certain point in development), by the pro-life style reasoning. Showing that no fetus (before that point) is a person is Warren's next step.  

\section{Warren's Criteria for Personhood}
Warren argues that there are 5 features which are needed to be a person, and if a fetus has these features (after a certain point), then abortion would be murder (after that point). These features are listed here:

\begin{enumerate}
\item Consciousness 	
\item Reasoning
\item Self-Motivated Activity 
\item Communication
\item Self-Awareness
\end{enumerate}

\subsection{Feature 1: Consciousness}

The first of these features seem to be the most intuitive. This is consciousness. We have touched on this before, in the Mind-Body Problem module. Although there is much debate about some of the features of consciousness, we do have a general understanding of when it's had and when it's not. For example, does it react to external stimulus? Is there some behavioral or other kind of evidence that shows that this thing is thinking, has an internal life? 

\subsubsection{Is this thing conscious?}

There are a few tests to tell whether a being is conscious. The relevant one here awareness (as in reaction to external stimulus) and evidence of internal thoughts. Many other tests include the self-awareness, which is not necessarily the same thing, but that one is another feature. We will limited this test to merely something like ``does it feel pain?". If it reacts, it's conscious.

\subsubsection{Are we the only conscious things?}

There are several creatures which have consciousness (by this definition) aside from humans. It is worth noting also that not all humans are conscious. Consciousness in humans is only there when the human is developed beyond a certain point and without certain impairments. In fact, most animals do have this and so do most fish. It's possible for some plants to even have this, though that is easily debated against.

\subsubsection{Why does this matter?}

If Warren is correct, several beings, including some humans, are immediately excluded from person-hood. Plants are (more than likely) taken out of the moral community, severely disabled humans, clams, and some animals. Even with this feature alone, it could be argued that fetuses aren't persons (before a certain point, we will see this later). One way to think about this is that you take the set of all things out there, and then slowly add criteria to whittle the total down to just persons. It is important to realize that even in international law, `human' is not the same as `person'.  Framing the question in this way moves it out of ethics and into metaphysics. So, ``is abortion permissible?" is an ethical question, while ``is a fetus a person?" is a metaphysical one. The answer to the second gives us the answer to the first.

\subsection{Feature 2: Reasoning}

The second feature which seems to be necessary for person-hood according to Warren is reasoning. This is the developed capacity to solve new and relatively complex problems. This, too, is not found in all humans. Most people might think that this is a necessary part of consciousness, so it should not have its own section. But this is not quite true. Consciousness, as we are using it, is having a `what-it's-like'-ness. Having sensations. A being can feel pleasure and pain without having the ability to reason. It also should be noted that we are worried about the mental capacity to solve the problems, not the physical ability. Infant humans, if they are not disabled to certain degrees, have the mental ability to solve the problems, but not the physical ability (strength, dexterity, so on).

\subsubsection{Does this thing have reasoning?}

There are some basic tests to tell whether a being has reasoning. These tests are likely going to be more involved than merely trying to tell whether the thing can feel pain/pleasure. The task is to give the creature a puzzle and see whether it can solve the puzzle. For example, in the case of a raven, put some food just out of its reach and see whether it can come up with a way to get the food.  

\subsubsection{Are we the only reasoning creatures?}

Just like with consciousness, we also find this capacity in many non-human animals (chimpanzees are an easy example, same with dolphins). We also find this capacity in some fish (octopuses are a great example). In general, if the being can figure something out, or at least shows that it's thinking about a problem in a more abstract way, then we can say that it has reasoning. But, it's also true that some humans lack this feature. There are some which are severely mentally handicapped, those in the later stages of dementia, and so on. These humans are certainly human but they are not, according to Warren, persons.

\subsubsection{Why does this matter?}

As before, if reasoning is an essential part to being a person, we can further whittle down our list of potential persons. As before, humans developed to a certain point have reasoning, but before that we don't. So, fetuses don't count here if they are prior to a certain stage of development. Also, many non-human animals do stay in the list. For example, we have more complexly intelligent animals and some fish (octopuses, cetaceans, new world monkeys, apes, chimpanzees, bears, otters and so on, basically, if you can train them, they have this), but plants are now certainly out. 

\subsection{Feature 3: Self-Motivated Activity}

The third feature is a bit more restrictive. Self-motivated activity is closely tied with reasoning and consciousness. This is activity which is relatively independent of either genetic or direct external control. Some may think that this requires some kind of libertarian free will, which seems to require some kind of soul. Warren, however, gets around this worry by adding in that it's independent of `direct' external control. This allows for indirect external control to have a part in it. To see the difference, direct external control would be a case where a mad scientist puts a microchip in your brain and controls you with a remote. This would not make you free, your actions would not come from you at all. On the other hand, indirect external control would be the sort of thing which you, more than likely, are experiencing right now. You have been heavily influenced by the past and your experiences, your choices are dictated by those, they are not necessarily instinctual. Both the hard determinists and the compatibilists would be fine with saying that actions can be independent of direct external control, but both will claim that they are indirectly controlled, by the past and the laws of nature. The hard determinist, however, would not be cool with the idea of morality, however, they would say that (although it's there) it doesn't count.

\subsubsection{Is this thing `free'?}

In general, the tests which we apply to figure out whether a being has reasoning will apply here. We could call those tests a `two for one'. When it comes to the tests for reasoning, we are asking whether they can solve puzzles, and to even engage with a puzzle to solve it, without really strange external factors being included (such as, being a remote-controlled robot), requires you to have self-motivated activity.

There are potential ways for isolating this feature and testing only it. For example, we would need to make a scenario where the creature (human, animal, robot) is denied external motivation for acting, it would have no instictual reason to act. If the creature still engages in the activity, then it would be self motivated. For example, a spy-camera in your house watching your pet. If we see that the pet acts without direct external interaction, then we could say that the action is self-motivated. 


\subsubsection{Are we the only `free' things?}

Some, like Descartes and Kant, will claim that humans are the only creatures which can be free, and even some humans (non-person humans) lack this. This mostly stems from their Libertarian Free Will intuitions. But, if we limit the scope of self-motivation to something within the range of a compatibilist, then we have that there's no reason to think we are the only ones with it. In the case of Descartes, as we will see later, he claimed that animals lacked a soul, so (as a consequence, though not the one he was shooting for), non-human animals can't have self-motivated activity. 

As with the previous, we see again that we aren't the only beings which count as having self-motivated activity. Much of the same beings from the reasoning section remain, but this is mostly because I don't know of a way to tell that a being has reasoning without getting that it has self-motivated activity.

We also have that some humans lack self-motivated activity. This list is much the same as before. Fetuses, yet again, lack this feature prior to a certain point in development.

\subsubsection{Why does this matter?}

Intuitively, it seems clear that for a being to have moral rights and be in the moral community (be a person), they would need to be able to act. Reasoning and consciousness can only get you so far. We also need assurance that these beings are acting freely. Without this, it just does not seem to have the kind of weight needed. If we say that a being is a person, then they must have moral responsibility, to at least some degree, which gets us, by definition, self-motivated activity.  

Similarly to the previous two, there are some humans which lack this ability. For example, the severely disabled or humans prior to a certain point in development. But there are certain other, non-human, animals which have this feature. In fact, it could be argued that most animals have this and fish. As before, also, some great examples are chimpanzees, dolphins, various new world monkeys, and octopuses. As with the previous two, fetuses, prior to a certain point, lack these. 

\subsection{Feature 4: Communication}

This fourth feature is where person-hood becomes far more restrictive for Warren. Communication is the ability to express messages, by whatever means (not just vocal, but signing counts, and so do other methods), messages of various types and with a large variation of contents. 

Warren goes, I think, a little too far in her definition of communication; claiming ``by whatever means, messages of an indefinite variety of types, that is, not just with an indefinite number of possible contents, but on indefinitely many possible topics." This seems very extreme to me, and I doubt that I even qualify here. But, limiting it to as I have said above helps. This criterion, as she phrases it, would mean that I would need to be able to talk about anything and  be able to do so in indefinite number of ways, which I think is just not possible, my brain is just not that big.
\subsubsection{Can this thing communicate?}

The real discrepancy here, and the reason Warren makes such and extreme standard for her communication, is that we don't want it to be just simple messages. The messages need to be more complicated than, for example, the chemical trails which ants leave or the dancing movements of bees. Rather these messages need to convey complicated information and they need to be able to convey it in several different ways.  The real point is to raise the bar on how smart the being needs to be to make the cut for personhood. The tests here are going to be a little more relative to the kind of creature which we are testing. For example, with some particularly primitive human languages, the messages might not be able to be expressed in several different ways, but the speakers can learn the different ways (though the older members will have more difficulty). The basic test would be to watch the beings interact with each other and notice the kinds of messages they are able to understand and convey to each other. Is there a grammar? Is the communication structure learned or instinctual? Can they understand abstract concepts?   
\subsubsection{Are we the only talkers?}

Despite what my friends who study linguistics might think (claiming that humans are the only language-users), when it comes to communication, we are far from the only ones. Animal communication is very wide-spread, with, I would argue, the complexity necessary for personhood. For example, dolphins have communication, and this is not instinctual but learned, we have even figured out some words in at least one of the variety of languages spoken.\autocite{tedtalksdirector2013} Chimpanzees have this capacity, but it does not seem to have one naturally arising. Some new world monkeys certainly have this capacity and have a naturally arising and learned languages (my personal favorite example is the cotton-top tamarin)\autocite{tededucation2014}. It may be also the case that octopuses are in this category.  If we treat this as a standard for intelligence, then individuals with the metal capacity to communicate, but not the physical ability, would qualify. That being said, some humans lack the mental capacity to communicate, not just the physical ability. This can be due to a variety of reasons.  And, as it applies to the relevant topic, fetuses before a certain point, lack this capacity all together.
\subsubsection{Why does this matter?}

In general, the ability to communicate our thoughts and intentions is the biggest sign of intelligence but it's also a bench-mark for degrees of intelligence. My father, who is mono-lingual, has often claimed that speaking multiple languages is a sign that the person is really smart, encouraging me to learn several (which I have). Though I have met very smart mono-lingual people and not-so-smart multilingual people, the ability to express oneself is a fairly intuitive standard. Similarly, if a being which had this mental capacity and later lost it, we often, in the real world, hold them to a different moral standard. For example, a person with sever mental disabilities is not held as responsible for their actions as a person without those disabilities, even if they behave the same way (in this case). And, it seems that they should not be held to those standards. This criterion limits the scope of person-hood yet again. Sure fish can reason, but they certainly aren't smart enough to be persons. Similarly, cats and dogs are in the same boat. So, the examples I have been giving thus far remain as potential non-human persons, but some are excluded. Yet again, fetuses are not included in this list. 
\subsection{Feature 5: Self-Awareness}

This is the fifth and final feature of person-hood for Warren. This is self-awareness. This one, one could think, should be earlier than communication, as there are many creatures which have this but lack communication. To be self-aware one must have self-concepts, understand that they are a different being from others. You are an individual, not a `hive-mind'. You can think about you, what you want.
\subsubsection{Is this thing self-aware?}

This is where I give the tests which one can use to tell whether a being is self-aware. In the sciences, the commonly used test is to show whether the being can recognize themselves in a mirror, know that the reflection is not another being, then they have this self-awareness.  But this method is flawed in several ways. There are some creatures who clearly have self-concepts but lack interest in mirrors. In the previous feature, I gave cotton-top tamarins as being talkers, and there's some solid evidence that the nature of their communication does require self-concepts, but they have failed the mirror test (when tested at least once, but there may have been something wrong with the testing perimeters). For example, when I was paying for my community college, o so long ago, I was working in Dementia Care. In trying to put an elderly woman to bed, she saw herself in a mirror. She called out to the reflection, asking  it to leave her room and consistently looking back and trying to get it to leave. Eventually, I had to cover up the mirror. But, it's clear that, though her dementia was extreme enough for her not to have the same moral privileges as you or I, she still had self-concepts.  So, though the mirror test works in most cases, we need to be careful about calling it a definitive test. A more holistic testing model is appropriate, so that we can weed out false positives and negatives.  
\subsubsection{Are we the only ones self-aware?}

It is certainly not the case that we are the only ones self-aware. Other creatures certainly are too, even the ones which can't communicate. For example, various cetaceans (whales and dolphins), primates, and some other creatures, such as some new world monkeys.  Some octopuses have failed to show that they are self-aware in a way which I am willing to accept, namely in the presence of a mirror, they behaved as if there was another octopus present. But the jury is out on this for more intelligent species of octopus. In general, if a being is able to communicate to the degree necessary for person-hood, then it's going to have this feature, but not the other way around. It's worth noting that not all humans are self-aware. We have the extreme cases of humans in irreversible comas or humans born without certain portions of the brain. Fetuses, yet again, lack this feature. 
\subsubsection{Why does it matter?}

As with the other standards, this raises the bar on what it takes to be a full-fledged person. Lacking the presence of self-concepts does not entail the sort of moral duties to them which we would place on ourselves for beings with these concepts. Persons are individuals, we have duties to them as individuals. If something can't identify as an individual, then we don't have the same duties to it. 

\subsection{An Interesting Tangent}

As we have seen through our analysis of these features which make a being a person, there are some humans which are not persons. Warren herself does not go down this rabbit hole, but using her requirements for person-hood gives us a very interesting tangent:
\begin{center}Is it possible for a non-human to be a person?\end{center}

This question is very interesting for the discussions of Animal Rights. To have rights, to have aspects which others have a moral duty to ensure, you certainly need to be a person. If you are a person, then you have rights. Though it is possible for something to have rights and not be a person in the fullest sense, but those rights would be limited. Applying Warren's 5 criteria to other creatures in the world, we see that there are non-fictional, real, actual, non-human persons on the planet right now. Now, I am not an Area 51 conspiracy theories, saying that their are aliens in the base. But, looking at these, we have some fascinating arguments to show that humans aren't the only persons. Proving and establishing legally that these beings are non-human persons will give animal rights activists a strong argument and a more powerful footing in making their case. As persons, these creatures will need to be given the same moral consideration as you or me. 
Below, I will give the 5 (five) categories of creatures where at least some of them have person-hood. For some, the category will have more than one example in them. The first are the least contentious and the last are the most contentious. 
\subsubsection{Primates and (some) New-World Monkeys}

Of course, humans are primates, but we are human, so not exactly non-human persons. But we do find the capacities for the aspects of person-hood in the non-human members. To start, they clearly react to external stimulation, so there's something going on upstairs, that's the first box ticked. Second, they can solve reasonably complex puzzles and even can make tools, so that's a second aspect met. Third, through their ability to solve puzzles and makes choices, we can see that their actions are not merely instinctual, they do have a moral compass, so to speak. Fourth, though they don't have a naturally arising language, they can be taught it and will use it even when not prompted to speak with each other if they know that the other will understand. Also, once the primate understands a language, it will teach its children the skill, with them even, sometimes, inventing new words. For more information on this, check out Washoe the Chimp. And fifth, primates, by and large, do have the ability to recognize themselves in the mirror and they do have self-directed thoughts and awareness. All of these features make it so that at least some primates, not just humans, are persons and, from that reasoning, deserve the same moral consideration as we would a human with the same mental capacity. 

When it comes to new-world monkeys, we have a very interesting case, these are the cotton top tamarins. These little guys are quite amazing. Not only do they have all of the features of person-hood, like the primates, but they have an added aspect. They have their own, learned, naturally arising language. Their language was not taught to them by us. They react to external stuff, and so forth. The only area which requires proving is whether they have self-concepts, but their language aspects seem to show that they would. 
\subsubsection{Cetaceans}

These are your dolphins and some whales. At present, some countries have recognized these creatures as non-human persons and have granted them various rights, even though they did not follow the same kind of reasoning given here. Dolphins, in particular, do have all of these features. First off, they react to external stimuli, which is going to be true for all animals worth mentioning. Second, they can solve problems and learn from each other. For example, one pod of dolphins independently learned to use a sponge to root the sea floor and others have learned this behavior from them. Third, their behaviors are not always instinctual and are independent of direct external control. For example, they play and will even engage in behaviors contrary to what we would think they would instinctively. Fourth, they do have communication, and a rather sophisticated one. The language is learned by the children, meaning that different pods might not understand another and we have even learned aspects of it able to communicate with them. Similarly, they are able to learn more than one language. Dolphins which we have trained will learn the whistle patterns for various tricks (like a dog), but will also mimic them to attempt at communication with other dolphins and even understand that the whistles in different orders will mean different things. The fifth aspect is self-awareness. This can be seen in both how they will pass the mirror test and also how they have naming customs for their young and how they introduce themselves. These make some dolphins non-human persons and worthy of our protection.   
\subsubsection{Extraterrestrials and Advanced AI}

Both of these are at the bottom merely because their existence is controversial. No, I do not think that such beings are currently on the planet, rather they are worthy of mention because of their possibility. The section regarding the Mind-Body Problem shows us that proving person-hood for an AI will be tough, it will need to have Strong AI, if they are possible at all. However, a Strong AI would obviously be able to speak, have consciousness, self-awareness, self-motivated activity, and reasoning to the same degree as human persons, if they are possible. Not recognizing that an AI machine has reached the level of person-hood is the base-line for pretty much every robot-uprising Sci-Fi. 

Extraterrestrials will likely have an easier time proving their person-hood than the machines. This is because they will have likely came from a process much like the one humans did. But, it would not surprise me at all if, on the day the first contact is made, there's a group out there who think that they aren't persons, not worthy of our consideration, because they have built into it the idea that person-hood is exclusively human. This, it would seem, would be mistaken. 
\input{iwasonceafetus}
\chapter{Part 22: The Abortion Debate (Pro-Life)}
For this module, thus far, we have been looking at one of the best arguments which can be made on the pro-choice side. This argument goes after the core beliefs of the pro-life side. Namely, that if a fetus is a person, then abortion is murder (and thereby wrong) and that a fetus is in fact a person. The argument does not rely on consequentialist reasoning nor on a woman's bodily rights.

For the sake of fairness, I feel that it's appropriate for us to also look at one of the best arguments from the pro-life side.  This is a purely optional reading as you can complete the assignments without reading it. However, it may serve you well if you ever encounter puzzles like this or if you are on the pro-life side of the debate, this paper will give you a far stronger argument for your stance. The paper is called I Was Once a Fetus: That is Why Abortion is Wrong by Alexander Pruss. His argument does not rely on religious beliefs nor on any notions of person-hood (per se). The argument goes like this:
\begin{earg}
    \item[1 ]I was once a fetus.
    \item[2 ]If I was once a fetus, it would have been wrong to kill that fetus.
    \item[3 ]If it was wrong to kill me when I was a fetus (that fetus), then it would be wrong to kill anyone while they were a fetus.
    \item[4 ]Therefore, it is wrong to kill anyone while they are a fetus.
\end{earg}
The conclusion, the fourth line, makes all cases of abortion morally wrong. It may seem a bit strong, but that is what is argued. We will go through the paper point by point.
\section{I was once a fetus}
This is the first line of his argument and it is the most seemingly obvious of the lines. Pruss starts off by asking us a few rhetorical questions (avoid them in philosophy papers, they just lead to confusion if they are not well crafted and obvious). The point of them is to show that at some point in the far past `I' was an embryo, then a fetus, then a neonate, then an infant, then a toddler, a child, an adolescent, and then, finally, an adult. There are several ways in which a person can show that they were once a fetus. Pruss does this in an interesting way, one which I would not have thought to apply to this debate. he uses what we in the philosophy biz call a `continuum', or an indiscrete series. To make these philosophically interesting, you need to have it such that something clearly holds on one side of the spectrum and clearly doesn't hold on the other. These are often called Sorites Paradoxes. `Sorites' is Greek for `heap' or `pile'. These sort of cases are where we get the stereotypical philosophical question ``how many grains of sand does it take to make a heap?" So, imagine that you have a pile of sand, a large pile, something which is clearly a heap of sand. If you take 1 singular grain of sand off of the top, is it still a heap? Clearly and obviously it is. Now, I remove another, and another, and another. All the while, 1 grain of sand doesn't make a difference, all the way down until I have only 1 grain of sand left. This grain of sand is clearly not a heap, but the reasoning shows that it is. For another example, take a spectrum of colors, from blue to red. The far side is clearly blue and the other is clearly red. What if I start at the blue  end and move over, ever so slightly, is it still blue? Yes, yes it is. Now, what if I move over a little more, still blue, a little more, still blue, and so on. In the middle, we will likely find something which we would call `purple', but, by the reasoning, it would still be blue. As we continue, the responses get seemingly more and more absurd, until we are saying that red is blue.   

Any line which we draw in this spectrum, saying that before this point, it's blue and after this point it's red or some other color would, frankly, be unnatural and arbitrary. In a third example, one which was given to me when I learned about this paper in the equivalent of 101 which I took (that course was exclusively epistemology and metaphysics, no ethics), concerns baldness. My professor, at the time, was balding, and was the example person for the entire department for this concept. On one side, you have him when he was young, full head of glorious hair. On the other side you have what he will become, a chrome-dome. Now, where is he now? Somewhere in the middle, is he bald? No, but is he `haired'? No. There's the paradox, something is both bald and not bald. 

Now, at this point, the same sort of reasoning can be used for the developing embryos through pre-birth infants all the way to you reading this in-front of some screen. You, now, are clearly you. You 1 second ago is also clearly you. You 2 seconds ago is also clearly you.\footnote{ I am not sure about the grammar of these two sentences.} We can keep going back, and back, and back, deducting seconds off the clock, until we get to your birth. Before this point, it gets a little more tricky to imagine it. The continuous nature of the transitions are even more evident in this case. We have, now, that at your birth, that's you, and on the opposite side, at conception, we have a zygote, or some such. But this is again, a continuous spectrum. There's no hard-line, no non-arbitrary position where we could, within reason, say that you appeared and the zygote disappeared.  Without such a line, then you are that zygote. So, this first line is pretty obviously true. I was once a fetus.

\section{If I was a fetus, it would have been wrong to kill that fetus}

The last line of the argument showed that I was once a fetus. It is worth noting, and this will come up again later, that I could have used any one of you in place of myself. Now, our job is to show that if I was once a fetus (I was), then it would have been wrong to kill that fetus (me). We need to show that there's a causal connection between me (or you, or anyone) being a fetus and the wrongness of killing that fetus. There are two ways which I would go in trying to show this connection, if I were Pruss. Here's the route which he takes  (along with my connections to previous content in this course) and then there's my personal preferred route. We will start by looking at the route which Pruss takes and then move on to the route which I would take. Both involve an intermediary proof for the stance that it's wrong to kill me now. 
\subsection{Pruss' way of showing that it's wrong to kill that fetus}

If you remember way back in the course, we covered Nagel's work Death. In that work, Nagel came from the idea that death is always bad for the person who died because something of great value is lost. That thing must be of such great value that loosing it is worse than anything which can come from having it. For Nagel, this is experience. Having experience is so valuable that death is always worse than whatever experiences may come from the experiences. Connecting this into Pruss, my untimely death now would be cutting off experiences which I would have otherwise had. This means that killing me has deprived me of some good, and that makes it wrong. This is, interestingly, in line with the typical Consequentialist kind of thinking (as in the way of approaching it is how the Consequentialist would do so). This is not Utilitarian thinking, however, because experience is included as an ultimate, super, good, which Utilitarianism does not incorporate.

So, we have that it's wrong to kill me now, which is awesome, but how do we move from this to killing me as a fetus? Well, we need to look at why it's wrong. As I have likely mentioned several times now, the why is the big aspect, this is the aspect which we can use in debate, either for or against, and make headway. To reiterate, it's wrong because I was deprived of experience. Suppose that I'm 28 and I would have otherwise lived to be 65. Killing me now would deprive me of 37 years of experience. Had I been terminated as a fetus, I would have been deprived 65 years of experience. From this, because 65 is greater than 37, we can say that killing me as a fetus is at least on par with killing me now, if not worse than killing me now. Because of this, we get that if I was a fetus, then it would be wrong to kill that fetus. Putting this as a more formal argument, we have the following:
\begin{enumerate}
    \item Depriving someone of a good which they would have otherwise had is wrong. 
    \item Killing someone deprives them of a good which they would have otherwise had (namely experience). 
    \item If I was a fetus, then killing that fetus would have deprived me of a good which I would have otherwise had.
    \item Therefore, if I was a fetus, then killing that fetus would have been wrong. 
\end{enumerate}
\subsection{An Argument Which He Could Have Used}

Pruss' way of getting to the premise relies on the intuition which Nagel uses. This is not shared by many, but there are those who have it. An alternative way which Pruss could have used to get this line, but does not (for reasons which we will see later), is to reapply the Sorites style thinking to this problem. All one needs to make this reasoning work is that it's wrong to kill me now and that the morality of killing another person does not change with time (which is Non-consequentialist style thinking).  The first half, that it's wrong to kill me now, can be gotten through your preferred method, though non-consequentialist thinking will likely be preferred. To start, if it's wrong to kill me now, then it would have been wrong to kill me a second ago. This is because the morality of killing someone does not change with time. This also means that if it's wrong to kill me now, then it would have been wrong to kill me two seconds ago, and three seconds, four, and so on. This goes all the way back to when I was a fetus. This means that if I was once a fetus, then it would have been wrong to kill that fetus. Put as a more formal argument, we get the following:
\begin{enumerate}
    \item It is wrong to kill me now (at T).
    \item If it is wrong to kill me now (T), then it was wrong to kill me a second ago (T-1).
    \item If something is wrong at T-n, then it is wrong at T-(n+1).
    \item Therefore, If I was fetus, then it would have been wrong to kill that fetus.
\end{enumerate}
There is a very innocent jump between lines 3 and 4 which can be glossed over. Basically, to make that move, there are some additional, veiled, premises which concern the time I was a fetus and the identity between myself and that fetus. The main point of this argument is that there's no hard line which makes it OK to kill me before a certain point, and wrong to kill me after that point. 

\section{If it was wrong to kill me when I was a fetus, then it was wrong to kill anyone when they were a fetus}

The author's original line used the gendered pronoun which I have replaced, but the point remains the same. Proving this line in a way that makes the entire thing still valid is a bit more difficult than normal. You see, this will appear to be inductive, which is not strong enough for philosophy, but it's not. As I have mentioned a few times in this proof, I am not special. I could have put that you were once a fetus and ran the argument in the same way and I could have put any person in this argument in place of myself and gotten the same result. Since there are no cases where the validity of this argument does not transfer to another, we are safe in making the generalization to all people. Take these two arguments, which are oddly similar:


\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
    I Argument&Sally Argument\\\hline
    I was once a fetus.&Sally was once a fetus.\\\hline
    If I was once a fetus, then it would have been wrong to kill me when I was a fetus.&If Sally was once a fetus, then it would have been wrong to kill Sally when Sally was a fetus.\\\hline 
    Therefore, it would have been wrong to kill me when I was a fetus.&    Therefore, it would have been wrong to kill Sally when Sally was a fetus.
 \end{tabular}

The first argument is the one which we currently have a proof for. But, if we take the reasoning and replace the references to myself with references to Sally, we get the same result. Now, what if I was to replace the references to Sally with some generic, like, I don't know, a human person. Take these for the comparison:


\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
    Sally Argument&Human Person Argument\\\hline
    Sally was once a fetus.&A human person was once a fetus.\\\hline
    If Sally was once a fetus, then it would have been wrong to kill Sally when Sally was a fetus.& If a human person was once a fetus, then it would have been wrong to kill them when they were a fetus.\\\hline 
    Therefore, it would have been wrong to kill Sally when Sally was a fetus.& Therefore, it would have been wrong to kill a human person when they were a fetus. 
\end{tabular}

But, a human person is just a generic stand in, if this reasoning works (and it does, soundness is a different story), then the reasoning applies to all, real or potential, human persons. This means that I can move from the individual case, that it would have been wrong to kill me when I was a fetus, to the stance that it would have been wrong to kill anyone when they were a fetus, because the justification applies just as well to them as it does me. This gives us the final premise of the argument, namely, if it was wrong to kill me when I was a fetus, then it would have been wrong to kill anyone when they were a fetus. This last line connects everything together, and with some pretty easy reasoning, we have that killing anyone when they were a fetus is wrong, which makes abortion always wrong. 
\begin{earg}
    \item[1] I was once a fetus.
    \item[2] If I was once a fetus, then it would have been wrong to kill that fetus.
    \item[3] If it would have been wrong to kill that fetus, then it would have been wrong to kill anyone when they were a fetus. 
    \item[4] So, it would have been wrong to kill anyone when they were a fetus.
    \item[5] If it would have been wrong to kill anyone when they were a fetus, then abortion is wrong (this is implied from the definition of an abortion). 
    \item[6] Therefore, abortion is wrong.
\end{earg}
With that, we have what is likely the best pro-life argument which you will find. Your typical pro-life argument which we saw earlier relied on an equivocation between `human' and `person'. This one, however, relies on sorites reasoning and, maybe, some intuition about experience. This could, maybe, also be used to make a hole in Warren's reasoning as well. But, that being said, there are some objections to this argument which need to be addressed. It should be noted that Pruss' replies rely on the Nagel intuition which we have seen previously. To reply to these objections without the value of experience being so much would require a different line of thinking than the one which Pruss uses. 
\section{But I was wanted!}

This is rather cold, I admit, but there may be one important difference between fetus-me and some other fetuses, namely that I was wanted. But, you have to notice that no-where in the arguments for the impermissibility of abortion were the claims that I was wanted. It solely relies on the fact that I am a person and that I was a fetus. There are some cases where the pregnancy and child-rearing will result in undue and heavy burdens on the mother. In these cases, it seems like abortion should be available (according to some). This is not to say that those burdens will always be present in that person’s life, but rather that they would not be removed or be very difficult to remove if she went to term. This sort of reasoning falls in line with one of the popular pro-choice arguments which I have given for Warren, namely the ones which rely on the undue burdens on poor women. 
\subsection{Pruss' Reply}

It is worth noting that nowhere in the argument was there something about the want of a child, nor was there anything about the life of the child after birth. Even if I wasn’t wanted, there’s still a loss, namely the fact that I would have otherwise had experiences. A person is a person, no matter how small. Morality, often, will force us into situations which we don't want to do, which might not be in line with our self-interest. 
\section{But I didn't endanger my mommy!}

That was me making a joke of the objection a bit. The point of it is that, as far as I am aware, there were no complications with my birth/time in the womb which added risk to my mother's life, which may make a relevant difference between me and other fetuses. There are some cases where a person must choose between the life of the mother and the life of the child. There are other cases, even more extreme, where the process (not just child-birth) of going to term will kill both. Such as cases of unidentified ectopic pregnancies. These cases make up 10\% the deaths during pregnancy.
\subsection{Pruss' Reply}

But, you have to notice, that we have to chose between the life of one and another person from time to time, though it is not common. More over, in cases where a fetus is endangering the life of the mother, it does so unintentionally. In cases where it is acceptable to kill a person who is endangering another, it is intentional. So, since it is unintentional, it is not acceptable to kill the fetus. Also, if it comes to a choice between the mother and the child, if we were to apply Nagel’s thoughts, we should choose the child. Namely because the experiences had by the mother up until that point plus those which will be had by the child is greater than the amount had by the mother if she were to continue living. This, however, does not work in cases where the fetus is not able to develop and the process of development will only kill both. For example, extreme cases of ectopic pregnancies. 
\section{But I was a healthy fetus!}

This becomes a debate about euthanasia (the killing of a person painlessly when they have some extreme health problem). Killing me as a fetus would have deprived me of a long life, but killing an unhealthy fetus would not have deprived them of much. For example, Peter Singer, who I have used as examples before, is a rather extreme Utilitarian, both in his work as well as in his actions. His charity work and donation practices are top-tier. However, when Singer was directing a bioethics center in Australia (which is a universal healthcare country), he was contacted by doctors concerning their ethical dilemmas. In one case, he was contacted concerning issues in the Neonatal Intensive Care Units, ICUs for newborns.  Many of these newborns had a range of very serious medical conditions which often lead to a very slow and painful death. If the newborns actually did survive, they would need multiple extremely expensive surgeries which would drain the resources from the collective system (it's even worse in a system without universal healthcare) and they would be very disabled in various ways. In the extreme cases of these conditions, the ones which the doctors were contacting Singer about, all people concerned, the  doctors, nurses, and even the parents, believed that the babies should not survive. In these cases, all the newborns could do is experience pain. As a result, many infants were being left in the ICU untreated, with minor adjustments to alleviate the suffering where possible.  This was an very painful experience for all those concerned, on an emotional level and took up resources which would have otherwise been provided to healthy newborns. Almost all cases like these die before they are 6-months old.  Peter Singer advised that the newborns with the extreme cases of this condition be given a quick and painless death. Would it not be better for the parents to have this condition identified early and be able to have an abortion?  
\subsection{Pruss' Reply}

This becomes a debate about euthanasia (the killing of a person painlessly when they have some extreme health problem). The connection is that, for the author, killing an unhealthy fetus is the same as killing a person with a terminal illness. So, if you think that killing a person with a terminal illness is permissible, then you must also think that killing an unhealthy fetus is permissible. The author thinks that euthanasia is wrong, full stop. Again, this is a lot like Nagel's reasoning. The core principle is that no experience can make life not worth it. Because of this, every human life, for Pruss, must be worthwhile. 
\subsubsection{A Reply to this Reply}

There is an issue with this reply. Namely, if experience is what makes human life always worthwhile, what if the fetus is going to be brain-dead? For example, what if the child has a certain genetic abnormality or some other factor which results in anencephaly (Greek for no-in-head (AKA brain)). This is a real thing, the baby, if it has the main brain stem, will be alive, but will not have any experiences whatsoever. No consciousness at all. The life expectancy for a baby born with this is between a few hours and a few days. This means that the human's life would not have anything to it, there would be none of what makes life always worth it, according to Pruss. This means that he does not have a way of saying that abortion would be wrong in this case. 
