\part{What is Philosophy?}
\label{ch.modone}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}
\chapter{Part 1: What Philosophy Is and Isn't}
Philosophy is not just a way of life, it is not just some outlook which a person takes on. It is also not the product of some really deep thinking. Philosophic thought is not philosophy. One does not have a philosophy. That's not how it works. 'My philosophy on that is...' is like saying 'my science on that is...' What they should say is 'my belief is that...' or 'I have thought long and hard about it and I think...' Philosophy is critical thinking taken to the ridiculous extreme. Since it is, at its core, thinking, one does not simply study philosophy. Philosophy is an activity, it's something that you do. Doing philosophy involves finding a stance on some issue and then coming up with reasons for that stance. This could be a stance that you hold or it could be the stance that your opposition holds in some debate.\footnote{This would be putting yourself in 'their shoes', so to speak, so that you can see where they are coming from and thereby exploit the flaws in their stance.} In order to really truly prove your point, you will want to know why another might believe otherwise and have the skills to show the errors in that reasoning. Engaging in philosophy is how you get the skills to find the errors in the reasoning. 

Since philosophic activities tend to go really abstract really quickly and since the questions which philosophers tend to think about are very difficult to answer, some have claimed that Philosophy is a pointless endeavor. This is, however, not accurate. All fields of study are defined, roughly, in terms of the questions which they are trying to answer or have the ability to answer. The questions in Philosophy tend to be those which other fields can't answer. In those cases, most of the time, the standards are very high, higher than other fields can match. Sometimes, those standards are so high that some claim that they are not possible to answer. Again, this is not accurate. There are answers, even if those answers are difficult to get. Doing philosophy tends to involve the following activities:
\begin{tabular}{p{40mm}|p{50mm}}\hline
Philosophy Involves&Example\\\hline
Resolving Confusion &Noticing that people misunderstand something because of a word-choice.\\\hline
Unmasking Assumptions &Seeing that you/another hold a position for a reason which was not stated.\\\hline
Revealing Presuppositions &Seeing that there are unexamined/unexplained jumps in reasoning or an aspect of the reasoning which is hidden.\\\hline
Distinguishing Importance &Figuring out which things you need to do vs which things you want to do.\\\hline
Testing Positions &Putting yourself in another's shoes to see whether their stance makes sense or to see whether your own stance holds water.\\\hline
Correcting Distortions &Fixing/identifying misrepresented stances or facts, seeing that you are being misinformed/lied to.\\\hline
Looking for Reasons &Asking yourself why a person would think/do what they are thinking/doing\\\hline
Examining World-Views &Seeing whether another's explanation for something or some course of behavior is correct/accurate to the world itself. People can have wrong opinions.\\ \hline
Questioning Conceptual Frameworks &Seeing whether a particular way of carving the world, a way of thinking about it, is better or worse than another way. We may think that the world is one way, but is it?
\end{tabular}
\section{Part 1.1: How Does One Do Philosophy?}
\label{s:p1.1}
As a kind of inquiry, philosophy is aimed at establishing knowledge and understanding. Even when you can't get exact and precise knowledge about a topic, you can often find interesting things to learn (for example, why you can't get that knowledge). So, rational inquiry (critical thinking, doing philosophy) may be interesting and fruitful even when we don't get straight-forward answers. Once we raise a philosophical issue, a question or puzzle, whether about the nature of justice or about the nature of reality, we want to ask what can be said for or against the various possible answers to our question. For example, if my question is 'when is it morally permissible to lie?', I look at cases where it seems OK and cases where it just seems wrong. I look at the \textbf{reasons} why they seem that way.

At this point, I am making \glspl{argument}. These are my sets of reasons for a stance. Some arguments give us better reasons or accepting their \glspl{conclusion} (what is being argued for) than others. Once we have an argument, we want to evaluate the reasoning it offers. So, if you want to know what philosophers do, this is a pretty good answer: philosophers formulate and evaluate arguments. We look at the reasons for some stance and figure out whether those reasons actually hold water, so to speak.  Your introduction to philosophy should be as much a training in how to do philosophy as it is a chance to become acquainted with the views of various philosophers. To that end, you should carefully study the sections below on arguments. I personally approach teaching philosophy from the perspective that there's still a lot going on, I show you the relevance of philosophy to questions which we are puzzled with today (like Artificial Intelligence, the Abortion Debate, the Existence of God, and so on). Through this, you will become accuanted with philosophers, both historical and contemporary, but this is more of a byproduct than a goal.

Once we have some philosophical position to think about, we want to ask what arguments can be made for or against it (\textbf{formulate}). We then want to examine the quality of the arguments (\textbf{evaluate}). Evaluating flawed arguments, figuring out why they are wrong, often leads us to other, stronger, arguments and the process of formulating, clarifying, and evaluating arguments continues.

This circular method of question and answer (make an argument, evaluate it, make a stronger one, and repeat) is known as a \gls{dialectic}. A dialectic looks a lot like debate, but a big difference is in the goals of the two activities. The goal of a debate is to win by persuading an audience that your position is right and your opponent’s is wrong.

\newglossaryentry{dialectic}
{
name=dialectic,
description={the art of investigating and determining the truth of some stance or opinion},
plural=dialectics
}


A dialectic, on the other hand, is aimed at inquiry. The goal is to learn something new about the issue under discussion. Unlike debate, in a dialectic your sharpest critic is your best friend. Critical evaluation of your argument brings new evidence and reasoning to light. The person you disagree with on a philosophical issue is often the person you stand to learn the most from (and this doesn’t necessarily depend on which of you is closer to the truth of the matter).

Dialectic is sometimes referred to as the Socratic Method after the famous originator of this systematic style of inquiry. For this module, the reading is one of the more famous of Plato’s dialogues, namely The Apology. This will give you a good sense for how the Socratic Method works. Then watch for how the Socratic Method is deployed throughout the rest of the course.

Doing philosophy boils down to thinking really hard, but in a structured and organized way. The thinking must be critical and comprehensive.

The point of arguments is to get at the \textbf{truth}. But, in the kind of world we live in today, it is worth taking a look at what philosophers mean by 'truth', before we get too much into arguments.

\subsection{Part 1.1.1: Truth}
\label{s:p1.1.1}
Since both science and philosophy are mainly concerned with getting at knowledge and understanding about the world (though the kind of knowledge may be argued to be different), it is natural to think that both are after the truth about things. There are some interesting and some confused challenges to the idea that philosophy and science are truth oriented. But for now let’s assume that rational inquiry is truth oriented and address a couple of questions about truth. Let’s focus on just these two:
\begin{earg}
    \item[\ex{truth1}] What is it for a claim to be true?
    \item[\ex{truth2}] How do we determine that a claim is true?
\end{earg}
It’s important to keep these two questions separate. Questions about how we know whether something is true (like the second question) are epistemic questions. Epistemic questions, which we will return to in a later module, are questions about knowledge, beliefs, reasons, or, to a certain extent, faith. But the question of what it is for something to be true (the first question) is not an epistemic issue. The truth of a claim is quite independent of how or whether we know it to be true. Questions about whether or not something is true are, more or less metaphysical questions.

For example, consider these two claims:

\begin{earg}
    \item[] There's extraterrestrial life.
    \item[] There's no extraterrestrial life. 
\end{earg}


I assume we don’t know which of one of these is true, but surely one of them is, for any sensible claim that you make, it's either true or false. Whichever of these claims is true, its being true doesn’t depend in any way on whether or how we know it to be true. So, there is a correct answer to the question "is there extraterrestrial life?" but as it sits, we don't know what that answer is. There are many truths that will never be known or believed by anyone, and appreciating this is enough to see that the truth of a claim is not relative to belief, knowledge, proof, or any other epistemic notion.\footnote{There are certain facts within mathematics which are not knowable, this could be because of the physical limitations of the computations or because of the nature of the question which the fact is an answer to. For example, there is a first digit to Graham's number, but it is so large that it's physically impossible to determine.} Similarly, unless you hold certain philosophic positions about certain things (we will see this in the Relativism/Meta-Ethics Module), just because you believe something, that does not make it true. How much weight you give a belief also doesn't change how true it is. Unlike somethings which you may encounter, like knowledge, certainty, belief, and so on, truth does not come in degrees (and this is regardless of the stances you take). 

But then what is it for a claim to be true? The ordinary everyday notion of truth would have it that a claim is true if the world is the way the claim says it is. And this is pretty much all we are after for this class. When we make a claim, we represent some part of the world as being a certain way. If how my claim represents the world fits with the way the world is, then my claim is true. Truth, then, is correspondence, or good fit, between what we assert and the way things are. There are other accounts of truth out there, which do better or worse jobs at various things (for example, this account can't make "Harry Potter is a wizard"\autocite{HarryPotter1}, "Stepan Arkadyevich Oblonsky's wife is mad at him"\autocite{AnnaKarenina}, or "Sherlock Holmes is a detective"\autocite{Sherlock} true, because those beings don't exist), but this works for our purposes.  

\subsection{Part 1.1.2: Truth and Meaning}
\label{s:p1.1.2}

A potential confusion about truth comes from confusing a sentence with the meaning behind it. This is where we get into a very basic introduction into Philosophy of Language. This area of philosophy deals with things such as meaning, reference, and so forth (Philosophy of Language is what I mostly work in, it deals with semantics while Linguistics deals with syntax, but the fields overlap quite a bit). Even when we aren't trying very hard, we can use words and sentences in a ton of different ways and there seems to be some vagueness or ambiguity in natural languages. For example, I can easily make examples where sentences might have two or more different interpretations, like double or triple entendres. 
\begin{earg}
\item[\ex{ambi1}] GOP grills the IRS chief over lost emails.
\item[\ex{ambi2}] If your dog poops, you must put it in the trash can.
\item[\ex{ambi3}] A woman gives birth in the UK every 48 seconds.
\item[\ex{ambi4}] People wanted for pickling and canning.
\end{earg}
All four of the above examples have at least 2 different interpretations and, in each case, one seems true while the other seems false. In Example \ref{ambi1}, it could be interpreted to mean that the GOP are having an IRS chief BBQ over a server or it could mean that they are harshly questioning the chief. In Example \ref{ambi2}, one way to understand this is that it's telling you to put your dog in the garbage can, while another is telling you to put the poop therein. In Example \ref{ambi3}, we could say that there's a single woman who is having a baby every 48 seconds (she must be tired) or that some woman, not necessarily the same woman, is having a child in that time. In Example \ref{ambi4}, it could be that we have a cannibal looking to store their meats for the winter using canning and fermenting techniques or it could be that a food fermentation business is looking to hire on some more people in those departments. Given these examples, and how often we misunderstand each other on a daily basis,  it might be tempting, therefore, to think that the truth of a sentence must be relative to its interpretation.  In an even more robust example, imagine the following:

\factoidbox{Suppose that we all collectively switched the sorts of things the words 'dog' and 'cat' pick-out. So, the word 'dog' now is used to point out the meowing critters and the word 'cat' refers to the barking ones.  In this case, it would seem that the sentence 'dogs are canines' would be false and 'dogs are feline' would be true. If we, again, flipped the meanings of the words 'feline' and 'canine', we would get that 'dogs are canines' as true, but for a totally different reason than it was originally.}

But does this make truth open for interpretation? Well, no, but in a sense, yes. When we look at a language, from one perspective, we notice that things like words and sentences are nothing more than characters on a screen/page, random collections of sounds, or certain structured gestures (in the case of sign languages). Those things, on their own, don't have meaning. There must be something extra, beyond the sound and signs, which has the meaning.  Philosophers call the meaning behind a sentence a proposition. A proposition, itself, is not a sentence or a word, rather it's the meaning behind the sentences.

\begin{tabular}{p{1in}|p{1in}|p{1in}}\hline
Sentence &Language &Proposition\\\hline
Snow is white &English &that snow is white\\\hline
Schnee ist weiss &German &that snow is white\\\hline
Nix alba est &Latin &that snow is white\\\hline
La neige est blanche &French &that the snow is white\\\hline
\end{tabular}

All of the above examples are different sentences, made clear because they are in different languages, but they all express the same proposition. The truth of a sentence is relative to the truth of the proposition attached to it, propositions are the things which are true or false, and then the truth of the proposition is determined by its correspondence with reality. Translators, at least the good ones, often take the sentence in one language, figure out the proposition connected to it, and then express the same proposition in the necessary language.\footnote{Sometimes there can be cases where a language has the ability to express a proposition which is not possible to express in another without adding something to that language or giving information beyond what was in the original statement. For example, take Cicero's Epistulae ad Familiares 9.22, the examples Cicero gives in the letter of profanity and why they make sense are  not translatable without explaining some aspect of Latin phonology and semantics.}

So the proposition expressed by a sentence is not itself a linguistic thing. Propositions themselves don't have meaning, but rather they are the meaning behind a sentence. For a bit of language to be open to interpretation is for us to be able to attach different propositions (meanings) to it. But the meanings themselves are not open to further interpretation. And it is the proposition, what is meant by the sentence, which makes the statements, sentences, true or false. So, when I speak of arguments consisting of premises (the supporting evidence and their arrangement in the argument), I am talking about the core meaning behind the sentence, not the sentence itself. If we misinterpret the sentence, then we haven’t yet gotten on to the claim being made and hence probably don’t fully understand the argument. Getting clear on just what an argument says is critical to the dialectical process.

\subsection{Simplified Breakdown}

Even if you are exceptionally bright, you probably found the last couple paragraphs rather challenging. That’s OK. You might work through them again more carefully and come back to it in a day or two if it’s still a struggle. The path to becoming a better critical thinker is more like mountain climbing than a walk in the park, but with this crucial difference: no bones get broken when you fall off an intellectual cliff. So you are always free to try to scale it again. We can sum up the key points of the last few paragraphs as follows:
\begin{earg}
    \item[] We use sentences, bits of language, to express propositions.
    \item[] The proposition, what is meant by the sentence, represents the world as being some way.
    \item[] The proposition is true when it represents the world in a way that corresponds to how the world is.
    \item[] Truth, understood as correspondence between a claim (a proposition) and the way the world is, is not relative to meaning, knowledge, belief, or opinion.
\end{earg}
Hopefully we now have a better grip on what it is for a claim to be true. A claim is true just when it represents things as they are. As is frequently the case in philosophy, the real work here was just getting clear on the issue. Once we clearly appreciate the question at hand, the answer seems pretty obvious. So now we can set aside the issue of what truth is and turn to the rather different issue of how to determine what’s true.

\subsection{The connection}

Arguments are how philosophers, and scientists, and the rest of us get at the truth. We are using them to structure our reasons and prove that the conclusion is true.

\section{Part 1.2: Arguments}
\label{s:p1.2}


The common-sense, everyday, way to tell whether a claim is true or false is to look at the reasons for or against it. Sometimes our observations give us good reasons. For example, I have a good reason for thinking my bicycle has a flat tire when I see the tire sagging on the rim or hear air hissing out of the tube. But often the business of identifying and evaluating reasons is a bit more involved. Logic is the business of identifying and evaluating reasons. You do this all the time. You give yourself reasons to choose one shirt over another for a job interview, you reason your way through making a choice about which classes to take, you believe certain things based on evidence, and you give reasons for why something isn't your fault (when you make excuses). In all of these cases and more, you are making \textbf{arguments}.

This is very different from how you might use the term ‘argument'. In everyday language, we sometimes use the word ‘argument’ to talk about belligerent shouting matches. If you and a friend have an argument in this sense, things are not going well between the two of you. Logic is not concerned with such teeth-gnashing and hair-pulling. They are not arguments, in our sense; they are just disagreements. For a humorous example of how arguments and disagreements differ, take a look at this funny exchange from Monty Python:\autocite{ArgumentClinic}

\factoidbox{\begin{multicols}{2}
Man: Is this the right room for an argument?\\
Other Man:(John Cleese) I've told you once.\\
Man: No you haven't.\\
Other Man: Yes I have.\\
M: When?\\
O: Just now.\\
M: No you didn't!\\
O: Yes I did!\\
M: You didn't!
\ldots\\
O: Oh I'm sorry, is this a five minute argument, or the full half hour?\\
M: Ah! (taking out his wallet and paying) Just the five minutes.\\
O: Just the five minutes. Thank you.\\
O: Anyway, I did.\\
M: You most certainly did not!\\
O: Now let's get one thing quite clear: I most definitely told you!\\
M: Oh no you didn't!\\
O: Oh yes I did!
\ldots\\
M: Oh look, this isn't an argument!\\
(pause)\\
O: Yes it is!\\
M: No it isn't!\\
(pause)\\
M: It's just contradiction!\\
O: No it isn't!\\
M: It IS!\\
O: It is NOT!
\ldots\\
M: (exasperated) Oh, this is futile!!\\
(pause)\\
O: No it isn't!\\
M: Yes it is!\\
(pause)\\
M: I came here for a good argument!\\
O: AH, no you didn't, you came here for an argument!\\
M: An argument isn't just contradiction.\\
O: Well! it CAN be!\\
M: No it can't!\\
M: \emph{An argument is a connected series of statements intended to establish a proposition.}\\
O: No it isn't!\\
M: Yes it is! ‘tisn't just contradiction.\\
O: Look, if I \emph{argue} with you, I must take up a contrary position!\\
M: Yes but it isn't just saying ‘no it isn't'.\\
O: Yes it is!
\ldots\\
M: No it ISN'T! \emph{Argument is an intellectual process.} Contradiction is just the automatic gainsaying of anything the other person says.\\
O: It is NOT!\\
M: It is!
\ldots 
\end{multicols}}

An argument is a reason for taking something to be true. \Glspl{argument} are made out of two or more claims, one of which is a \gls{conclusion}. The conclusion is the claim the argument purports to give a reason for believing. The other claims are the \glspl{premise}. The premises of an argument taken together are offered as a reason for believing its conclusion. In the above exchange, the person looking for an argument claims that "An argument is a collected series of statements to establish a definite proposition". This is a close approximation to how philosophers use the term, but a better, more exact definition is:


\newglossaryentry{argument}
{
name=argument,
description={A connected series of sentences, divided into \gls{premise}s and \gls{conclusion}},
plural=arguments
}

\newglossaryentry{premise}
{
name=premise,
description={A sentence in an \gls{argument} other than the \gls{conclusion}, often indicated with a premise indicator},
plural=premises
}

\newglossaryentry{conclusion}
{
name=conclusion,
description={The sentence which an argument is intended to support or prove, often indicated with a conclusion indicator.},
plural=conclusions
}

\begin{center}
An argument is a collected series of propositions intended to establish others in the series.
\end{center}
The propositions (statements, in this case) which support others are our premises. The one being supported is the conclusion. That 'intent' bit is a heavy hitter as we will soon see.

Some arguments provide better reasons for believing their conclusions than others. In case you have any doubt about that, consider the following examples:

\begin{tabular}{p{1in}|p{1in}}\hline
Example \exarg{linecook1}: &Example \exarg{linecook2}:\\\hline
Sam is a line cook. &Sam is a line cook.\\
Line cooks generally have good kitchen skills. &Line cooks generally aren't paid well.\\
Therefore, Same can probably cook well &Therefore, Sam is probably a billionaire. 
\end{tabular}

\begin{tabular}{p{1in}|p{1in}}\hline
Example \exarg{boston3}: &Example \exarg{boston4}:\\\hline
Boston is in Massachusetts. &Boston is in California.\\
Massachusetts is east of the Rockies.&California is west of the Rockies.\\
Therefore, Boston is east of the Rockies.&Therefore, Boston is west of the Rockies.
\end{tabular}

The premises in Example \ref{linecook1} provide pretty good support for thinking Sam can cook well. That is, assuming the premises in the first argument are true, we have a good reason to think that its conclusion is true (at this stage, we are not interesting in whether the premises are actually true, only the structure). Whether or not the argument is any good depends on how well they establish the conclusion, relative to the intent. So, we can say that the reasoning in Example 1 is pretty good (at least). The premises in Example \ref{linecook2} give us no reason to think Sam is a millionaire, let alone a billionaire. So whether or not the premises of an argument support its conclusion is a key issue.

Looking at Examples \ref{boston3} and \ref{boston4}, we see some similarities. The intent is different (in the first two, the goal was to get something with likelihood, in these, the intent is to get something with certainty), but we seem not to like the second (Example \ref{boston4}) and the first seems OK (Example \ref{boston3}). The main problem with Example \ref{boston4} is very different than the problem with \ref{linecook2}. And, in fact, the issue is an entirely different animal. With Example \ref{linecook2}, the issue was with the structure, here, the issue is with the truth. Notice, the structure of the arguments, Examples \ref{boston3} and \ref{boston4}, are exactly the same, there's no difference in how they are being reasoned.  If its premises were true, then we would have a good reason to think the conclusion is true (the best reason, 100\% certainty reasoning). That is, the premises do support the conclusion. But the first premise of the second argument just isn’t true. Boston is not in California. So the latter pair of arguments suggests another key issue for evaluating arguments. Good arguments have premises which support the conclusion, the best arguments have true premises.

That is pretty much it. The best arguments are those which have true premises that, when taken together, support its conclusion. So, evaluating an argument involves just these two essential steps:
\begin{enumerate}
    \item Determine whether or not the premises are true.
    \item Determine whether or not the premises support the conclusion (that is, whether we have grounds to think the conclusion is true if all of the premises are true).
\end{enumerate}
Often, figuring out whether the premises of an argument are true involves looking at further arguments for those premises individually. An argument might be the last link in a long chain of reasoning. In this case, the quality of the argument depends on the whole chain. Really high-quality philosophy papers (I don't expect these in this class) often involve an argument and further arguments for each of the premises. And since arguments can have multiple premises, each of which might be supported by further arguments, evaluating one argument might be more involved yet, since its conclusion is really supported by a rich network of reasoning, not just one link and then another. While the potential for complication should be clear, the basic idea should be pretty familiar. Think of the little kid who would constantly ask their parent “why?” after being an explanation. Even at a young age we understood that the reasons for believing one thing can depend on the reasons for believing a great many other things.

However involved the network of reasons supporting a given conclusion might be, it seems that there must be some starting points. That is, it seems there must be some reasons for believing things that don’t themselves need to be justified in terms of further reasons. There needs to be some sort of bedrock, ground level, foundation at the bottom. Otherwise the network of supporting reasons would go on without end, and the kid would happily ask 'why' until the end of time. The problem here is about how do we tell where the ultimate foundations of knowledge and justified belief. This is a big epistemological issue and we will return to it later in the course. For now, let’s consider one potential answer we are already familiar with. In the sciences, our complex chains of reasoning seem to proceed from the evidence of the senses.

We think that evidence provides the foundation for our edifice of scientific knowledge. Sounds great for science, but where does this leave philosophy? Does philosophy entirely lack evidence on which its reasoning can be based? Philosophy does have a kind of evidence to work from and that evidence is provided by philosophical problems. When we encounter a problem in philosophy this often tells us that the principles and assumptions that generate that problem can’t all be correct. This might seem like just a subtle clue that leaves us far from solving the big mysteries. But clues are evidence just the same. Sensory evidence by itself doesn’t tell us as much about the nature of the world as we’d like to suppose. Scientific evidence provides clues, but there remains a good deal of problem solving to do in science as well as in philosophy.

So we can assess the truth or falsity of the premises of an argument by examining evidence or by evaluating further argument in support of the premises. Now we will turn to the other step in evaluating arguments and consider the ways in which premises can support or fail to support their conclusions. The question of support is distinct from the question of whether the premises are true. When we ask whether the premises support the conclusions we are asking whether we’d have grounds for accepting the conclusion assuming the premises are true. In answering this question we will want to apply one of two standards of support: deductive validity or inductive strength.

\subsection{Part 1.2.1: Deductive Validity}
\label{s:p1.2.1}
There are two kinds of arguments which we deal with, one is \glspl{deductive argument} and the other is inductive arguments, here we are going to be concerned with deduction. When you are dealing with this kind of argument, the standard for the 'goodness' of the argument is validity. An argument counts as deductive whenever it is aiming at this standard of support. Deductive validity is the strictest standard of support we can uphold. It is also the one which is used in the vast majority of philosophy (with the exception of some, very contemporary, fields).\footnote{This would be Experimental Philosophy, which involves testing philosophic intuitions. This can be useful when you are trying to get a common person, average Joe view.} In a deductively \gls{valid} argument, the truth of the premises guarantees the truth of the conclusion. This standard is not concerned with whether the premises are actually true, that is a different standard. Basically, if you assume that the premises are true, the conclusion must be true. Here are two equivalent definitions of deductive validity:

\newglossaryentry{valid}
{
name=valid,
description={A property of arguments where the truth of the premises guarantee the truth of the conclusion; i.e. it is impossible for the premises to be true and the conclusion false}
}

\begin{tabular}{p{0.5in}|p{3in}}
D &A valid argument is an argument where if its premises are true, then its conclusion must be true.\\\hline
D* &A valid argument is an argument where it is not possible for all of its premises to be true and its conclusion false.
\end{tabular}

That (D*) standard is the more formal way, and exact way, of claiming (D). Here are a few examples of deductively valid arguments:

\begin{tabular}{p{1in}|p{1in}|p{1in}}
Example \exarg{socrates1} &Example \exarg{mammals2} &Example \exarg{raining3}\\\hline
If Socrates is a man, then Socrates is mortal &All primates are mammals &If it's wet outside, then it's raining\\
Socrates is a man &All humans are primates &It's not raining\\
Therefore, Socrates is mortal &Therefore, all humans are mammals &Therefore, it's not wet outside.
\end{tabular}

If you think about these two examples for a moment, it should be clear that there is no possible way for the premises to all be true and the conclusion false. The truth of the conclusion is guaranteed by the truth of the premises. \ref{raining3}, on the other hand, should have given you pause. Yes, that argument is valid, but it's not \gls{sound}. Soundness involves whether or not the premises are in fact true, validity concerns the structure of the argument. If an argument is sound, then it's valid, but not the other way around. In contrast, the following arguments are not valid:

\newglossaryentry{sound}
{
name=sound,
description={A property of arguments that holds if the argument is valid and has all true premises}
}


\begin{tabular}{p{1in}|p{1in}|p{1in}}
Example \exarg{socrates4} &Example \exarg{cookies5} &Example \exarg{raining6}\\\hline
If Socrates the Cat is a man, then Socrates the Cat is mortal &Billy or Sally stole cookies from the jar &If it's wet outside, then it's raining\\\hline
Socrates the Cat is mortal &Billy stole cookies &It's not wet outside\\\hline
Therefore, Socrates the Cat is man &Therefore, Sally didn't steal cookies &Therefore, it's not raining
\end{tabular}

These examples, again, are not valid, they have a flaw in their reasoning, in arguing this way, you will have an error which will lead you astray. To see why, it might require a bit of imagination, but it's a reasonably simple test. Imagine a case where the premises are true and the conclusion is false. If you can't do it, then the argument is valid, if you can, then there's a flaw in the reasoning. Think of this test as like a trial by worst-case scenario. For Example \ref{socrates4}, Socrates the Cat is obviously mortal, and \emph{were} the cat a man, he still would be mortal, but that doesn't mean that a cat is a man. For Example \exarg{cookies5}, it's perfectly possible that both Billy and Sally were naughty, bad, kids and stole some cookies from the jar, just because one did it doesn't mean that the other didn't.\footnote{Some languages, like Latin, have two different words to express the different kinds of disjunctions (or-statements). There are inclusive disjunctions, which allow for both to be true and there are exclusive disjunctions, which only allow for only one of them to be true. English uses both but does not have a built in way to tell the difference between them consistently. Context is your best bet for this, but also, sometimes 'either... or...' is used for exclusive and just '... or...' is used for inclusive.} For example \ref{raining6}, this might be easier to imagine for people who have spent time in the desert: In those sorts of regions, there's a phenomenon which I know as sun-showers. This is where the sun is still shining, the ground is perfectly dry, but there's rain coming down from the sky. So, it's perfectly possible for it to be raining and yet not be wet outside.
\factoidbox{
Deductive validity is the gold standard for an argument. Soundness is the platinum standard.}

The deductive arguments we’ve looked at here are pretty intuitive. We only need to think about whether the conclusion could be false even if the premises were true. But most deductive arguments are not so obvious.These example arguments only use one logical rule and only two supporting premises; most of the ones which are really powerful use several logical rules/forms in conjunction with each other and several more lines of supporting evidence.  Logic is the science of deductive validity. For a class on this subject, check out PHIL\&120, which is the barebones, no fluff, mathematics of Philosophy. Philosophy has made some historic advances in logic over the past few centuries, with great advancements happening in the last century.  If you want to see some of the practical side of their good work, what it's actually been used for, just look at your computer. The background coding for that machine, I know because I worked with it for a time and I know the history behind it, is all based on the logical structures which philosophy has used forever.

\subsection{Part 1.2.2: Inductive Strength}
\label{s:p1.2.2}

|n the section, we looked at deductive arguments and their standards are validity and soundness, but those aren't the only kinds of arguments which are used, other fields use a different kind of argumentation, \glspl{inductive argument}. The standard for the goodness here is strength, not validity. Like with deductive arguments, an argument counts as inductive when it's shooting for this kind of support for the conclusion. Inductive strength is a weaker standard of support we can shoot for. That being said, it's also the standard which is found in science.\footnote{As well as in Experimental Philosophy.} An inductively strong argument is one where the premises make the conclusion more likely, give probabilistic support. Strength, again, is not concerned with the truth of the premises, that's a different standard still. Here are some examples of inductive arguments:

\newglossaryentry{deductive argument}
{
name=deductive argument,
description={An argument such that the intent behind it is to guarantee the conclusion. There is no ‘wiggle room'. In general, deductive arguments move from broad general principles and to more particular instances.},
plural=deductive arguments
}

\newglossaryentry{inductive argument}
{
name=inductive argument,
description={An argument such that the intent behind it is \emph{merely} to make the conclusion more likely, assuming the truth of the premises. There is some ‘wiggle room'. In general, inductive arguments move from a collection of particular instances and then use those to support the truth of some general principle},
plural=inductive arguments
}


\begin{tabular}{p{1in}|p{1in}|p{1in}}
Example 1: &Example 2: &Example 3:\\\hline
Sam is a line cook &Most things cancerous to mice are cancerous to humans &Oscar was born in North America\\
Line cooks generally can cook well &Various chemicals in tobacco products are cancerous to mice &Oscar was not born in Mexico\\
Therefore, Sam can probably cook well &Therefore, various chemicals in tobacco products are probably cancerous to humans &Therefore, Oscar was probably born in the USA
\end{tabular}

Examples 1 and 2 seem like pretty decent arguments. The premises do support the conclusion (we are looking at inductive strength here). But, none of the above arguments are, in fact, valid. It's perfectly possible for the premises to be true and the conclusion false. Sam could be a brand new cook hired because he’s the manager’s son who has never cooked in his life. The chemicals in tobacco products could be cancerous to mice but not humans.\footnote{In this case, we can only look at the information provided in the premises, same with validity, so if you disagree with this sentence, you would be correct, but that's bringing information not given as support in the argument.} Many arguments give us good reasons for accepting their conclusions even if their premises being true fails to completely guarantee the truth of the conclusion.  The intent behind all of these arguments is different. The point of these arguments is not to guarantee the conclusion, but rather to make them more likely to be true.  We judge how good such an argument is according to how strong it is.  Unlike validity, strength is a bit more wishy-washy, it comes in degrees, which is why I use the terms 'probable' and 'improbable' in the below definitions:

\begin{tabular}{p{0.5in}|p{3in}}
I &An inductively strong argument is an argument in which if its premises are true, its conclusion is probably true.\\
I* &An inductively strong argument is an argument in which it is improbable that its conclusion is false given that its premises are true.
\end{tabular}

As in the case of validity, when we say that an argument is strong, we are only claiming that if the premises are true then the conclusion is likely to be true. Corresponding to the notion of deductive soundness, an inductive argument that is both strong and has true premises is called a cogent inductive argument. Unlike the case with deductively sound arguments, it is possible for an inductively cogent argument to have true premises and a false conclusion. When we are asking about the validity of an argument, we are asking whether it's possible for the premises to be true and the conclusion false. When we are asking about the strength of an argument, we are asking about the probability of the conclusion being false if we assume that the premises are true.  Possibility does not come in degrees, it's either possible or impossible. Probability does come in degrees. In the simplest case, inductive reasoning involves inferring that something is generally the case from a pattern observed in a limited number of cases.\footnote{One way to think about this is that deduction goes from general to particular and induction goes from particular to general.}

    Suppose we conducted a poll of 1000 Seattle voters. The results showed that 600 of them claimed to be Democrats. We could inductively infer that 60\% of the voters in Seattle are Democrats. The results of the poll give a pretty good reason to think that around 60\% of the voters in Seattle are Democrats. But the results of the poll don’t guarantee this conclusion. It is possible that only 50\% of the voters in Seattle are Democrats and Democrats were, just by luck, over represented in the 1000 cases we considered, but it may not be probable.

There are a few factors which tell us how strong an inductive argument is. One is how much evidence we have looked at before inductively generalizing. Our inductive argument above would be stronger is we drew our conclusion from a poll of 100,000 Seattle voters, for instance. And it would be much weaker if we had only polled 100. Similarly, if we were trying to figure out the political stances of all of Washington but solely looked at Enumclaw, we would get a radically different view about our standings than if we had looked at Seattle and Enumclaw.

Also, the strength of an inductive argument depends on how much of the evidence represents the amount in reality. So our inductive argument will be stronger if we randomly select our 1,000 voters from the Seattle phone book than if they are selected from the Ballard phone book (Ballard being a notably liberal neighborhood within Seattle).

So far, we’ve only discussed inductive generalization, where we identify a pattern in a limited number of cases and draw a more general conclusion about a broader class of cases. Inductive argument comes in other varieties as well. In the example we started with about Sam the line cook, we inductively inferred a prediction about Sam based on a known pattern in a broader class of cases. Argument from analogy is another variety of inductive reasoning that can be quite strong. The strength here is in how many commonalities the two cases have, we will see an argument like this in the section concerning the existence of God.

\section{Part 1.3: Common Argument Structures In Philosophy}
\label{s:p1.3}

Since Philosophy relies on arguments to get at the facts about the world, like science, but we go with the gold-standard for arguments, validity, often arguments will come in certain forms, structures. These forms can be just on their own, or in combinations with others. If you structure your arguments in these ways or you make arguments out of combinations of them, you are guaranteed to have a valid argument, soundness is a different story. 

\subsection{Part 1.3.1: Modus Ponens}
\label{s:p1.3.1}

This is another easy and intuitive argument structure. It follows a very similar model to cause-and-effect. If I knock over the glass, it will break; I knocked over the glass; so it broke. Here, I have an if-then statement, which can be phrased in several different ways, I affirm the antecedent (the if-part), and thereby get the consequent (the then-part). But, that doesn't mean it's fool-proof. For example, there's a very common logical fallacy called "affirming the consequent", this is where you have an if-then statement, you affirm the then-part, and thereby think you get the antecedent (the if-part). This is not valid, it's not a good structure. For example, take this quick argument (which is Modus Ponens done properly):

\factoidbox{If my car won't start, then there's something wrong with the battery terminals. My car won't start. So, there must be something wrong with the battery terminals.}

Using cause-and-effect again, there can be many different possible causes for some event. One of them definitely caused the event, but you can't say with certainty which one caused it just because it happened. For example, if an apple is too heavy, it will fall off a tree. Yes, when an apple is too heavy it will fall from a tree, but loads of other things can cause it to fall, so you can't say that it's weight caused it to fall, it could have been the wind.

\subsection{Part 1.3.2: Modus Tollens}
\label{s:p1.3.2}

Modus Tollens is like Modus Ponens in reverse and negated. In this case, you have an if-then statement, you have that the then-part is false, and so you get that the if-part must be false. This makes sense in the case of cause-and-effect too. You know that one thing would cause another and you know that that other thing didn't happen, so you know that the first thing didn't happen. For example, if I sleep in, I will be late, I wasn't late, so I must not have slept in. 

Again, this is a pretty simple idea, but it's easily misused. The fallacy here is called 'denying the antecedent'. This, again, is not good reasoning; you have an if-then statement, you have that the if-part is false, so you think that the then-part must be false. But, just because something wasn't caused by one thing, it doesn't mean that it wasn't caused by another. For example, if God exists, then humans exist; God doesn't exist; therefore, man doesn't exist. Poof, all atheists  just disappeared, right? No, they didn't. This is not a valid argument. 

Modus Tollens, properly, we will mostly see in this class, as it's the easiest for the sort of teaching style which I employ. For example, take the following:

\factoidbox{Ethical Egoism is the stance that the morality of an action is determined by how much it benefits the doer personally. By this theory, an action is moral when it benefits me and immoral when it does not. So, if Ethical Egoism is correct, then donating to charity, which will in no way benefit me, is always morally wrong. But, it seems obvious that the most moral actions we can do are selfless (in this case, donating when it won't benefit me). So, Ethical Egoism is incorrect.}
    
\subsection{Part 1.3.3: Disjunctive Syllogism}
\label{s:p1.3.3}

This is by far the most simple argument structure used in Logic and it's probably the hardest to misuse (think that you have a valid argument, but don't). This does not mean that I have seen bad formulations of this which I will show, but first, here is a story:
\factoidbox{Years ago, while I was living in Arizona, I had some car trouble. Periodically, my car just would not start, or it spontaneously let me start it after a few attempts, seemingly randomly. I eventually called my local mechanic and had them take it to a shop. The mechanic, maybe knowing I was a philosophy professor, said the following (and this is a real quote): "Either it's your ECM or it's your fuse-box. We tested and it's not your fuse-box. So, the issue must be your ECM."}

Basically, a disjunctive syllogism takes 2 possible options, has that one of them is false, and thereby gets the other. This can be used in various different ways, for example, you could have 3 possible options, show that one of them is false and thereby get that it must be one of the other 2. This is a sort of process of elimination sort of argument structure. This is a very simple, easy, argument structure, but it's still possible to structure it poorly. For example, I have seen people take two options, show that one of them is true and thereby claim that the other is false. This is not a good structure. Disjunctions, in English, without context, don't allow for this sort of move. It is possible for both options to be true. There are some cases where this works, but to play it safe, and stay on the windy side of validity, assume that both can be correct. 

\subsection{Part 1.3.4: Hypothetical Syllogism}
\label{s:p1.3.4}

This is the next, basic, argument structure which I will cover with you here. This argument structure does not require you to have anything other than conditionals, you don't need any proven facts. But, at the end, you don't get any proven facts out either, all you get are conditionals. We often use this sort of reasoning when we are thinking about what will happen next if we go through some possible scenario. For example, take the following:

\factoidbox{If you give a mouse a cookie, he will ask for a glass of milk. If he asks for a glass of milk, then he will ask for a straw. Therefore, if you give a mouse a cookie, then he will ask for a straw.} 

When the antecedent of one conditional is the same as the consequent of another, you can collapse them into one conditional by taking out the middle man. This is a driving force for the slippery slope fallacy, but in that case, the fallacy is not in the reasoning itself, but rather in the truth of the conditionals which it employs. Often in arguments, this is used to shorten the work of Modus Ponens or Modus Tollens (below), but it can be used all on its own to get a point across. 

\subsection{Combining the Structures}

Arguments aren't always merely 3 lines long, sometimes they are far longer and more involved. For example, take the following perfectly valid argument:

\factoidbox{If I eat a ton of food this Thanksgiving, then I will get yelled at by my doctor. If I get yelled at by my doctor, then I will need to work out. I did eat a ton of food this Thanksgiving. Therefore, I will need to work out.}

In this case, I am using several different structures together to get my conclusion. I am using Hypothetical Syllogism and Modus Ponens all at once to get my answer. Similarly, I can have an argument like this:

\factoidbox{If I need to get a math credit, then I will either need to take PHIL\&120 or a college level math course. I need to get a math credit and I really don't want to take a college level math course, so I will need to take PHIL\&120.} 

Here, I am using both Modus Ponens and Disjunctive Syllogism to get the conclusion. And yes, PHIL\&120 is a math credit, but it can be more difficult than other options. All of these examples come from PHIL\&120. 

\section{Part 1.4: Fallacies and Biases}
\label{s:p1.4}

A fallacy is just a mistake in reasoning. Humans are not nearly as rational as we’d like to suppose. In fact we are so prone to certain sorts of mistakes in reasoning that philosophers and logicians refer those mistakes by name. For now I will discuss just one by name but in a little detail. Watch for explanations of other fallacies over the course of the class. For pretty thorough catalogue of logical fallacies, I’ll refer to you The Fallacy Files.\autocite{Curtis1} There is also a section in the appendix of this textbook which lists and explains various informal fallacies. Formal fallacies are cases where the structure of the argument seems fine, but it actually relies on an improper move (outlined when I covered the common structures). Informal fallacies are when improper or incorrect reasons are used in the argument, which are outlined in the module I linked to. Here are some examples of fallacies which are rather egregious and should never be seen in any work, in philosophy or elsewhere: 
\subsection{Ad Hominem Fallacy}

“Ad hominem” is Latin for “against the man.” It is the name for the fallacy of attacking the proponent of a position rather than critically evaluating the reasons offered for the proponent’s position. The reason ad hominem is a fallacy is just that the attack on an individual is simply not relevant to the quality of the reasoning offered by that person. Attacking the person who offers an argument has nothing to do whether or not the premises of the argument are true or support the conclusion. Ad hominem is a particularly rampant and destructive fallacy in our society. What makes it so destructive is that it turns the cooperative social project of inquiry through conversation into polarized verbal combat. This fallacy makes rational communication impossible while it diverts attention from interesting issues that often could be fruitfully investigated.

\factoidbox{Here is a classic example of ad hominem: A car salesman argues for the quality of an automobile and the potential buyer discounts the argument with the thought that the person is just trying to earn a commission. There may be good reason to think the salesman is just trying to earn a commission. But even if there is, this is irrelevant to the evaluation of the reasons the salesman is offering. The reasons should be evaluated on their own merits.}

Notice, it is easy to describe a situation where it is both true that the salesman is just trying to earn a commission and true that he is making good arguments. Consider a salesman who is not too fond of people and cares little for them except that they earn a commission for him. Otherwise he is scrupulously honest and a person of moral integrity. In order to reconcile himself with the duties of a sales job, he carefully researches his product and only accepts a sales position with the business that sells the very best. He then sincerely delivers good arguments for the quality of his product, makes lots of money, and dresses well. This salesman must have been a philosophy major. The customer who rejects his argument on the ad hominem grounds that he is just trying to earn a commission misses an opportunity to buy the best. The moral of the story is just that the salesperson’s motive is logically independent of the quality of his argument.
\subsection{Strawman Fallacy}

The strawman fallacy is one which I encounter occasionally in student works but you really will see it commonly in the political sphere (much to our misfortune, regardless of the side you hold). A strawman is like a scarecrow, a mock façade of a person used as a distraction or a trick. The strawman in the fallacy is not a façade of the person making the argument or claim, per se, but rather it is a façade of the claim or argument. Often, when we are explaining a view or defending our own views, we will need to explain the opposing side and give their arguments for their stance. This is so that we will be able to explain why our own is more reasonable or better. Putting your opponent's stance in your own words is not strawmanning, in fact I encourage you to (because then you will need to put yourself in their shoes and potentially find flaws in your own stance). This becomes strawmanning, however, when the argument or stance which you are attributing to your opponent is not their argument or stance. You have left out key details, misrepresented their findings or claims, or imposed your own stance (which they reject) onto them in order to make them seem absurd. Strawmanning another person's stance doesn't only harm your quest for the right answer to a problem (because you have removed the view of a person (who I am assuming is reasonable) from the discussion) but it also harms others who could take what you say seriously. In the case of politics, if the party or group you support presents the opposing side as having absurd or outlandish views which could never work or have some fundamental flaw which anyone could see, often, rather than thinking that they must have missed something, you will take it as gospel and move on to spread this misinformation. The vast majority of the time, however, the absurd or outlandish view was actually a strawman of the original. The original, if the people behind it are reasonable, would lack the fundamental flaw or have some explanation about how the flaw was handled in connection with other policies. 

The way to avoid this fallacy is to paint all arguments for any claim in the best possible light. If there is a subtle flaw in the argument which could easily be overlooked, make the patch yourself (even and especially if you disagree with the conclusion to it). For example, in this class, when we talk about the first cause argument for the existence of God, I present the argument without a 'fatal' flaw which is seen in the ordinary presentation of it because it was easy to see and fix; I did not want to present you with a strawman.  If a stance seems absurd, do some digging into the actual stance and see whether the person presenting it to you missed or intentionally left out something. For example, suppose that a study showed that some program, which would remove a service from the private sector and made it public (have the government provide the service rather than a collection of private companies), would cost the tax payers 2.6 trillion USD. They present this as an absurd amount of money and then show how much your taxes would increase by. What they fail to mention, however, is that the current cost for that program on tax payers is closer 3.6 trillion USD, but rather than this money going to the government, it goes to the private companies. Others may strawman the cost by inflating it by ignoring the basic fact that many middlemen would not exist in the proposed system and the lack of them would reduce the cost. Doing a little deeper digging will save you from this fallacy. 

\section{Part 1.5: Critical Thinking}

This page roughly finishes up our crash course on logic and critical thinking, but we will be seeing more logical forms and critical thinking structures as the course progresses. Most of it was dedicated to logic, and if you are more interested in that, I strongly recommend that you take PHIL\&120. But, for good critical thinking, there are some further standards which are worth noting. These standards follow from the others which we have discussed, but are a wee-bit more practical. These standards are Consistency and Coherency, but there are other aspects which come in when philosophy goes empirical.

\subsection{Consistency}

The first, basic, bare-bones standard for good critical thinking, and thereby good philosophy, is that the thinking must be consistent. This is more basic than saying that it's logical or that it lines-up with reality. For some train of thought to be consistent, it needs to lack two different things. First, it can't have any contradictions. A contradiction is a case where some proposition must be both true and false (at the same time) in order for your reasoning to work. In everyday life, it is (hopefully) very rare for us to encounter a contradiction either in our own or someone else's thinking, but there are cases where they show up, such as in conspiracy theories, and we only notice them well after the stance was explained. Take, for example, a conspiracy theory which has that both the Earth is the center of the universe and that gravity is just the Earth moving upwards at an ever increasing rate. This might not seem too contradictory on the face of it, but once you start digging into what must be true for both of those statements to be true, you will find that some proposition must be both true and false at the same time. There are other more blunt examples, such as:

\begin{tabular}{p{1.5in}|p{1.5in}}
Contradictory &Non-Contradictory\\\hline
The non-existent ghosts stole the painting &The ghosts stole the painting\\\hline
Bobby Joe committed the crime in New York (in person) while in LA. &Bobby Joe made it seem like he was in LA in order to cover up the crime.\\\hline
No one drives in New York because there is too much traffic.\footnote{This is from the Futurama episode "The Lesser of Two Evils", season 2, episode 11.} &The traffic in New York is so bad that most people walk.\\\hline
All animals are equal, but some are more equal than others.\footnote{This is from George Orwell's Animal Farm.} &Either all animals are equal or they aren't all equal.\\\hline
I make my own choices, with my wife's permission. &What I ultimately choose should be a joint decision between my wife and I.\\\hline
\end{tabular}

The Non-Contradictory column is there to show you how, most of the time, there is an easy way to resolve the contradiction, but there are other cases, which we will see in this class, where the contradiction is more deep seeded and can't be so easily resolved (like the contradictions in Moral Relativism). In order to avoid these, make sure that all of the propositions you are using are consistent, that they are always true or false (depending) throughout your thinking.

The second requirement for consistent thinking is that there aren't any equivocations. This is where you use the same statement or word/phrase in two or more parts of your reasoning and your reasoning relies on them meaning the same thing, but they don't mean the same thing in the two different instances. For example, take a look at these arguments:
\begin{earg}
   \item[] Nothing is better than God.
    \item[]A cheese sandwich is better than nothing.
    \item[]Therefore, a cheese sandwich is better than God.
\end{earg}
In this argument, the term 'nothing' is being used in two different ways. In order for us to accept the two lines of this argument, individually, we have different meanings in mind. In the first case, 'nothing' means that something along the lines of 'it is not the case that there is exists a being such that that being..." or "no being". So, the proper way to understand this first line is "it is not the case that there exists a being such that that being is better than God." The second line of the argument uses a different sense of the word 'nothing', namely "not having anything", so the correct way to understand this is that it means "a cheese sandwich is better than not having anything." Taking these two together, the argument looks like this:
\begin{earg}
    \item[]There isn't anything better than God.
    \item[]Having a cheese sandwich is better than not having anything.
    \item[]Therefore, ...
\end{earg}
As you should be able to see, the flow of the argument isn't there any more, the equivocation was the glue holding it together. Some other examples can be in quick reasoning like:
\begin{earg}
    \item[]André the Giant is so called because of his great size.
    \item[]André René Roussimoff is so called because that's what his mother named him.
    \item[]André the Giant is André René Roussimoff.
    \item[]Therefore, André René Roussimoff is so called because of his great size.
\end{earg}
That last line should seem off to you. This is because there is equivocation in this reasoning. In the first two lines we are talking about the names, but in the third we are talking about what the names pick out in the world. The equivocation is in the shift between talking about the word and the person.

\subsection{Coherency}

The next basic requirement for good critical thinking is that it needs to be coherent. Coherency is not the same as consistency, though there is a fair bit of overlap. Each premise, each proposition used in the reasoning need to relate to each other in a reasonable way. There can't be any strange or unorthodox jumps in the argument. You can think of this like a spider web, each of the strands is connected together and it would not be as strong if some of those strands were removed. For example, take this argument:
\begin{earg}
    \item[]If the moon landing was fake, then the Government did so to deceive us.
    \item[]If the Government did so to deceive us, then it was to make us lose faith in our religion.
    \item[]The moon landing was fake.
    \item[]Therefore, the Government faked it in order to make us lose faith in our religion.
\end{earg}
So, I am not going to make an argument that the moon landing wasn't fake (we really did go to the moon), rather, assuming that it was fake, it does make sense that it would have been faked to deceive us in some way. The real issue with this argument is in the second line. Does it really make sense that the Government would deceive us to make us lose faith in this way? There is a jump in the reasoning. Much more evidence and premises are required to show a connection between the assumption that the Government deceived us with the moon landing and that the purpose of it had something to do with religion. There are many other closer, more relevant potential reasons which need to be discounted first. For example, that the Government did so in order to reaffirm a sense of exceptionalism.

\subsection{Empirical Thinking}

There was a time in which most philosophic thought was empirical, but as the sciences diversified, that became less common (though the trend is swinging back that way in philosophy). Empirical thinking is reasoning which relates to or explains the outside world. In the other sciences, you may have heard a distinction between empirical (applied) and theoretical. Theoretical science is the arm-chair, hypothetical models which are thought about and debated and later tested using empirical methods (like experiments). Empirical science is the science which actually does the tests and uses the materials in question. There are two features to reasonable empirical thinking, which the other sciences should take note of and explain from the beginning. These are that the thinking must be adequate and applicable.

Empirical thinking is adequate when all of the cases you are trying to explain are accounted for. There shouldn't be too many exceptions to the account you are giving and those exceptions should be easily accounted for. For example, although this story is apocryphal, when Galileo presented his findings about there being objects which orbit something other than the Earth, one person claimed that it was because of interference in the telescope. So, they tried it in a different location, and they got the same result. Eventually, the objector said something like "my hypothesis works so long as you don't look through a telescope". The hypothesis that all objects orbit the Earth wasn't adequate because there were cases which it could not explain and those exceptions could not be easily accounted for. 

Empirical thinking is applicable when there isn't anything in the explanation which doesn't relate back to experience and evidence or data gained from testing in the relevant environment and the explanation/hypothesis is useful (as in, it leads or gives way to further understanding and can be used to make other explanations). Many examples of empirical thoughts failing in this regard can be found when you look at the claims and tests involving pyramid power, magic crystals, feng shui, various vitamins and supplements seen on Dr. Oz, and even copper/magnetic bracelets claimed to treat arthritis. The explanations for each of these contain claims which either can't be demonstrated with experiments or don't relate to experience. For a more scientific example, take this quote from the famous physicist Richard Feynman:
\factoidbox{
    The next question was — what makes planets go around the sun? At the time of Kepler some people answered this problem by saying that there were angels behind them beating their wings and pushing the planets around an orbit.\autocite{physicallaw}}

 The hypothesis that the angels are making the planets move isn't applicable because it doesn't relate back to experience and evidence. If, in some strange world, we could see the angels sweating and pushing the planets really hard, then it would be. However, we can't see the angels and if someone were to claim that they are there, they are just invisible, the hypothesis would be even more inapplicable.

\input{apology}

\chapter{Part 2: Life and Times of Socrates}

With the introduction to philosophy out of the way (as in, the barebones of the method and the quest for answers), we will now move on to the life and times of Socrates, the protagonist and real world person who you should be reading in the assigned reading for this module. This should serve as an ample starting point to understand the raise and fall of Philosophy's most famous proponent.
\section{Birth and Parents}

Socrates was born around spring 469BCE in Athens. At that time, the Persians had just attempted  (and failed) to invade Athens and Athens, on a global scale, was forming an alliance with the other city-states in the region (this alliance, called The Delian League, would grow into the Athenian Empire). This is on the Attic Peninsula and, politically and socially, it was divided into 139 districts, which were in turn, broken up among the 10 tribes which made up Athens. The members of these tribes were automatically Athenian. Socrates was a member of the Antiochis, which was located outside of the city walls, to the south-east. Keeping with the customs of the time, 5 days after Socrates' birth, his father, Sophroniscus, looked the infant Socrates over while walking around the hearth, to make sure that the kid was his, and then admitted him into the family. 5 days after this, Socrates was actually given a name and his father presented him to the local officials for the relevant paperwork (think of this as the equivalent of a birth-certificate and all that). In doing this, Sophroniscus took on the responsibility of ensuring that Socrates got a proper education and was made into a respectable member of society.
\section{Education}

In Athens, the ability to read and write was commonplace since around 520BCE and it was unheard of for a young man not to have that skill. When Socrates turned 5, he started the equivalent of elementary school. His education consisted of learning to read and write, gymnastics (it was expected that Athenians be physically fit for military service, think of this like PE), music, and basic mathematics. It is because of this that Socrates, frankly, became a nerd. Socrates, as to be expected, loved philosophy. At that time, philosophy was, and still is, the mother of all other subjects, physics, mathematics, biology, and so on. The term 'philosophy' comes from two Greek roots, 'philein' meaning love and 'sophia' meaning wisdom. For Socrates and other Greeks (as well as any good philosophy/science today), philosophy was done with unaided reason and careful observation of the world. Over time, the careful observation aspect became what we call the sciences. But, to do philosophy, there are certain things which you don't get. You don't get anything in a religious scripture (so I don't want to see you reference a religious text in this class), you don't get what some authority figure says (just because they are an authority), you don't get myths, and you don't get things just because they were always done that way (no tradition/cultural beliefs). For philosophy, all you get is your observations and your own good reason. At that time, philosophers were inventing Geometry, proposing a heliocentric model of the solar system (the Earth revolves around the sun), and also various aspects of the natural sciences which would develop into their own fields (EG biology).

Socrates, by all accounts, was a remarkably ugly person. Many accounts say that his childhood classmates would call him “frogface”. He had bulging eyes, broad, flat nose, and thick lips. He also walked bow-legged and sideways like a crab. If you ever see a bust of Socrates, understand that the carving is an angel compared to him in reality.
\section{"Adult" Socrates}

At the age of 17, Socrates graduated from the schools and was sworn in as a full-fledged citizen of Athens (think of it as like turning 18 in the US). Socrates' father took him to the ceremony, but died shortly afterwards. His mother, Phaenarete, remarried and had another son, named Patrocles.  In those days, Athens had one of the first constitutional democracies which enshrined freedom of speech and, because of this, public discussion, voting, and debate about all matters were quite commonplace. Athens had many festivals and gatherings throughout the year, attracting many of the great minds from around Greece. According to some accounts, at 19, Socrates was often found at these festivals discussing things with the philosophers of the day, allowable because of the freedom of speech. Freedom of speech is a definite perk which was not had by many societies in those days, but being an Athenian did come with some responsibilities. Upon becoming a citizen, Socrates was also put on the draft for military service.
\section{Military Career}

After being called on for the draft and completing the required 2 years of training, Socrates served in the military for Athens. During his first posting, it was a time of relative peace, so Socrates likely practiced a trade (stonecutting, like his father). But, as the years progressed, Athens was starting to get into a war with Sparta. Socrates served on the front lines for many of the battles in 432BCE,  Socrates was called on to serve in the military for Athens on a few different occasions. After putting down a revolt, Socrates' troop entered into a very heavy conflict near Spartolus, where they suffered heavy casualties. In this fight, Socrates' bravery became legendary. He refused to retreat until he was the last person there and fought off enemy soldiers with another soldier, Alcibiades, on his back, saving his life. This deployment kept him away from home for 3 years. Socrates returned to duty again in 424BCE, where his bravery was again noted by the generals. His commander, Laches, noted Socrates' bravery when writing about the nature of courage, stating that he refused to retreat, even after the order was given, until he was the last person to leave. A year later, Socrates returned to duty and fought in another battle. After this, as far as we can tell, he did not serve in the military any longer. After this point, Athens and Sparta signed a treaty, which gave Athens a few years without the struggles of war. During this time, Socrates married Xanthippe, who, because of Socrates' military exploits, came with a large dowry.
\section{Family Man}

Socrates cared little for material possessions. Often wearing the same clothes for many days in a row, including sleeping in them. While in the military, as well as throughout his life, Socrates rarely, if ever, wore footwear, even on ice or snow. Athenians of his day described him as “frugal”. Although it was well within his means to afford various things, like shoes, he practiced “voluntary simplicity”. Xanthippe gave birth to three sons, the third being born while Socrates was in prison awaiting execution, so they did the ceremony in the prison. According to many accounts, Xanthippe had a volatile personality and was very unhappy with her hubby, and with good reason, he had a habit of spending all day and night in the agora (public market space) discussing questions and arguing with people when he could be out there making money (he was a stonecutter, like his father). According to one story, after chewing him out, Xanthippe climbed onto the roof of their house and dumbed a bucket of urine on his head as Socrates went out to debate people.

Because of his willingness and constant engagement in questioning and debate, by his middle age, Socrates became a very recognizable (today we would say famous) person on the streets of Athens. In the Greek comedies of the day, an ugly caricature of Socrates was a re-occurring character. Socrates was made-fun of for his appearance and his love of philosophy in at least 3 prize-winning plays. Socrates took these in stride. At one point a foreigner asked “who’s this loon, Socrates?” in the middle of a play. At which point Socrates stood up and said “ME!”

\section{Part 2.1: Trials in Athens}

\section{How Trials Worked in Athens}

In Socrates' time, the procedure for a trial and court cases like this was pretty well established. In that time, also, there was no such thing as what we would call today a 'public prosecutor', this is similar to how in London, prior to their police force, they had a hew and cry system. If a citizen suspected another of a crime, then they would report it to the officials to have the case looked over. Any actual citizen of Athens could initiate the procedure. Once such a crime was reported, the trial consisted of three parts. First, there was the Initiation of Criminal Proceedings. Next, they had The Preliminary Hearing (Anakrisis). And third, they had The Trial itself. As evidence for how common and stream-lined this process was in Athens, historically, just prior to his trial, Socrates was engaged in a conversation with a prosecutor for a trial just ending. This conversation was on the nature of goodness and the question related to it is the discussion for this module. However, most cases were settled prior to reaching the actual trial because, as we will see, making the arrangements is quite the undertaking.
\subsection{Initiation of Criminal Proceedings}

As I just mentioned, any Athenian could raise charges against another. In the case of Socrates, the proceedings began when Meletus, a poet, arranged a meeting, for a specific date, with the legal magistrate, or in some cases King Archon, in a colonnaded building called the Royal Stoa to answer charges of impiety. Meletus, then delivered an oral summons to Socrates in the presence of witnesses (or callers). Once the magistrate determined – after listening to Socrates and Meletus (and perhaps the other two accusers, Anytus and Lycon) – that the lawsuit was permissible under Athenian law, a date was set for the "preliminary hearing" (anakrisis) and terms for the hearing were posted as a public notice at the Royal Stoa.
\subsection{The Preliminary Hearing (Anakrisis)}

The preliminary hearing before the magistrate at the Royal Stoa began with the reading of the written charge by Socrates' accuser, Meletus. Socrates then formally answered the charge. Both the written charge and denial were then attested to by each, under oath, as being true. The next phase of the preliminary hearing was an interrogation. First, the magistrate questioned both Meletus and Socrates. This is to see whether the issue could be settled out of court and to see whether there was any merit to the charges. Second, both the accuser and defendant were allowed to question each other. This was another chance to change the course of events. The third and final phase, supposing that the magistrate found the case worthy of consideration, the magistrate would draw up formal charges against the accused and set a date for the public trial. For Socrates, these charges (relating to impiety and corruption of youth), the actual paperwork, were preserved as a public document (antomosia), and they survived until at least the second century C.E., but were subsequently lost. 
\subsection{The Trial}

The trial of Socrates took place over a nine‐to‐ten hour period in the People's Court, located in the agora, the civic center of Athens. The jury consisted of 500 male citizens over the age of thirty, chosen by lot from among volunteers. People today might think that this is a ridiculously high number for a jury.  But, sometimes, the juries could be as large as 1501 men. This was a protection against bribery. For example, in order to ensure that a case went a certain way, the person would need to bribe at least 251 people, and it's not likely that many could afford to do this.  All jurors were required to swear by the gods of Zeus, Apollo, and Demeter the Heliastic Oath: "I will cast my vote in consonance with the laws and decrees passed by the Assembly and by the Council, but, if there is no law, in consonance with my sense of what is most just, without favor or enmity. I will vote only on the matters raised in the charge, and I will listen impartially to the accusers and defenders alike."

Most of the jurors were probably farmers, as that was the principal occupation of the day. For their jury service they received payment of three obols. An obol was a currency of the day and it was around a 40TH of an ounce of silver. In general, three obols fed you an paid for a leisurely night. The jurors sat on wooden benches separated from spectators by some sort of barrier or railing. Given Socrates's fame and the notoriousness of the charge against him, the crowd of spectators was most likely large – including, of course, the most famous pupil of Socrates, Plato.

The trial began in the morning with the reading of the formal charges against Socrates by a herald. Few, if any, formal rules of evidence existed. The prosecution presented its case first. Meletus, Anytus, and Lycon had three hours, measured by a waterclock, to make their argument for a finding of guilt. Each accuser spoke from an elevated stage. No record of the prosecution's argument against Socrates survives. Following the prosecution's case, Socrates had three hours to answer the charges. Although many written versions of the defense – or apology – of Socrates at one time circulated, only two have survived: one by Plato and another by Xenophon. After the arguments, the herald of the court called on the jurors to consider their decision. In Athens, jurors did not retire to a juryroom to deliberate – they made their decisions without discussion among themselves, based in large part on their own interpretations of the law. The 500 jurors voted on his guilt or innocence by dropping bronze ballot disks into marked urns. Only a majority vote was necessary for conviction. Four jurors were assigned the task of counting votes. In the case of Socrates, the jury found him guilty on a relatively close vote of 280 to 220. (Interestingly, if less than 100 jurors voted for guilt, the accusers had to pay a fine to cover trial costs, which is similar to something which is found in court cases today.)
\subsection{The Final Phase}

If a defendant is convicted, the trial enters a second phase to set punishment. The prosecution and the defendant each propose a punishment and the jury chooses between the two punishment options presented to it. The range of possible punishments included death, imprisonment, loss of civil rights (i.e., the right to vote, the right to serve as a juror, the right to speak in the Assembly), exile, and fines. In the trial of Socrates, the principal accusers proposed the punishment of death. Socrates, if Plato's account is to be believed, proposed first the punishment – or, rather, the nonpunishment – of free meals in the center of the city, then later the extremely modest fine of one mina of silver. Apparently finding Socrates' proposed punishment insultingly light, the jury voted for the prosecution's proposal of death by a larger margin than for conviction, 360 to 140. The execution of Socrates was accomplished through the drinking of a cup of poison hemlock.

\section{Part 2.2: Socrates' Defense/Arguments}

\section{The Charges}

There were three formal charges which Socrates was going to need to reply to in his 3 hour block of time. The first, the real zinger, was corrupting the youth of Athens. This was essentially causing the youth to have rebellious mindsets and be against the state. There was some evidence to this claim as several attempted rebellions and revolutions had taken place with many of the young people which Socrates spoke with often playing central roles, though Socrates himself was more of a bystander than anything in these cases.  The second charge, inventing new gods, was a lesser case. This is more in line with the initial charge of impiety. According to the laws of Athens, blasphemy of this sort was a crime punishable by death.  The third charge, which could have either built up or dismantled the second, was not believing in the gods of Athens. This could have gone two ways, either the accusers could have said that in inventing new gods, Socrates abandoned belief in the Athenian pantheon (which would have built up the case regarding blasphemy) or they could have said that Socrates was an atheist, which directly contradicts the second charge. As it turns out, the accusers, not trained in philosophy, did not see the contradiction and went with the latter option (being an atheist).
\section{Socrates' Method}

Philosophy, in this period, was still very much in its infancy, it wasn't as formalized as it is today (throughout this class, you will be learning the philosophic method, which is the great and powerful grandmother to the scientific method which you may be familiar with). For the pre-socratics (which includes Socrates), you can think of philosophy as a game where they are still trying to figure out the rules. If you have ever seen little kids try and play soccer or football for the first time, then you will have the right kind of image in mind.  Socrates, for his part, in the trial, was unfamiliar and inexperienced in how to present publicly, especially given the large number of people in attendance, and was not really the best at being persuasive (rhetoric was an area which he really did not have much respect for). So, Socrates begins his rebuttal by explaining that he will treat this like he would a debate in the Agora. In those cases, Socrates would merely ask questions to people. Rarely, if ever, does Socrates actually state his stance. Rather, Socrates merely restates the stance which the person had stated in reply to his question and then adds further questions to this. If you ever get the chance to see me in a lecture format, I do this methodology for the first few days of lecture, off and on. As a result, Socrates did not address the jury directly, rather just questioned Meletus. Though this method is quite good for teaching and one-on-one debate, it's not the best for a presentation of this sort. 
\section{Socrates' Questions}

Since I understand that the translations for this are a bit wordy and the word choice can be a bit unnatural (I have a lot of experience translating Latin and if a person is not careful or doesn't think about the context of the wording, the translation will come out as quite formal), I have taken the liberty of rewording some of the content of The Apology to make the meaning come through a little clearer. It's still important that you read the original, as it's a wonderful work, however, this is here to help. For this, you can read it as a script, with 'S' being Socrates and 'M' being Meletus.
\subsection{Corrupting the Youth}

This charge, as I have mentioned, is the real big charge, the others, though still punishable by death, don't have as much going for them, as we will see. To prove his point, Socrates asked Meletus, around, 11 questions, which were leading him down a very particular rabbit-hole.
\factoidbox{
    S: Do you think that it is the greatest importance to make the youth as great as possible?

    M: I do.

    S: Who improves them? 

    M:The Laws

    S: You didn’t answer me, Who has knowledge of these laws?

    M: These Jurymen

    S: What do you mean? Do all these people improve the youth?

    M: All of them.}

So, at this point, Meletus starts off by essentially claiming that the laws of Athens are the sort of things which can improve the youth. But this is not correct, laws are passive actors, they aren't the sort of thing which can take an active part in the raising of the youth. That requires a person who knows the laws and has the ability to accurately teach. So, to his credit, Meletus noticed this and backtracked, claiming that the jurymen have the knowledge of the laws and the ability to teach them (think about how it would have sounded if he claimed that the jurors didn't know the laws).
\factoidbox{
    S: By Hera! That’s a lot of people to improve our youth, what great shape we must be in! What about the audience? So they improve them? 

    M: They do.

    S: What about the council?

    M: They improve them too.

    S: Everyone in Athens makes the youth better but me, is that what you are saying?

    M: Yep, everyone makes the youth better but you.}

At this point, Meletus doesn't seem to see where this is going. Socrates has, through his questions, basically forced Meletus to claim that all the people of Athens improve the youth, with Socrates being the sole bad influence. From a rhetorical side, this was an interesting ploy. If Meletus had claimed otherwise, then Socrates could have easily countered by asking why those people weren't on trial with him, why charges hadn't be posed against them. But, since Meletus said that Socrates was the only corrupter, Socrates needs to go a different route, which leads us to the final 'build-up' question:
\factoidbox{
    S: Man, you put me in a rough spot. But does this also apply to horses? That all people improve them and only one person corrupts them? Obviously, it is the opposite. The horse breeder (trainer) improves them and all others corrupt them. It would be a wonderful world if our youth only had one corrupting influence in their lives!}

The horse analogy has not aged well, as for contemporary English, calling a person a horse is a bit insulting. However, think about it this way, a good horse was like the nice car of its day. The vast majority of people aren't trained in how to repair and maintain a car, or in this case, care for and train a horse. Rather, there are very few people who really know how to maintain those things, others who use them, rather unintentionally, corrupt them (wear them out, make them forget their training, etc.). In the case of the youth, the vast majority of people aren't trained in the best child-raising techniques, they interact with children in unintentional ways which lead to negative behaviors, and so forth. It would be a great world if the children only had one corrupting influence, as that would get quickly drowned out by all of the other influences around them.

This leads us to the final set of questions concerning this particular accusation.
\factoidbox{
	S: You have made it clear that you care very little for the youth and haven’t put good thought into the reason you’re putting me on trial. Now answer me this: Is it better for a person to live among good fellows or wicked? This isn’t a hard question. Do wicked people harm their neighbors and good people benefit them?

    M: Certainly

    S: Now, is there a person alive who would rather be harmed than benefited?

    M: Certainly not

    S: Also, do you accuse me of corrupting the youth deliberately or unintentionally?

    M: Deliberately

    S: What then, Meletus? Are you so wise that you realize that the wicked do wicked and the good good but I am so stupid that I have not realized this? Have I failed to recognize that if I make anyone of my associates bad, they will harm me? But still you say that I do this great evil voluntarily? I can’t believe you, Meletus, nor do I think anyone else in the world does! But, leaving that aside, either I do not corrupt them, or if I corrupt them, I do it involuntarily. This shows that you are lying either way we go. But if I corrupt them involuntarily, the law is not to haul people into court, but to take them and instruct and admonish them in private. For it is clear that if I am told about it, I shall stop doing that which I do involuntarily. But you avoided associating with me and instructing me, and were unwilling to do so, but you haul me in here, where it is the law to haul in those who need punishment, not instruction.}

This is where we get Socrates' real argument against the charge. It follows a very interesting logical form, called Constructive Dilemma. It is one which Socrates, historically, had just used earlier in the day outside of the court concerning the nature of goodness. In an abstract form, the reasoning goes like this: One of two (or more) options is correct (A or B). If A is correct, then  C and if B is correct, then D. Therefore, either C or D is correct. Here is the argument for the case along with another example used by Socrates in another dialogue:
\begin{tabular}{p{2in}|p{2in}}
Corrupting the Youth& Piety\\

    Either I don't corrupt the youth or I corrupt them unintentionally.&Either something is moral because it's commanded by the gods or it's commanded by the gods because it's moral.\\
    If I don't corrupt the youth, then the charge against me is false (I'm innocent here).&If it's moral because it's commanded by the gods, then the morality of actions is arbitrary (random, not fixed).\\
    If I corrupt the youth unintentionally, then you should have instructed me in private and not taken me to court (I'm innocent here).&If it's commanded by the gods because it's moral, then the morality of actions is not determined by the gods (we don't need to reference them to figure out the morality of actions).\\
    So, either way, I am innocent of this charge.&  So, either morality is arbitrary or it's not determined by the gods.\\
\end{tabular}

The argument regarding piety can also work if you replace 'the gods' with 'God'. If you are interested in that particular argument, look into the Euthyphro Dilemma.