\part{What is Knowledge? How Do We Know Things?}
\label{ch.modsix}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}
\input{meditationsonetothree}
\chapter{Part 12: Ren\'e Descartes, Life, Times, and Meditations}
Descartes lived during an intellectually vibrant time. The Scholastics had supplemented Catholic doctrine with a tradition of Aristotle scholarship and early scientists like Galileo and Copernicus had challenged the orthodox views of the Scholastics. Surrounded by conflicting yet seemingly authoritative views on many issues, Descartes wants to find a firm foundation on which certain knowledge can be built and doubts can be put to rest. So he proposes to question any belief he has that could possibly turn out to be false and then to methodically reason from the remaining certain foundation of beliefs with the hope of reconstructing a secure structure of knowledge where the truth of each belief is ultimately guaranteed by careful inferences from his foundation of certain beliefs.

When faith and dogma dominate the intellectual scene, “How do we know?” is something of a forbidden question. Descartes dared to ask this question while the influence of Catholic faith was still quite strong. He was apparently a sincere Catholic believer and he thought his reason based philosophy supported the main tenants of Catholicism. Still he roused the suspicion of religious leaders by granting reason authority in the justification of our beliefs.

Descartes is considered by many to be the founder of modern philosophy. He was also an important mathematician and he made significant contributions to the science of optics. You might have heard of Cartesian coordinates. Thank Descartes. Very few contemporary philosophers hold the philosophical views Descartes held. His significance lays in the way he broke with prior tradition and the questions he raised in doing so. Descartes frames some of the big issues philosophers continue to work on today. Notable among these are the foundations of knowledge, the nature of mind and the question of free will. We’ll look briefly at these three areas of influence before taking up a closer examination of Descartes’ philosophy through his Meditations of First Philosophy.\autocite{Descartes1}

To ask “How do we know?” is to ask for reasons that justify our belief in the things we think we know. This is the branch of philosophy called \Gls{epistemology}, which is a fancy word for the study of knowledge (a little meta, I know). The epistemological project of providing systematic justification for the things we take ourselves to know was launched by Descartes and it remains a central endeavor in epistemology to this day. This project carries with it the significant risk of finding that we lack justification for things we think we know. This is the problem of skepticism. \Gls{skepticism} is the view that we can’t know. Skepticism comes in many forms depending on just what we doubt we can know. While Descartes hoped to provide solid justification for many of his beliefs, his project of providing a rational reconstruction of knowledge fails at a key point early on. The unintended result of his epistemological project is known as the problem of Cartesian skepticism. We will explain this problem a bit later in this chapter.


\newglossaryentry{epistemology}
{
name=epistemology,
description={A field of study within Philosophy dedicated to questions concerning knowledge, belief, and justification}
}

\newglossaryentry{skepticism}
{
name=skepticism,
description={The epistemological stance that knowledge is impossible either globally (it is impossible to know anything at all) or locally (knowledge is impossible within certain topics)}
}



Another area where Descartes has been influential is in the philosophy of mind (Module 3). Descartes defends a metaphysical view known as dualism that remains popular among many religious believers. According to this view, the world is made up of two fundamentally different kinds of substance, matter and spirit (or mind). Material stuff occupies space and time and is subject to strictly deterministic laws of nature. But spiritual things, minds, are immaterial, exist eternally and have free will. If dualism reminds you of Plato’s theory of the Forms, this would not be accidental. Descartes thinks his rationalist philosophy validates Catholic doctrine and this in turn was highly influenced by Plato through St. Augustine.

The intractable problem for Descartes’ dualism is that if mind and matter are so different in nature, then it is hard to see how they could interact at all. And yet when I look out the window, an image of trees and sky affects my mind. When I will to go for a walk, my material body does so under the influence of my mind. This problem of mind body interaction was famously and forcefully raised by one of the all too rare female philosophers of the time, princess Elizabeth of Bohemia.

Previously, we explored the philosophy of mind which was  launched in the wake of problems for substance dualism and because of the advancements there, Descartes' fruitful failure lead to neuroscience, cognitive psychology and information science. We also see how undeserved philosophy’s reputation for failing to answer its questions is. While many distinctively philosophical issues concerning the mind remain, the credit for progress will go largely to the newly minted science of mind. The history of philosophy nicely illustrates how parenthood can be such worthwhile but thankless work. As soon as you produce something of real value, it takes credit for 

The final big issue that Descartes brought enduring attention to is the problem of free will (Module 4). We all have the subjective sense that when we choose something we have acted freely or autonomously. We think that we made a choice and we could have made a different choice. The matter was entirely up to us and independent of outside considerations. Advertisers count on us taking complete credit and responsibility for our choices even as they very effectively go about influencing our choices. Is this freedom we have a subjective sense of genuine or illusory? How could we live in a world of causes and effects and yet will and act independent of these? And what are the ramifications for personal responsibility? This is difficult nettle of problems that continues to interest contemporary philosophers.

Descartes’ is also a scientific revolution figure. He flourished after Galileo and Copernicus and just a generation before Newton. The idea of the physical world operating like a clockwork mechanism according to strict physical laws is coming into vogue. Determinism is the view that all physical events are fully determined by prior causal factors in accordance with strict mechanistic natural laws. Part of Descartes’ motivation for taking mind and matter to be fundamentally different substances is to grant the pervasive presence of causation in the material realm while preserving a place for free will in the realm of mind or spirit. This compromise ultimately doesn’t work out so well. If every event in the material realm is causally determined by prior events and the laws of nature, this would include the motions of our physical bodies. But if these are causally determined, then there doesn’t appear to be any entering wedge for our mental free will to have any influence over out bodily movements.

\chapter{Part 13: An Overview of Descartes' Meditations on First Philosophy}

\Gls{epistemology} (from the Greek word $\eta\pi\iota\sigma\tau\eta\mu\eta$ (epist\-em\-e) meaning ‘knowledge’) is a branch of philosophy which deals with knowledge and belief. Some of the questions which are found there are:
\begin{enumerate}
    \item What does it take to know something?
    \item What is belief?
    \item What if faith? (This is a more recent question)
    \item What are the different kinds of knowledge? Are some better than others?
    \item What is wisdom?
    \item What separates knowledge from opinion?
    \item And there are many, many others.
\end{enumerate}
Despite what others may tell you, and I have debated this on many occasions, Descartes' Meditations concern epistemology, not metaphysics. So, Descartes' project in his meditations is to carry out a rational reconstruction of knowledge. Descartes is living during an intellectually vibrant time and he is troubled by the lack of certainty. With the Protestant Reformation challenging the doctrines of the Catholic Church and scientific thinkers like Galileo and Copernicus applying the empirical methods Aristotle recommends to the end of challenging the scientific views handed down from Aristotle, the credibility of authority has being challenged on multiple fronts. So Descartes sets out to determine what can be known with certainty without relying on any authority and then to see what knowledge can be securely justified based on that foundation.

The meditations are broken up into 6 parts and each serves a different purpose. I encourage you to read all 6, but the first 2 are enough for this module.
\section{The First Meditation}
Below is the first small section of this meditation, written in more plan language. The point of these meditations is to break down everything, doubt everything, until we can find something which can’t be doubted and use that as the basis for building up a 100\% guaranteed true set of beliefs/knowledge. The first meditation concerns what can be doubted.

\factoidbox{Several years ago, I came to the opinion that many of the things which I believed were true, were, in fact, false; and when that happened, I thought that I should do the job of trying to rid myself of these false opinions in order to get an indubitable structure for knowledge (science). But I waited until I was older and wiser so that I could have very little doubt in my abilities. On this account, I have delayed so long that it would be wrong to keep thinking on it and I should just put pen to paper. Today, then, since I don’t have anything which I need to do because I am retired, I will at length apply myself earnestly and freely to the general doubting of all my former opinions.}

For the second section of this meditation, Descartes starts us off by explaining his method for getting at that which cannot be doubted. Rather that dealing with each belief individually, he chooses to deal with classes, or categories, of beliefs (such as ones which all came from the senses, or were told to him by some person). If he finds reason to doubt some of the beliefs in a class, then he can doubt the entire class. The type of reason to doubt is also important. The reason must be one which is at the foundation of those beliefs. If you take out the foundation, the rest collapses. Here is an example of this general idea in practice:

\factoidbox{Take the beliefs that you got because someone told them to you. But, as we all know, people do sometimes lie or are mistaken about what they are telling others. This means that there is reason to doubt all of the things which you were taught in school, or by your folks, or by other people. Since there is reason to doubt those things, they can’t be the foundation of indubitable knowledge.}

Given the times Descartes was living in, with the rie of empirical science (the idea that knowledge is primarily gained from the senses, through experience and experiments), Descartes wanted to show that the fundamental basis for this knowledge was actually grounded in knowledge you can gain without appealing to your senses, called \emph{a priori} knowledge. So, he turns to whether one can doubt their senses. To get this off the ground, Descartes needs a way of getting rid of knowledge through your senses, as in, knowledge via experience. This means he needs a systematic way of making a worst case scenario (think Murphy's Law, if it can go wrong it will), making it so that the foundation of his knowledge is internal, not external.  According to Descartes, you can doubt your senses, meaning that they can't be at the foundation of an indubitable knowledge structure. He gives us two ways, with the second stronger than the first.
\begin{enumerate}
    \item What if he is having a really realistic dream? (this is in sections 6-8)
    \item What if he is under the influence of some evil demon? (this is in sections 9-12)
\end{enumerate}
The third section of this meditation gives us a basis for making the claim that our senses cannot be the foundational basis for knowledge. Put in more plain language, he says:

\factoidbox{Everything that I have held with certainty, I have gotten through my senses. But, I know that my senses have sometime misled me; as a general rule, you should not trust something (100\%) if it has lied to you.}

Every generation or so has a movie or show which draws inspiration from Descartes' Meditations or Plato's Cave (they both use similar thought experiments, but draw different, though related, conclusions. One great example is the first Matrix movie. So much inspiration was drawn that we have this little quote from Morpheus to Neo:

\factoidbox{Morpheus: Have you ever had a dream, Neo, that you were so sure was real? What if you were unable to wake from that dream? How would you know the difference between the dream world and the real world?}

The point here is that we could be deceiving ourselves, what if we are having a really, super, realistic dream? It would seem that, then, we couldn't trust our senses, rather what we experience would be an illusion.

\subsection{The dream case (this is section 6)}
At this point, Descartes gives the first thought experiment, or possible scenario, where his senses could be misleading, allowing him to discard them as the foundation for knowledge. Put more plainly, this section goes like this:

\factoidbox{Hey, you know, I am a man and, you know, I need to sleep. When I sleep, I tend to dream and be in those dreams. Sometimes those dreams are really crazy, but other times they make a lot of sense. Sometimes, I cannot tell the difference between my dreams and reality. How can I be sure that I am not, right now, dreaming something really realistic?}

This case, however, is not strong enough for Descartes' purposes. If he were in a very realistic dream, then Descartes could stil gain some knowledge from his senses, but this knowledge would be fairly abstract. Here is the plain language version of his reasoning:
\factoidbox{So, let's suppose that I am having this really realistic dream. This would mean that everything I am seeing now are nothing more than illusions from the unconscious mind. But we have to admit that the things in my dreams are based on things in the real world, no matter how mangled and jumbled they may be. Because of this, we can say that although the objects in dreams may be doubted, we must claim that there are things which they are based on. So, you can still not doubt your experienced whole-hog, like I'm trying to do.}

As you can see, if he were in a very realistic dream, then he would still have some knowledge about the external world. This is because dreams are based on experiences in the waking world. One can still have knowledge of abstract things like shape, color, motion, and the like. In order to completely discount knowledge gained from the senses, he needs a stronger possible case.
\subsection{God and Deception}

He believes that there is an all good God, but how could this fit with being mistaken in this way? If allowing a person to be mistaken did not jive with God, He would not allow it. But He does, we are mistaken all the time, so allowing a person to be mistaken must be just fine by God.

\subsection{The Evil Demon Case (this is section 12)}
Here we have the Evil Demon, which is the thought experiment strong enough to discredit all knowledge gained from the senses. This thought experiment is the basis for many books, movies, and shows.\footnote{need to list examples} In plain language, his introduction of this case goes like this:
 
\factoidbox{I will suppose that some Evil Demon has employed all of their power to deceive me. All of my senses and the things which I learned from them are nothing more than either illusions or are the product of the Demon’s deceptions. I will consider myself as without  a body or anything external to the mind, as they are, too, illusions by the Demon. But this this is a really hard job, and difficult to maintain; so I tend to fall back into thinking that the external world is indubitable when I leave my armchair. This tendency does not mean that they are, in fact, indubitable, but rather, that we have a built in bias.}
\section{Descartes' Second Meditation}


\factoidbox{The last meditation left me with no knowledge at all, everything could be doubted, and I have a hard time forgetting this. Also, I don’t see a way out of the situation, it is like I have been plunged in deep water and can’t find my way up or down. Despite this, I will forge on, treating everything with the slightest doubt of its truth as if it were false. I will continue to do this until I have found something which I know for certain or, at the very least, that I know for certain that nothing else is know for certain. For Archimedes to move the world from one place to another only demanded a  firm and immovable point. In the same way, if I can find one thing which is indubitable, then I can have great hopes for this project.}


\factoidbox{Right now, I suppose that everything I see are illusions or are not actual. I can’t trust my memory, so I must assume that the things I remember never existed. My senses are also dubitable, so I must assume that I don’t have them. Because I learned about things like body, figure, extension and motion from my senses, I must assume that they are not actual as well. Since I can’t have any of those things, what can I have which is true? Maybe only that nothing is certain.}



\factoidbox{How do I know that there isn’t something different than those things which I can’t doubt? Didn’t I suppose that there was some kind of Demon who could cause these thoughts to come to my mind? But why do I even need to bring up this Demon, when I would deceive myself. Even if there is some demon, I am still something, I still exist. I have supposed that I don’t have a body, but then I must ask what comes from that? Am I so dependent on the body and the senses that without them, I don’t exist? But I had been persuaded that there was nothing in the world. When  I was persuaded, was I also persuaded that I don’t exist? Not at all, I certainly exist, since I was persuaded. At the same time, I supposed that there was some great demon deceiving me. This further shows that I exist, since I am being deceived. No matter how much this demon deceives me, he can’t make it that I don’t exist, so long as I am conscious, I am something. There is at least one thing which I cannot doubt, I am, I exist.}


\factoidbox{Even though I know that I exist, I don’t know what I am. The next step is one I need to take care in making, because if I say that I am the wrong sort of thing, then I will have strayed from the truth. This means that I need to think about what I thought I was and cast those aside. I will apply the same methods as before in order to get at what is undoubtable.}


\factoidbox{But what can I say I am if I suppose that there is this demon? Can I say that I still have all of the things which belong to the body? After looking into it, I can say that done really belong to me. Let’s move on, then, to the attributes of the soul. The first mentioned were the powers of nutrition and walking; but, if I don’t  have a body, then I can’t be able to either walk or be nourished.

Second, we had perception. In the same way as before, perception requires that I have a body, besides, I have often thought my self having perceptions in dreams.

The third and last attribute is thinking; and this is what really belongs to me. This alone is inseparable from me. I am--I exist: this is certain; but, when do I exist? I exist whenever I am thinking. If I stop thinking, it would seem that I cease to exist. I am therefore only a thinking thing, that is, a mind. I am, however, a real thing, and really existing; but what kind of thing? The answer was a thinking thing.}

\subsection{Commentary:}

Here, Descartes has his Evil Demon case, think of it like the Matrix without the mad glitches. This puts him in a state where he can doubt almost everything. Descartes ends up finding one thing which he can't doubt.

Even an evil deceiver could not deceive Descartes about his belief that he thinks. At least this belief is completely immune from doubt, because Descartes would have to be thinking in order for the evil deceiver to deceive him. In fact there is a larger class of beliefs about the content of one’s own mind that can be defended as indubitable even in the face of the evil deceiver hypothesis. When I look at the grey wall behind my desk I form a belief about the external world; that I am facing a grey wall. I might be wrong about this. I might be dreaming or deceived by an evil deceiver. But I also form another belief about the content of my experience. I form the belief that I am having a visual experience of greyness. This belief about the content of my sense experience may yet be indubitable. For how could the evil deceiver trick me into thinking that I am having such an experience without in fact giving me that experience? So perhaps we can identify a broader class of beliefs that are genuinely indubitable. These are our beliefs about the contents of our own mind. We couldn’t be wrong about these because we have immediate access to them and not even an evil deceiver could misdirect us.

The problem Descartes faces at this point is how to justify his beliefs about the external world based on the very narrow foundation of his indubitable beliefs about the contents of his own mind. And this brings us to one of the more famous arguments in philosophy: Descartes’ “Cogito Ergo Sum” or “I think, therefore I exist.” Descartes argues that if he knows with certainty that he thinks, then he can know with certainty that he exists as a thinking being. Many philosophers since then have worried about the validity of this inference. Perhaps all we are entitled to infer is that there is thinking going on and we move beyond our indubitable foundation when we attribute that thinking to an existing subject (the “I” in “I exist”). There are issues to explore here too. 

\section{Descartes' Third Meditation}
At this point, Descartes only has two justified beliefs. One is about his thoughts (the content of his mind) and the other is about his own existence (I think therefore I am). At the moment, he doesn't have any knowledge, or at least justified beliefs about anything outside of his mind. For example, he can't say "I know I have hands" or "I know there's a piece of paper in front of me."

Also, Descartes doesn't have any information about truths we know through reason alone, like 2+2=4 or other things like that. Those just don't have the justification yet. To get some basis to justify that stuff, Descartes sets out to prove that God exists and is not like his evil demon. Once the evil deceiver hypothesis is taken out because of God, facts we get from reason and perhaps those from our senses could get the justification to be knowable. However, not just any argument for the existence of God and that God's all-good will do the trick (we will be looking into a lot of different arguments for the existence of God in Module 5). The trick for Descartes’ project of a rational reconstruction of knowledge is to prove the existence of a good God by reasoning using only those beliefs that he has identified as indubitable and foundational.

Descartes argument for the existence of a good God goes roughly as follows:
\begin{enumerate}
    \item I find in my mind the idea of a perfect being.
    \item The cause of my idea of a perfect being must have at least as much perfection and reality as I find in the idea.
    \item I am not that perfect.
    \item Nothing other than a good and perfect God could be the cause of my idea of a perfect being.
    \item So, a good and perfect God must exist.
\end{enumerate}
This argument is a really simple version of the argument which Descartes gives (it's deep and complex) but the core features are there and can show the flaw in his reasoning (all arguments for the existence of God have at least one flaw (most of them have the same flaw, as we will see). Let's grant that this argument is valid and move over to look at it's soundness. Keep in mind that for Descartes to do the job he is trying to, all of his premises in this argument need to be indubitable and foundational. The first premise is a belief about his own thoughts, so we will give it to him. Though it is not as clear, premise three might arguably count as a foundational belief about the contents of Descartes' own mind. An evil deceiver, being evil, would lack perfection found in Descartes idea of a perfect being. So as powerful as such a being could be, the cause of Descartes idea of a perfect being must be more perfect than any evil deceiver. Perhaps any being so perfect would have to be a good God.

But the issue for Descartes is in the second premise. What reason do we have for thinking that the cause of something must have at least as much perfection as its effect?

The idea that there are degrees of perfection and the notion that the less perfect can only be explained in terms of the more perfect traces it's way back to Plato and his theory of Forms. Like many ideas of this type, it will strike many of us as implausible or even incomprehensible. What, exactly, is perfection supposed to mean here?  And even once we’ve spelled this out, why think causes must be more perfect?

It doesn't seem all that rare for less prefect things to make more perfect things (if you are a parent, it might seem clear that your kid is more perfect (better) than you are, or the ugly depressed artist making a beautiful painting).

In any case, whether the second premise can be explained and defended at all, the fatal flaw for Descartes’ project is that it is not foundational. It is not an indubitable belief about the contents of Descartes’ own mind, but rather a substantive belief about how things are beyond the bounds of Descartes’ own mind. So Descartes’ attempt to provide a rational justification for a substantive body of knowledge leaves us with an enduring skeptical problem. All we have immediate intellectual access to is the contents of our own minds. How can we ever have knowledge of anything beyond the contents of our own mind based on this? This is the problem of Cartesian skepticism.

Having diagnosed the fatal flaw in Descartes’ project, we should briefly consider how his rational reconstruction of knowledge was to go from there. Given knowledge of God’s existence and good nature, we would appeal to this to assure the reliability of knowledge had through reason and later also through the senses. God being the most perfect and good being would rule out the possibility of interference by an evil deceiver. We might still make mistakes in reasoning or be misinformed by the senses. But this would be due to our failure to use these faculties correctly. A good God, however, would not equip us with faculties that could not be trusted to justify our beliefs if used properly. This is a very cursory summary of the later stages of Descartes’ attempted rational reconstruction of knowledge in his Meditations. But it will suffice for our purposes.

\chapter{Part 14: Knowledge and Justification}

In the previous section of this module, we were discussing Ren\'e Descartes and his quest for certainty and knowledge. He did this by employing a form of methedological skepticism, doubt everything which can be doubted. In philosophy, \gls{knowledge}, as we are going to use it, is justified true belief. This allows us to make a nice little equation:\footnote{This equation will appear later.}

\begin{center}

K=TJB

Knowledge is True Justified Belief
\end{center}

Different stances in Epistemology regarding knowledge can be reduced to different ways of adding variables to this equation or even removing them. Some might claim that one can know something without believing it, others might claim that knowledge just is belief, and so on. The different stances in Epistemology can also be characterized by what they take each of these terms to mean (with the exception of knowledge). 

Justification for knowledge is different than other forms for justification which you might encounter (and even then there is some debate about the nature of that justification). In ordinary life, you are entitled to your medical records, which means that you are justified in asking for them. This kind of justification is related to, but not the same as, the kind of justification which we have for knowledge. Epistemic justification is the kind found when it comes to knowledge. This is background reasoning which supports the belief. A belief is epistemically justified if we have sufficiently good reason to think that the belief is true. 
\section{Why justification is important}

Some might try to answer the question ``what is knowledge?'' with a relatively simple answer ``you know it if you are correct.'' Using the equation from before, this stance could be characterized as K=T\footnote{if it's true then you know it}, K=B\footnote{knowledge is belief, there is no difference.}, or  K=TB\footnote{if it's true and you believe it, then you know it.} All three of these have a problem with getting lucky, where you guess and it turn out you were right. As an example, take this case:

\factoidbox{Recently, my little brother and I were binging the show Dragon Ball Z. My little brother came to me with a theory about the time travel based plots in the show.  As an expert on theories of time travel, I thought this was great. He came up with the idea that there had to be four different timelines, given what he had seen thus far. Explaining that there had to be two different versions of the same character and so forth. As it turns out, he was correct, there did need to be two different versions of that character and there needed to be 4 timelines, but for completely different reasons than the ones he laid out. He later said that it didn't feel like he was correct, something was missing.}

I quickly recognized what was missing: adequate \gls{justification}. The structure of his reasoning, though it lead to the correct answer, wasn't accurate. As a result, merely being correct is not enough for us to claim that we knew it. At the same time, merely having reasons to think that the belief is true is not enough to say that you knew it. Here is a, relatively, classic example of justification going astray. There have been various thought experiments which follow this line of thought, but I will go with the one which is earliest, as far as I am aware, from Alexius Meinong: 

\factoidbox{Imagine that you live near a park and most days at noon, a violinist plays. You can hear the beautiful music from your dinning room and, it seems right to say that if you hear the music, then you know that the violinist is playing. However, one day, because you, unknowingly, drank some bad tea, at noon, you go temporarily deaf and have an auditory hallucination of a violin playing. While it is true that the violinist is out there and you hear the music, does it follow that you know the violinist is out there?}

For cases like these, most people would say that, though you were right, you don't know that the violinist is playing. The reason, as before, is that there was something wrong with the justification. 

It should be noted that 100\% certainty, no chance of being wrong, is not necessarily required for justification. For example, in a real world case, you can know that your sink isn't leaking without being in the room with the sink. There is a chance you are wrong, given your evidence. Justification can come in degrees and the more justified a belief is, then the more likely it is to be accurate/true. In the violinist example, experience and prior evidence certainly gives ample justification for knowledge in most cases, but you would have had the same evidence if the violinist had been sick and unable to play. This is why good critical thinkers question, doubt, or even reject beliefs if they lack adequate justification.

We see that the that the real area of debate, what really separates knowledge from a lucky belief, is the justification. As a result, much of the debate in Epistemology centers around the standards for justification. If you believe something and that thing happens to be true, what else is needed for you to know it? Examples like this can be found all over the place, where you were correct, but for totally unexpected reasons. This is why good critical thinkers look deeply at the structure of their justification, how their beliefs relate to each other, which also calls back to one of Socrates' core questions; what is your evidence?



\newglossaryentry{knowledge}
{
name=knowledge,
description={Justified True Belief; You have good reason to believe it, you do believe it, and your belief is, in fact, true}
}


\newglossaryentry{justification}
{
name=justification,
description={The reasons one has to believe something, why they think it is true}
}

Justification for knowledge is always how your beliefs relate to each other. You can think of this as a building or a puzzle (these ways of picturing the structure lead to two different theories of justification). After doing just a little introspection, you will find that your beliefs relate to each other in various ways, and some make others more likely to be accurate. For example, take this case:

    \factoidbox{My belief that the weather-forecast in Washington is unreliable is based on my belief that they have claimed many times that it would be sunny and on my belief that the weather on many of those occasions was, in fact, rainy. My belief about their claims is based on my belief that I remember their statements correctly. My belief about the weather on those occasions is based on my belief that my memory is correct.}

For this, we see that my beliefs form a pyramid, or tower, like structure, which leads us right into the first theory of justification, Foundationalism, which is what Descartes used in the Meditations.

\section{Foundationalism}
Much of the debate in Epistemology, as I mentioned, centers around justification. There are three dominate theories of justification which we will cover. Doing a little introspection should make it clear that some beliefs are built upon others. One belief is justified by another belief. For a particularly strange example, take this case: 

\factoidbox{I believe that the Earth is actually a dodecahedron. It is not spherical but it is also not flat. I believe this because I think that the middle of two extremes is the most  likely to be accurate. I also believe that there are two extreme stances in the debate over the shape of the Earth. One side claims that the Earth is spherical while the other side claims that the Earth is a flat disk. The two beliefs about the extreme stances with the general belief about the accuracy of a middle ground leads me to the conclusion that the Earth must be a dodecahedron.} 

You might notice that this reasoning is much like an argument which we have seen previously. I have given reasons and some general principles to connect them. Laid out this way, we can diagnose what went wrong with the justification. It is correct that there are two extremes in the debate over the shape of the Earth and their descriptions are, roughly, accurate. The issue lies in the principle which links them together. The middle of two extremes is not always the most accurate. This causes us to come to the wrong conclusion. The debate about justification is how the beliefs either need to relate to each other or the sort of cement used to link them together.

The most basic and common stance in Epistemology is \Gls{foundationalism}. This is the stance that our beliefs, in order to be justified, need to be ultimately justified on the basis of certain, core, foundational beliefs. Here is an example of a system of reasoning using Foundationalism: 

\factoidbox{I have the belief that the weather-forecast in Washington state is unreliable. I believe this because I also beleive that that the forecasters have claimed many times that it would be sunny and on my belief that the weather on many of those occasions was, in fact, rainy. My beliefs about their claims is based on my belief that I remember their statements correctly. My beliefs about the weather on those occasions is based on my belief that my memory and senses are correct.} 

Notice that the reasoning and the beliefs used to justify my stance that the weather forecasts in Washington state are unreliable is structured like a tower or maybe a pyramid. At the bottom of this structure are the foundational beliefs, which support all of the others. In this case, the foundational beliefs are that my memory and senses are accurate. 

Foundational beliefs are your rock bottom, these aren't supported or justified by anything else. If your beliefs are structured in this way and you want to get at knowledge, certainty, then you need to look very closely at the foundational beliefs. These beliefs need to be self-evident, they are not  supported by other beliefs, they need to be strong enough to carry all the weight, all by themselves.  If you have any experience with construction, or Minecraft, this analogy will work nicely for you. The foundationalist thinks of the intellectual structure of knowledge, how the beliefs relate to each other, as a tall building. Buildings can take many different shapes, from tall towers to pyramids, to strange Brutalist architecture, with a small foundation and an expanding roof. But, at the end of the day, the foundation, the concrete slab it's built on, needs to be tough enough to take the weight. 

Finding your foundational beliefs might require a little help, or a highly questioning mind. Little kids will often continually ask `why'. It could start with something fairly simple like `mommy, why is the sky blue?' and then you will answer with something accurate to the best of your knowledge, and then they will ask `why', you will reply, and they will ask `why' again, and again, and again. Eventually, you will get to a point where you just need to throw your hands up and say `that's just the way it is.' For example, last time I had this happen, I just stopped at `because some things exist and other things don't.' This final belief is your foundation. If you keep on digging, there's nothing left. Is this strong enough, on its own, to hold up the weight? In my weather forcast example, the foundations could be that my senses and memory are accurate (for those cases). Is that strong enough, all on its own? 

Foundational beliefs tend to be very basic logical or experiential truths. For example, something like `my senses are accurate' or `if P implies Q and P is true, then Q is true', or that there can't be contradictions. These all can be used to build up a structure of beliefs, which, though possible to be inaccurate, still count as justification. Additionally, Foundationalism holds that the beliefs you have, if they are justified, cannot directly or indirectly justify themselves. This is to say that there cannot be circular justification. For example, take this strange structure of justification:

\factoidbox{I believe that Mt. Rainer is an active volcano. I believe this because my middle school teacher told me so and I believe that she was accurate in her claim. I believe she was accurate in claiming this because I believe that Mt. Rainer is an active volcano.}

Notice that the first and final beliefs in this structure of reasoning are the same. Ultimately, it does not reach out to anything more than my belief that Mt. Rainser is an active volcano. If someone explained their stance in this sort of way, most people would be, correctly, very skeptical of their reasoning. Some might claim, however, that circular reasoning like this is not always bad. This case, in isolation, is but if we add more beliefs and increase the network of beliefs, showing more connections, then the circles will improve rather than deminish the justification. 

\newglossaryentry{foundationalism}
{
name=foundationalism,
description={The stance in Epistemology concerning justification which states that in order to know something the justification necessary must be structured in such a way as they fundamentally rely on basic, foundational beliefs which are self-evident and do not need further justification. The structure is often compared to a pyramid or a building}
}


\section{Infnite Regress Theory}

The \glspl{infinite regress theory} of justification points the finger at the foundational beliefs. For a belief to count as knowledge, according to the foundationalist, need to be built up from basic, foundations, which are self-evident. Those beliefs don't require any further justification. This theory claims that all beliefs need to be justified. It agrees with foundationalism that beliefs can't directly or indirectly support themselves, but it claims that foundational beliefs can't lead to knowledge. For a belief to count as knowledge, according to this theory, it must be justified by some other beliefs, which in turn need to be further justified by others, and those need to be justified by still others, and so on. For you to know something, there would need to be an infinite string of justifying beliefs, never making a circle, which support it. Such an account of justification makes knowledge, frankly, impossible.

\newglossaryentry{infinite regress theory}
{
name=infinite regress theory,
description={The stance in Epistemology concerning justificaton that all beliefs must be justified, there can be no `foundational' beliefs but also beliefs cannot indirectly justify themselves. The structure is like an infinitely long chain of branching beliefs},
plural=Infinite Regress Theory
}


Some of you might have heard the phrase 'it's turtles all the way down', which has a humorous story behind it.  This story has changed and mutated quite a bit with time, but it gets the general idea:

    \factoidbox{Long ago, a philosopher was giving a lecture to the public. In it, they mention the structure of the solar system and the planets. Afterwards, an older lady comes up to the philosopher and says ``your theory that the sun is at the center and the Earth is a ball is very convincing, but I have a better one." Amused, the philosopher asks about it and she replies ``we live on the crust of the Earth which rests on a turtle's back." Instead of demolishing her stance with evidence, the philosopher chose to follow Socrates and ask follow-up questions, ``if that is correct, ma'am, then what does the turtle stand on?" Equally amused, the lady replied ``the first turtle stands on the back of a second turtle." ``But," replied the philosopher, ``what does the second turtle stand on?" She then said ``why, the back of yet another, bigger turtle, it's no use, Mister, it's turtles all the way down."}

Infinite Regress Theory follows a very similar line of thought. Rather than their being foundational beliefs, in order to have knowledge, there must be beliefs all the way down.  However, having an infinite chain of knowledge doesn't, on its own, give us knowledge. For example, suppose that there's at least 0 soda cans on my desk. Here is a chain of beliefs; if there's at least 2 soda cans on my desk, then there's at least 1; if there are 3 soda cans on my desk then there are at least 2; if there are 4... Continuing for all numbers. None of these beliefs are really justified,  the chain will never reach back to reality and justify my claim that there's no soda cans on my desk. This also means that, if this theory is correct, knowledge is impossible and Global Skepticism is the way to go. 

\section{Coherentism}

Building off of the previous two, Foundationalism claims that there are bedrock beliefs which don't require further justification, knowledge is justified on the basis of other beliefs (aside from the foundational ones), and beliefs can't directly or indirectly justify themselves to count as knowledge. Infinite Regress Theory claims that there are no foundational beliefs, knowledge must be justified on the basis of further more basic beliefs, and beliefs can't directly or indirectly justify themselves.  \Gls{coherentism} is like Infinite Regress Theory in that it claims that there are no foundational beliefs in the structure of knowledge, but it rejects that beliefs can't indirectly justify themselves. Beliefs are justified in how they cohere or fit together. Several philosophers have called the structure involved in this justification a ``web of belief''\autocite{Quine1} and this is an apt picture to keep in mind when thinking about this structure of justification. In a spider web, every strand is connected, any where you start on the strand, you can follow the strands to any other one. But, it is not enough that the beliefs are all interconnected to be justfied. One could come up with radically strange reasons to move from ``green is a color'' to ``the Earth is flat'' but it doesn't follow that the person is justified in claiming that the Earth is flat.  

First, all of the beliefs in the web need to be consistent with each other. This should seem basic at this point, but in ordered to be properly justified, the beliefs you have can't be in conflict with each other. One cannot, for example, believe that Spiderman is a fictional character, that fictional characters do not exist, and that Spiderman is waiting for them in their apartment.\footnote{Unless they mean something different by `Spiderman' in these beliefs. In such a case, the beliefs may be consistent but they would lack a point.} 

Second, the beliefs must all be connected (as we saw before with the `web of belief') but weak connections between beliefs weaken, rather than strengthen, the degree of justification one has for any belief. At the same time, strong connections between the beliefs strengthen one's justification. One way to think about this is to consider likelihood. How likely is it that the inference you have made connecting two beliefs is accurate? The more likely the connection is, the stronger the connection. 

And third, one's justification is increased by the number of connections one makes or can reasonably make between their beliefs. This is to say that the more links one has connecting beliefs to others and the more integrated the beliefs are together, the more justified they are in their beliefs.\footnote{This is not to say that the more beliefs you have, the more justified you are in any of them. Rather, the more densely packed the connections are, the more they link together, the more justified you are.}   

It is worth noting that without an infinite number of beliefs, which no person can have, and without foundational beliefs, it's not possible to have a system of beliefs fit the coherentist model without there being some kind of circle, at least one. Coherentists embrace this circularity and claim that while small obvious loops are clearly a sign of bad reasoning, large, interconnected, multidirectional loops, are a sign of good justification. Because of this circularity, Coherentism allows for a sort of back and forth process; with your beliefs being shaped by your experiences and your experiences being shaped by your beliefs. Justification can come in two different directions. 

For example, consider a, probably young, person, Skie, who is unsure about the shape of the Earth. She goes out and looks at photos taken from orbit, listens to first hand accounts from astronauts, and even goes up in a space shuttle to see the shape for herself. In this case, the many individual observations, testamonies, and facts build up evidence for a more grand generalized claim that the Earth is round. The particular observations give us a basis for the principles. Here, the justification is going `bottom-up', moving from particulars to generalizations. 

At the same time, sometimes, we have cases where some evidence, an observation, just can't fit with a generalization we have made; sometimes, the evidence disproves our generalization. For example, suppose that after years of bird watching in states aside from Washington and seeing thousands of ravens, I generalize to the belief that all ravens are black. But, one day, while bird-watching in Washington, I spot two white ravens (yes, there is a beach in Washington with a family of white ravens). This will clearly conflict with my core belief that all ravens are black, but that's OK. After confirming that those are in fact ravens and doing my due diligence, I amend my belief to something like ``most ravens are black." But that is still fine with the Foundationalist, they would say that the process is still bottom-up.  

Coherentism allows for our beliefs to shape our experiences, going `top down'. In these cases, the generalized beliefs justify our particular evidence. The generalized claims stick around by categorizing or making sense of our observations.This can be because of the biases which we have covered in this class or it could be because of other reasons. Coherentism claims that our observations can be `theory laden', or influenced by our general theories. For example, a person, Mary, strongly believes in miracles. When she sees, for example, a statue of Jesus with water dripping from the cheeks, Mary will instantly color her experience of the `weeping statue' as a miracle. Similarly, Norm, who strongly holds that miracles don't happen, will see such an event and reject it almost instantly as his eyes playing tricks on him or will jump to some explanation in order to explain the event. This is the kind of back and forth which Coherentism entails. People who strongly believe in Sasquatch are likely to see Sasquatch because their beliefs color their experience and their experience shapes their beliefs. In Mary's case, her belief that the weeping statue is a miracle justifies her stance that miracles happen and her stance that miracles happen justifies her stance that it was a miracle. In Norm's case, his belief that miracles don't happen justifies his belief that the weeping statue was not a miracle. 


\newglossaryentry{coherentism}
{
name=coherentism,
description={The stance in Epistemology concerning justificaton that all beliefs must be justified, there can be no `foundational' beliefs. For this theory, it is necessary that some beliefs indirectly justify themselves. For this theory, beliefs are justified in how they cohere or fit together. The structure is like a spider web or loosely woven fabric}
}


\section{Native American Epistemologies}
By ``Native American Epistemologies'', I am referring to the systems of justification and knowledge used by the peoples indigenous to what is today the continental United States. The peoples, cultures, and epistemological systems native to the Americas (North and South) are very diverse, with each having subtle differences from their neighbors and those differences become greater the greater the distances between them. It pays, therefore, to be mindful that this is a cursory overview of the common features had by some of the peoples in a relatively small section of the Americas as a whole.\footnote{There are also the Epistemological systems used by the peoples native to Mesoamerica and various regions across South America. These systems may or may not have all or some of these common features.}  

All cultures, the world over, can have their requirements for knowledge be characterized, roughly, as `True Justified Belief'. In fact, we can make this into a nice little equation, like so:

\begin{center}
K = TJB
\end{center}

Thanks to this, we can see that the Epistemologies employed by different peoples and cultures can be reduced to disagreements about the answer to one or more of these questions:
\begin{enumerate}
\item What is `truth' as it relates to knowledge? Does this differ from `truth' as in a fact about the world?
\item What is a `belief'? Can I believe something without being able to use it?
\item What does it take to be justified in claiming something? What structure does the reasoning take?
\end{enumerate}
The `old world' Epistemologies, those found in Europe, North Africa, the Middle East, and Asia, all tend to agree, on an abstract level, about the answers to these three questions. Starting with the last, those systems all tend to be Foundationalist. There are basic, foundational beliefs (which don't need justification) and those build up to more complex beliefs in a tower like structure. These old world Epistemologies also all tend to hold that the nature of beliefs (as they relate to knowledge) is propositional, you know \emph{that} the sky is blue, \emph{that} carbon has 6 protons, \emph{that} a person needs to sleep, and so on. And, finally, they all agree that the proposition must correspond with reality, be a fact about the world, independent of you, in order for you to know it. This is to say that there is no difference betweem truth as a condition for knowledge and truth as in a fact about the world. 

Native American Epistemologies disagree with the `old world' ones on all three points.\autocite[18]{Burkhart1} Truth for knowledge is different than truth as in a fact about the world, there's something more to it. Beliefs are not just propositional, they are far more `global', all encompassing, they are not isolated. And, finally, the justification for knowledge is not Foundationalist, but rather Coherentist.   
\subsection{Truth}
As mentioned previously, the sense of truth used as a requirement for knowledge in the Epistemologies of Europe, Asia, the Middle East, and Africa\footnote{At the very least North Africa.} is the same as the sense of truth used in statements like ``it's true that there is no largest prime number'' or ``it's true that the Earth is spherical''.\autocite[57]{NortonSmith1} This makes truth a feature or property of propositions. It's true when it accurately corresponds or represents the way the world is, independent of the mind. This kind of knowledge or truth is sometimes called `propositional knowledge' or `propositional truth'. In English, we say things of the form `I know that...' the `...' there is the proposition, in isolation, which is claimed to be true because it corresponds with the what the world \emph{really} is. In these Epistemological systems, this `knowledge-that' is treated as fundamental, the basic form which knowledge takes and other forms of knowledge are explainable in terms of a collection of known propositions. 

Native American Epistemologies do not treat propositional knowledge, knowledge-that, as fundamental. Rather, the fundamental kind of knowledge is \emph{procedural} knowledge, or `knowledge-how'.\autocite[19]{Burkhart1} Procedural knowledge is measured by practicality in addressing some practical concern or reaching some goal. The kind of truth related to knowledge in Native American Epistemologies is procedural as well. Rather than being a property or feature of statements, it is a property of a performance or action.\autocite[63]{NortonSmith1} 

This points out the first of a few consequences of taking on this notion of truth: Truth comes in degrees. Some methods are hands-down more successful in reaching a goal than others, making them more true, in this sense. Compare, for example, two different farming practices. First, we have the Native American method of the three sisters. Corn (maize), squash, and peas are all planted together and grown together in the same spot (each corn stalk has peas climbing it and squash around the base). This method produces a large quantity of food and you can plant in the same mounds for generations without loss of soil integrity, plant vigor, or other concerns. Second, we have the methods used by many farmers today. This is large scale monoculture,\footnote{Only one kind of plant grown in the field} using synthetic fertilizers, pesticides, and the like. These methods are commonplace because they produce far more food than the three sisters or other similar methods can. Even without the synthetic fertilizers and pesticides, such methods produce more foods. So, if our goal was simply to produce as much food (for humans) as possible, the monoculture method would be more true than the three sisters method. But, the use of the monocultural methods harms soil quality and the various additives and pesticides harm the rivers, fish, and the rest of the food chain from there.\autocite[21-22]{Burkhart1} 

This points out that there must be an overarching goal, a greater goal, such that the lesser goals are judged by how well they achieve or progress towards meeting it. This goal is to ``walk the right path'', be a good person and make good actions. This is fundamentally a moral component:

\factoidbox{The idea is simply that the universe is moral. Facts, truth, meaning, even our own existence are normative. In this way, there is no difference between that is true and what is right. On this account, then, all investigation is moral investigation. The guiding question for the entire philosophical enterprise is, then: what is the right road for humans to walk?\autocite[17]{Burkhart1}}
At least from this perspective, the monocultural practices are not on the path which a human should walk. It follows, then, that the monocultural methods, though more successful at achieving the lesser goal of producing the most food, are less successful at achieving the greater goal of being morally right, because it causes harm to our non-human counterparts. This, however, does not imply that what is the case about the world is right or as it should be. Remember that `true' or `truth' in this sense is different from `truth' in the sense of being a fact of the world, independent of us. Everything we do, say, and even think has a moral component to it. From this perspective, the `truth' required for knowledge is that it is good, moral, on the road you ought to walk. 

Following Thomas M. Norton-Smith, a Shawnee philosopher of clan Turkey, we can characterize this notion of truth as a biconditional:
\factoidbox{For subject(s), S, her, his, or their purpose or goal g, and action or performance p, p is true for S for g if and only if p is \emph{respectfully successful} in achieving g.\autocite[64]{NortonSmith1}}
This \emph{respectful success} of an action just means that the action is the morally right one to do, especially with respect towards the consequences, not just for humans, but non-humans as well.\footnote{This hints at a Consequentialist notion of ethics, which we will explore in Module \ref{ch.modeight}, or at least a form of Dao Consequentialism. Dao Consequentialism or `Way Consequentialism' states that an action is right if and only if it is a part of a way or pattern of life which as a whole causes more good and less bad. Different forms of Consequentialism propose different goods and bads which ought to be promoted or mitigated respectfully.}  

\subsection{Belief}
As the Choctaw Native American philosopher Lee Hester puts it, western science and philosophy is based on \emph{belief}.\autocite[321]{Hester1} Similar to the nature of truth in those Epistemologies, beliefs are taken to be isolated propositions which one claims to be the case; true beliefs would be those which one claims to be the case and accurately correspond with reality, independent of the person in question. The beliefs, taken together, form a map or a depiction of the world (territory). This map is taken to be the true depiction of the territory, there can only be one. 

In Native American Epistemologies, both the map and the territory are real, but the map is not mistaken for the territory.\autocite[321]{Hester1} The map is taken to be one of many ways to see the world around us, different ways of seeing may be more true than others according to how well they guide us and help us walk the path we ought to.  
\factoidbox{An old chief of the Crow tribe from Montana was asked to describe the difference between his tribe and the whites who lived nearby. Pausing slightly and drawing his conclusions, he remarked that the white man has ideas [beliefs], the Indian has visions\autocite[15]{Deloria1}}
\emph{Visions}, as the Crow chief calls them, or \emph{recognitions}, as Vine Deloria Jr. later calls them,\autocite[362]{Deloria1} are far more robust than beliefs. They are \emph{integrating experiences} which shape and inform how one sees the world in all matters, ``culturally, spiritually, psychologically, politically, and in matters of subsistence and use of technology''.\autocite[321]{Hester1} Unlike beliefs, visions are not isolated, they are fundamentally interconnected with all aspects of the person. One could, digging a little deeper, think that these Native American Epistemologies have propositional beliefs just as much as the old world ones, maybe even more of them. This could be correct but they are not the basis for knowledge, rather the fundamental unit is the \emph{vision}, composed of a litany of propositional beliefs. 

\subsection{Justification}

Thus far, we have seen how Native American Epistemologies differ from the old world Epistemologies in two different ways. First, `truth' as a requirement for knowledge is not the same as `truth' as in an accurate description of the world. `Truth' in Native American Epistemologies is assessed more pragmatically, something is true for the person if it that thing is \emph{respectfully successful} at achieving the person's goal. This respectfulness relates to the overarching goal of walking the right path, of being a good person and acting in a moral manner. Second, the \emph{vision} is far more important than the belief. Recall that this is an integrating experience which links all of your beliefs into a cohesive network. This points out the fundamental system of justification in Native American Epistemologies, Coherentism. As Norton-Smith puts it: 
\factoidbox{All beings and their actions in the American Indian world are related and interconnected, so knowing about the world involves actively seeking out newly emerging connections between experiences.\autocite[58]{NortonSmith1}}
We ought to note that \emph{experiences} are not just the waking ones which are undiluted and un-muddied (so to speak); all experiences are to be treated equally and integrated. In the western epistemic tradition, at least, experiences like dreams, prophetic visions, and the like are seen as \emph{deceptive};\autocite[71]{NortonSmith1} they are not the sort of experiences which give an accurate account of the territory and ought not to be included in the map of the world. This sharp distinction between appearance and reality is absent from Native American Epistemologies, all experiences, even those which would be rejected by western sciences, are worthy of integration into the larger network. More over, in the western tradition at least, first-hand experiences are taken to be privileged. This is to say that the person who had the experience is in a better position to use that experience in their justifications than a person who merely heard about it second hand. This notion of privilege is also absent from Native American Epistemologies, as Burkhart says, ``What place do I have to tell you that your experiences are invalid because I do not share them?''\autocite[26]{Burkhart1} This allows for \emph{tribal knowledge} or communal knowledge, knowledge which comes from the merging of many minds and experiences, handed down by tribal elders to the young, often through stories, advice, and council.\footnote{This is, in part, why the stories and histories of a tribe are so important. They serve as an integral part of the tribal knowledge which has been building for generations upon generations.}

Following Norton-Smith once again, we can give the following biconditional for what it takes to be justified in Native American Epistemologies, noting that this is closely tied in with the notions of truth and visions which we have seen previously:
\factoidbox{For subject(s), S, her, his, or their purpose or goal g, and action or performance p, p is justified... for S for g if and only if either (1) S directly experiences the respectful success of p in achieving g, or (2) the respectful succes of p in achieving g is endorsed by S's trbal tradition.\autocite[73]{NortonSmith1}}
Using this and the other aspects of knowledge which we have seen in this section, we can construct an account of what it takes to know how to do something according to Native American Epistemologies and we can even make an account of what it takes to know something (have knowledge-that) within those systems. Starting with the procedural knowledge, again following Norton-Smith, we have:
\factoidbox{For subject(s), S, her, his, or their purpose or goal g, and action or performance p, S knows how to p to achieve g if and only if (1) p is true for S for g and (2) p is justified ... for S for g.\autocite[74]{NortonSmith1}}
Although Norton-Smith primarily focuses on the procedural knowledge, the know-how, which seems primary in Native American Epistemologies, it is possible for us to faithfully generalize the various aspects to give an account of propositional knowledge, knowledge-that, by amending the standards, only when necessary, like so:
\factoidbox{For a person, S, the proposition p is true if and only if p is \emph{respectful}, meaning that p is in accordance with morality and would assist in living the kind of life one ought.

For a person S, p is justified for S if and only if S had a vision concerning or related to p (remember that a vision is an integrating experience) AND either (1) S directly experienced p, (2) S is told, in a reliable fashion, that another experienced, or (3) p is a part of S's tribal knowledge. 

For a person S, S knows that p if and only if p is true for S and p is justified for S.
}

Note that the final biconditional does not make reference to beliefs directly. This is because, as mentioned before, the \emph{vision} is what matters in Native American Epistemologies and this appears as a fundamental component of the conditions for justification. 
   