\part{Is Morality Culturally Relative?}
\label{ch.modseven}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}
\input{culturalrelativismrachels}
\chapter{Part 15: What is Moral Relativism?}
\section{Going Meta}

You might have heard the term ‘meta’ before. For example, when I was in undergrad, I was explaining my Epistemology course to some friends, saying something like “it’s the study of knowledge, we are learning about knowledge”, and one person in the group said “that’s so meta!”. The term ‘meta’ comes from Greek, meaning ‘over’, and has there a similar sense to our prefix ‘trans-’ (as in ‘translate’ or ‘transfer’), which comes from Latin and has the same sense as the Greek. In English, also, there are two senses of ‘over’, the first means ‘across’, like to carry something over to another, and the second means ‘above’, as in to put something above another. ‘Meta-’ and ‘trans-’ differ in the same way. With very few exceptions (‘metaphor’, which comes from the Greek for ‘carry over’), the prefix ‘meta-’ means that you are doing something above, over, the area in question and the prefix ‘trans-’ means that you are doing something across, over, certain boundaries. In the case of 'translate', this comes from the Latin verb 'transfero', meaning "I carry over", that verb is also where we get the English verb 'transfer'.

More explicitly, when \gls{meta-} is attached to a field of study, it refers to a field of study one step more abstract than it. One way to think about this is that I am now asking questions about the questions in the field. This is not just true for fields in philosophy, though once the level gets high enough, it becomes an area of philosophy, but it’s also true for any science or subject you pick. Here is a table of some examples:

\noindent
\begin{tabular}{p{1in}|p{1.5in}|p{1.2in}|p{1.5in}}
Subject&Questions &Meta-Subject&Meta-Questions\\\hline
Biology&Concerning living organisms&Metabiology&What exactly is living? 

What exactly is an organism?\\\hline
Physics&Concerning matter and its actions in space and time&Metaphysics&Is matter all that exists? 

What is space? 

What is time?\\ \hline
Linguistics&Concerning language, meaning, and structure in humans&Metalinguistics&How does meaning relate to truth and intentions? 

How do words pick out things? \\
\end{tabular}

\newglossaryentry{meta-}
{
  name=meta-,
  description={a prefix which changes the level of abstraction for the questions in the field, namely by moving it up. This means, roughly, `asking questions about the questions in...'}
}


When you ‘go meta’, you are taking yourself out of the sphere of the field and starting to ask questions about the very nature of it. Due to the nature of philosophy and philosophers in general, we are far more inclined to go meta on a topic. So, for any field in philosophy, there’s likely a well-trotted path for the various meta-questions and sometimes a well-trotted path for the meta-meta-questions.

The various stances which can be generated once you choose to go meta on a topic tend to fall into nice, neat, easy to classify packets. The first level has two different kinds of stances, \Gls{realism} and Anti-Realism (sometimes called \Gls{nihilism}, which is what I prefer). Realism claims that the facts about the subject are real, hence the name, while the Anti-Realist claims that they aren't real. And then, depending on what the stance says about various meta-level questions, they are further subdivided. In the Realist side, there are two general stances which get our attention, Relativism and Objectivism.

\newglossaryentry{realism}
{
  name=realism,
  description={The stance that there are facts in a given topic. In other words, the stance that the facts are `real'}
}

\newglossaryentry{nihilism}
{
  name=nihilism,
  description={The stance that there are no facts in a given topic. In other words, the stance that the facts aren't `real', there is nothing true or accurate to be said about it. The name for this stance comes from the Latin word `nihil' meaning `nothing'}
}


There are different kinds of Relativisms and Objectivisms (the typical subdivisions of Realism) out there, and those tend to be more particular to the field and there may be different kinds of Nihilisms aside from Error Theory and Expressivism, but those go well beyond the scope of this class. Towards the end of this module, I will explain those stances in the Nihilism branch for Meta-Ethics in more detail.

\section{The Start of Moral Relativism}

Moral Relativism is a part of a family of stances, all of which say that the truth of something, or everything, is relative (generally called \Gls{relativism}). What is relative or to what it is relative all depends on the stance. This sort of idea is found in several areas of philosophy and the notion (applied differently) is found in physics. This, in its broad sense, is the stance that there is no absolute truth (objective truth), but rather the truth about things varies from person to person or from culture to culture, depending on the version of relativism which is held.

Relativism comes in many different flavors. The first is that everything is relative, this is global relativism. The other is that only certain things are relative, this is limited relativism. From this, we have to ask what are they relative to? These are either to your culture or to you individually.

\newglossaryentry{relativism}
{
  name=relativism,
  description={The stance that there are facts in a given topic but those facts are variable, they change according to the culture, context, or individual. In philosophy, generally, the facts change according to either what the individual believes or the culture in question believes, depending on the form of relativism held}
}


\glspl{moral relativism} is a limited form of relativism. It does come in two flavors, Moral Individual Relativism (what is moral is relative to the doer, and there is no way for another person to say that what they did is/was morally wrong) and Moral Cultural Relativism (what is moral is relative to the culture and there is no way of saying that they are wrong from an outsider's perspective). One can also say that Moral Relativism is the stance that there is no objective basis for claiming that a particular action, or a class of actions is right or wrong, permissible or impermissible. Or, that there is no fact of the matter regarding whether actions are right or wrong. Whether something is, in fact, right or wrong is relative to the culture or to the person. Some classic examples of this are killing babies, cannibalism, and funeral rights.

\newglossaryentry{moral relativism}
{
  name=moral relativism,
  description={The stance that there are facts about morality, that is there are standards of right and wrong, but those facts are variable, they change according to the beliefs of culture or individual in question (which depends on the form of Moral Relativism claimed to be true)}
plural=Moral Relativism
}


\section{The Cultural Difference Argument}
Often, people argue for moral relativism by pointing to examples, it is often claimed that due to the differences in moral beliefs, there must not be an objective, absolute morality. If you remember from my charts, the moral relativism still holds that there are moral truths, but those truths are relative to the culture. Cultural Moral Relativism is by far the most common stance found in popular culture, so that's the one we will point to, but individual moral relativism has similar problems as well as others unique to it. Here is a classical example used to argue for Cultural Moral Relativism:

    \factoidbox{Darius, a king of ancient Persia, was intrigued by the variety of cultures he encountered in his travels. The Callatians (a tribe of Indians) customarily ate the bodies of their dead fathers. The Greeks practiced cremation. Darius thought that an understanding of the world must include an appreciation of such differences. One day, he brought some Greeks who happened to be present and asked them what they would take to eat the bodies of their dead fathers. They were shocked and replied that no amount of money could persuade them to do such a thing. Then Darius called in some Callatians, and while the Greeks listened asked them what they would take to burn their dead fathers' bodies. The Callatians were horrified and told Darius not even to mention such a dreadful thing. The two people had a very similar reaction to the opposite actions.\autocite{Herodotus1}} 

From this, we can see that two cultures, when the encountered the customs of the other were equally mortified. but sometimes, even the most 'socially aware' of us will be mortified by customs found in other cultures. Take this example:  

    \factoidbox{In some Inuit cultures, there is a common practice know as ‘the dip’ (my translation). When a child is born in the dead of winter, it is common for the mother to carve out a hole in the ice and place the child in the water. This kills the child instantly (infanticide). This is done purely at the parents’ discretion and there is no negative stigma about it. Old people also, when they became too feeble to contribute to the family, were left out in the snow to die.}

The lesson from these two stories (the second of which I have seen it used as examples enough for me to think it's true) is that different cultures have different moral codes. What is seen as just fine in one culture is seen as horrifying in another. For a fictional case, take this example:

    \factoidbox{For most of the world, the elderly are treated with respect and cared for until their natural death. The Kaelon people are different. In this society, the prevailing view is that it's the duty of the 'elderly' to leave their tasks to the next generation, that forcing the next generation to care for the elderly is cruel,  and that having death come for a person  seemingly randomly is heartless. So, when a Kaelon turns 60 years old, they undergo the Resolution. In it, a great party is thrown, celebrating their life and accomplishments, and afterwards, the Kaelon commits suicide. Living past this point is seen as greedy, their time is up and their accomplishments after this point are seen as stolen from the next generation. This is expected of all Kaelons, refusing to kill yourself at the allotted time will cause even family members to be ashamed of you. If a Kaelon seeks asylum to avoid the Resolution, the Kaelons will declare war on the other culture in order to kill the person.\autocite{HalfaLife}}

Many have looked at examples like this and have thought that there must not be an actual objective truth about morality, that it must merely be opinion, either held by a person or collectively as a group. This leads us to:

\section{The Cultural Differences argument}

These examples, and my not so subtle  hints, point to a very interesting argument. It seems, given how I have framed it, the most socially conscious stance would be one which puts all cultural customs on equal footing, to not have some objective measure which can be used to discriminate against a culture. So, using this observation, we get this argument:
\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
Cultural Differences Argument&Disagreement Argument\\\hline
    Different cultures have different moral codes.&There is disagreement about morality.\\
    Therefore, there is no objective "truth" about morality.&Therefore, there's no objective fact of the matter about morality.
\end{tabular}

But, does this show that the next module in this class is going to be gibberish? Well, no. Though there are some moral relativists out there in philosophy today, there are very few, and no philosopher (no really good philosopher) is a moral relativist because of this argument. There are many problems with it which are worth exploring. 

\subsection{Problem 1: It isn't a good argument}

For an argument to be good, philosophers have a pretty high standard. The conclusion needs to follow from the premises necessarily. There can be no case where the premises are true and the conclusion false (this is called ‘validity’). If an argument is not valid, then it is not sound. The first line of the argument concerns what people believe while the conclusion concerns what is actually the case. The issue with this sort of reasoning can be seen using a similar sort of method as the Limitation of the Conclusion problems in the arguments for the existence of God. If you think that The Cultural Differences Argument is accurate, then you will, by the same reasoning, need to think that these arguments are accurate: 
\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
Shape of the Earth Argument&Global Warming Argument\\\hline
    Different cultures/people have different beliefs about the shape of the earth.& Different people have different beliefs about global warming.\\
    Therefore, there is no fact of the matter about the shape of the earth.&Therefore, there's no fact of the matter about global warming.\\
\end{tabular}

This argument follows the exact same reasoning as the cultural differences argument. If it works for one then it should work for the other. But, we can also look at the spirit of the argument, and attack that as well. 

\subsection{Problem 2: Cultural Oppression}

Though the argument may not be any good, all that shows is that the argument isn't good. It does not show that morality is objective. But, if we look at the last example I gave, involving the Kaelons, we see something interesting. If morality is relative to a culture, there's nothing stopping one culture, morally speaking, from oppressing people in their culture and oppressing people in other cultures. For example, if there are no objective moral facts, then there would be nothing wrong with the Kaelons declaring war on another culture for allowing someone to receive asylum. Similarly, there would be nothing wrong with the Kaelons forcing members of their culture to unwillingly commit suicide. This sort of case can be found in history as well, with the sear number of examples being too numerous to list. This is a problem which will appear again later in this class, because it's more obvious inside another argument.  

\section{The Cultural Imperialism Argument}

The Cultural Differences Argument is not the only argument in favor of Cultural Moral Relativism, there are several others. Most fall to the same sort of objections which we will see later, others have problems of their own. The Cultural Imperialism Argument is a little more advanced, claims to prove this sort of relativism, but it, equally, has issues.  Here is the introduction:

\subsection{Cultural Imperialism}

Throughout history, we have seen cases of one culture forcing beliefs and moral stances onto another. For example, we have the forced religious conversions of the Aztecs by the Spanish, we have the boarding schools which the US put Native American children in, and in China, we have the Uyghurs being put into 'reeducation camps' (this may still be happening, this is true as of 2018). All of these examples have/had the same purpose: changing the practices, religions, and moral ideologies of the people.  

Cultural Imperialism is the reason for these behaviors, why cultures force themselves on others in this way (and sometimes in even more violent ways). Cultural Imperialism is the stance your culture as all of the right answers, morally speaking, your culture is perfect, and because of this, you are justified in forcing other cultures to change (typically to your own). But, looking at the number of examples from history, we can see that this imperialism is wrong, it should not be done. This can be seen both from an insider's and an outsider's perspective. Naturally, this perspective on Imperialism leads easily to Cultural Moral Relativism, because it seems never to be right to judge the values of one culture against the values of another. Put more formally, we get this argument:

\subsubsection{The Cultural Imperialism Argument}

\begin{enumerate}
    \item Cultural Imperialism is morally wrong. (Intuition).
    \item This follows from the examples which I have given, many argue that this sort of behavior is wrong and should not be done at all. The main reason for the Prime Directive in Star Trek is because of the base line principle, you need to let cultures' internal matters be internal matters. 
    \item If Cultural Imperialism is morally wrong, then it is wrong to judge the values of one culture against the values of another. (Consequence of it being wrong)
    \item This, in turn, is a consequence of the wrongness of Cultural Imperialism (the first line). Basically, if it was OK to make judgments about other cultures' practices, then it would be OK to act on those judgments. Acting on those judgments would be Cultural Imperialism, which we said was wrong.   
    \item If it is wrong to judge the values of one culture against the values of another, then no person is ever justified in criticizing the moral norms of another culture. (Consequence of the consequence)
    \item This is a consequence of the wrongness of the judgments about a culture. Criticizing a culture's practices, in either word or action, is wrong by the same reasoning as Cultural Imperialism. The exact reasoning is found in the previous line. 
    \item If no person is ever justified in criticizing the moral norms of another culture, there are no non-relative moral truths (they are all relative to culture). (Consequence)
    \item This is the final line of support for the conclusion. If there were an absolute, objective, moral rule, or set of rules, then it would be possible for a person to be correctly justified in criticizing another culture's norms. All they would need to do is point to the objective rules and show that the culture is in violation of them. But, we have shown that a person can never be justified in this way, so, it would seem that there must not be an absolute moral standard (they are relative to the culture).
    \item Therefore, there are no non-relative moral truths (they are all relative to culture).
\end{enumerate}

Now, despite my build-up and explanation for this, there is a major issue with this argument, both in the way it's constructed (though, technically valid) and the spirit behind it. For example, I can use this argument, exact same premises, to show that unicorns exist, 2+2=5, and anything else I could want. 

\subsubsection{The Problem}

As I mentioned above, I can use this argument, not the structure, but the argument itself, to prove that mermaids exist. Any argument with this capability will be valid, but it will not be sound, there will be a factual error in it somewhere. For this one, let's try and work out the factual error, and here it's fairly easy to spot (hint, look at the first line and the last line):
\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
First Line&Last Line\\\hline
Cultural Imperialism is morally wrong.&There are no non-relative moral truths.
\end{tabular}
The problem with this argument is that it's contradictory. You are saying that the same is both true and false. In the first line, we are saying that Cultural Imperialism is wrong, which is not contradictory on its own (if you think that it can be OK, then you have included other aspects into it). Similarly, the last line says that all moral truths are relative. Again, not contradictory on its own. But, in combination, the lines conflict. The first line is a non-relative moral truth and the last line says that those don't exist. This also attacks the spirit of the argument, in that the core premises/intuition which build it up come from the idea that Cultural Imperialism is wrong, which contradicts the point it's trying to prove. 

There are a few ways out for the Relativist. First, they could accept that some cases of Cultural Imperialism are fine, making it not always wrong. For example, we can see that the Allies entering WW2 to stop certain behaviors was Cultural Imperialism, but that wasn't wrong. The issue with this is that if we go down this rabbit hole, we find that the cases where Cultural Imperialism is permissible are cases of objective moral truths, which limits the initial scope of Objectivism, but contradicts Relativism. 

Another way out, which limits Cultural Moral Relativism further, is to say that there's a distinction between internal and external behaviors. Internal behaviors are those which only apply to members of the same culture, while external behaviors are those which involve members of different cultures. Cultural Imperialism is an external behavior. The Relativist here can claim that the morality of internal behaviors is relative while the morality of external behaviors is objective. This has its own issues as well, mostly dealing with cases like the Kaelons which I gave before, and can be easily seen in that case. 

\section{Tests to Tell Whether You are \emph{actually} a Relativist}
These are some tests which you can preform on yourself or on others, or on groups, which will tell you your actual feelings about moral relativism. There are three tests total, sometimes these tests will work individually, sometimes jointly (depending on the case), but they can be very useful. These are based on \emph{Why I am an Objectivist about Ethics (And Why You Are, Too)} 
by David Enoch\autocite{Enoch1}.

\subsection{The First Test: The Spinach Test}

This is a pretty straight forward test for whether you actually are a moral relativist. But, as a fair warning, philosophy kills jokes. As an example, and where the test gets its name, lets look at the following joke:

    \factoidbox{A kid hates spinach, later, he says that he’s glad he hates spinach. When asked why, he says “because if I liked it, I would eat it; and it’s yucky!”}

More often than not, we will take this as funny. And, in a lecture format, if I present it right, I will get a laugh. But, the test is not whether or not you laugh at this joke, but rather whether you laugh at jokes of this form/structure, with different things put in the place of 'spinach'. For example, let's try it with entertainment and other foods:

    \factoidbox{A kid hates watching golf. He says that he’s glad he hates watching golf because “if I liked it, I would watch it, and it’s boring!”

    A metal-head says that he’s glad he hates country. When asked why, he replies ‘because if I liked country, I would listen to country, and that’s just bad music.’

    George H.W. Bush hates broccoli and says that he’s glad he hates broccoli. The reason? “Because if I liked broccoli, I would eat it, and it’s gross.”}

These are the same, if presented right, they will get a laugh. But, will this joke work for anything which I plug into the joke's formula? What if I plug in something which is more absolute? Take these examples:

    \factoidbox{A 20th century man believes that the Earth is round. He claims that he’s glad that he’s a 20th century man because “If I grew up in the first century, I would have believed that the Earth is flat, and that's just wrong.”

    A 20th century man believes that women should have the right to vote. He claims that he’s glad that he’s a 20th century man because “If I grew up in the 19th century, I would have believed that women shouldn’t vote, and this is wrong.”

    An Iraqi woman who grew up in the 2000s believes that women are not equal to men. She claims that she’s glad about this because “if I grew up in the 60s and 70s, I would have believed that we were equal, and this is wrong.”}

Something is very different here. If I tried to present these as a joke, even with the same tones and inflections, I would not get a laugh (and I don't, I have tried). But, what's the difference between these? What sort of things get a laugh and what sort of things don't? Well, the difference between these cases is seemingly unrelated, but it's the only one which works (they are related). The relative things, the things which are merely preference, those get a laugh when plugged into the joke, the things which aren't relative, the objective things, those don't get a laugh.  So, this test is basically if you get a laugh, it's relative, if you don't, then it's objective. The vast, vast majority of the time, we do not laugh about morality. So, we, at least at some level, think that morality is objective. 

\subsection{The Second Test: The Disagreement Test}

Often, in real life, we get into disagreements with people. Sometimes these are serious, like whether refugees should be allowed in the country or whether global warming is occurring. Other times they are just silly, like whether broccoli is yucky or whether dark chocolate is better than milk. The second test for relativism about a subject concerns the nature of such disagreements. For this test, you need to  imagine, or actually get into, a disagreement about something. Ask yourself what the disagreement feels like. For example, take these examples:
\factoidbox{
\begin{tabular}{p{2.5in}|p{2.5in}}
Disagreements A&Disagreements B\\\hline
A disagreement about whether the Earth is flat.&A disagreement about whether pineapple should go on pizza.\\\hline
A disagreement about a mathematical principle.&A disagreement about how enjoyable Buffy the Vampire Slayer was.\\\hline
A disagreement about whether refugees should be allowed in the country.&A disagreement about the direction of a toilet paper roll.\\\hline
A disagreement about whether slavery was bad.&A disagreement about the best metal band.\\\hline
A disagreement about whether God exists.&A disagreement about flavors of ice cream.\\\hline
A disagreement about what to do in the trolley problem.&A disagreement about a movie (how enjoyable it was).\\\hline
\end{tabular}
 }

These disagreements are very different. Think about what it’s like to be in those kinds of disagreements. In the case of broccoli or chocolate, you are trying to state your own preference and then, maybe, trying to get the listener to come to their senses and change theirs. However, when we are engaged in a disagreement about, say, global warming, the point of the argument is not to change the other person’s mind or to state how you feel. Rather, it is to get at the truth. Looking above at the examples, do the ones in the left column feel the same? Do they all feel like facts? Similarly, do the ones in the right column all feel the same? Do they feel like preferences? If the feeling of the disagreements in the left column all feel the same, then you are an objectivist about morality, in your heart.  

So, for this test about whether you are a relativist about some subject, you need to ask yourself what it feels like to be in a disagreement about the subject. Imagine that you are having a disagreement with someone about the morality of abortion. How does that feel? Does it feel like you are having a discussion about pineapple and green-olive pizza? Or does it feel like you are having a discussion about global warming?

Yet again, very rarely do we feel like disagreements about morality are disagreements about preference. When we have those debates, it feels like they are about facts.
\subsection{The Third Test: The What-If Test}

Like the other two, this one is a bit different. It involves what are called counterfactuals. They are called this because they are counter to the facts. We use these sort of statements all the time without realizing it. Any time you ask yourself what would happen if something were the case, you are using a counterfactual. These are insanely useful in philosophy as well as in science. There's an entire cottage industry in philosophy dedicated to coming up with how we use, accept, and reject these kinds of statements. That being said, since we use them so much, your gut intuition will suffice. For this test, we will present the subjects as yes-no 'what if' style questions. Such as, ‘what if I drove a 110MPH on a 60MPH highway, would that be dangerous?’ It is not true that I drive that fast, but I am trying to figure out what would be the case if I did. Here are some examples of what-if style questions which show this test at work.
\factoidbox{
If the vast majority of people believed that politicians were actually aliens, would they still be human?

If the vast majority of people believed that global warming wasn't happening, would it still?

If the vast majority of people thought that gender discrimination was fine, would it still be wrong? 

If the vast majority of people thought that the Earth was flat, would it still be round? 

If the vast majority of people thought that Philosophy was pointless, would it still be valuable? 

If the vast majority of people thought that pineapple on pizza was delicious, would it still be gross? 

If the vast majority of people wore top hats and thought they were cool, would they still be out of fashion?

If the vast majority of people thought that Buffy was the worst show ever, would it still be great?

If the vast majority of people thought that sweet potatoes were tasty, would they still be gross?

If the vast majority of people thought that spinach was gross, would it still be tasty? 
}
 

For each of these, there should be a yes or no answer, one which you can easily give. In some versions of this, I have presented it as "all people" rather than "the vast majority of people", but that has lead to some confusion. As before, the normal responses to the questions in the left column are all 'yes', while the answers to the questions in the right are normally 'no'. This test, if the question is formulated correctly, is the most definitive, I think. It points out the fact that objective truths are not up-to-us, they will be the case regardless of whether or not people discover them. 

The steps to preform this test on yourself are pretty straight forward. First, take a stance about something which you think is true. For example, you can have that slavery is bad, that pineapple on pizza is good, that the Kaelons shouldn't force their elderly to commit suicide, that the Earth is round, or something like that. Next, you imagine a case where the vast majority of people believe the opposite. And, finally, ask yourself whether, in this strange world, you would still be correct.  If the question is phrased correctly (it's possible to phrase them so that you will get, consistently, the opposite answer) and if you get the answer 'yes', then the subject matter is objective and if you get the answer 'no', then the subject is relative.

\section{Consequences of Taking Moral Relativism Seriously}

So far, we have seen evidence that we don't actually think that moral relativism is correct and we have seen that the arguments for the stance aren't any good (it's possible that there are others, and likely are others, but they will all, likely, fall you the same issues). But, like with the arguments for the existence of God, an objection to the argument doesn't show that it's conclusion is false. There are some true or false things which just can't be proven one way or the other (for example, "the first digit of TREE(3) is 1" is either true or  false, but due to the size of the number, it's just impossible for us to prove one way or the other). Now, to actually disprove Cultural Moral Relativism we need to attack it head on, not just the supporting evidence. To do this, we will use a method which you might recognize from science (science got it from Philosophy). We will assume that relativism is true and show that it does not line up with the world, or that the consequences of it are so outlandish we can dismiss it easily. There are, at the heart of Moral Relativism, 3 problems which we will address (and more, but 3 is enough). 

\subsection{The Criticism Problem}

If we remember back to the Cultural Imperialism Argument, one of the core reasons it works is because we want to say that criticism of other societies (through word or action) is morally wrong. Though this does contradict Cultural Moral Relativism, that stance does have a variant to it built in. Namely:
\begin{center}
We could no longer say that the customs of other societies are morally inferior to our own
\end{center}
In other words, if moral relativism is correct, we can't say, truthfully/correctly, that any culture has 'got it wrong', we can't say, in that way, that any cultural behavior is wrong. Sometimes, this seems fine, like in cases where doing so leads to the wrong sort of Cultural Imperialism. Other times, however, it leads to some very outlandish conclusions like:
\factoidbox{
We can't say that a culture torturing and/or killing other groups systematically is morally wrong.

We can't say that intolerance is bad.

We can't say that cultures which engage in hateful acts towards others are morally wrong.

We can't say that tolerance is good.}

The argument for this conclusion rests on the idea that if we accept moral relativism, there would be no gauge, objective standard, to measure morality.
\begin{enumerate}
    \item If MCR is true, then morality is relative to culture. (This is basically the definition)
    \item If morality is relative to the culture, then there is no standard to gauge the morality of a culture.
    \item If there is no such standard, there is no way to compare one culture to another (morally speaking)
    \item If there is no way to compare them, then there is no way to say that the Nazis were bad.
    \item So, if MCR is true, then there is no way to say that the Nazis were bad.
    \item But, the Nazis were bad!
    \item Therefore Moral Cultural Relativism is false.
\end{enumerate}
The way out, for the relativist, is to claim that the Nazis and every other culture which has engaged in horrible acts weren't bad. This can be quite the pill to swallow. 

\subsection{The Sheeple Problem}

Some of you may have heard the term 'sheeple' before. It's a mash-up of the word 'sheep' and the word 'people'. In this case, if we take Moral Cultural Relativism seriously, thinking for yourself about morality is always morally wrong. This is not directly because of some universal moral rule, which would contradict Moral Relativism, rather it's because if something is a moral rule, then not abiding by it is morally wrong, and since every culture will have moral rules, not blindly following them will be wrong. Rather than speaking in negatives, we have this statement: 
\begin{center}
We could decide whether actions are right or wrong just by consulting the standards of our society.
\end{center}
Moral Cultural Relativism gives us a very simple test to tell whether an act is right or wrong. Just look at the cultural standard. Also, at the same point, trying to change those standards, whatever they may be, is breaking those standards, and thereby wrong. This means that, for example: Suppose that a person in the deep south (Pre-Civil War) was curious about whether slavery was permissible. All they would need to do is ask whether it fit with the code of the society. If it did, then they would be OK with having slaves. Another way to think about this is to imagine that God, or whoever, once the criteria for being a culture are met, sends down a book with the rules. Doing anything other than blindly following those rules is wrong, regardless of the culture. This means that every civil rights movement in history, in every culture, was morally wrong. This includes Black Rights Movements and Women's Rights Movements.

But, no cultural system is ever perfect and they have room to improve. Not only can’t we criticize other cultural codes, but we can’t criticize our own. However, claiming that every civil rights movement in history was morally wrong is absolutely outrageous, so Moral Cultural Relativism must be incorrect.  

This leads us to the following argument:
\begin{enumerate}
    \item If MCR is true, then morality is relative to culture. (This is basically the definition)
    \item If morality is relative to the culture, then whether an action is right or wrong is determined by your cultural norms.
    \item If those are determined by the norms, then trying to change those norms is always morally wrong.
    \item So, if MCR is true, then trying to change moral norms is always morally wrong.
    \item But, trying to change moral norms is sometimes morally right.
    \item Therefore, Moral Cultural Relativism is false.
\end{enumerate}
The way out for the relativist here is to claim that moral change is always wrong. Basically, claim that Martin Luther King Jr. was about as immoral as they come.

\subsection{The Progress Problem}

Many of us, especially some of the more 'woke' members of our society, will claim that some moral change is for the better. But, the question quickly becomes "what is 'better'?" When we use this term, we are comparing one thing to another. We say that "pineapple and green-olive pizza is better than pepperoni" or "Pushing Daisies is better than The Big Bang Theory" or "carrots are better than potatoes", in the context of healthiness. But, when we say that some moral change is for the better, what are we comparing and what are we using to make the comparison?

When we claim that some moral change is for the better, we are comparing where we were, morally, and where we are, morally. To make this comparison, if it's correct, we are using an absolute moral standard, a universal standard. We mean that we have moved closer to getting morality right. We call this progress. If we take Moral Cultural Relativism seriously, there's no moral standard, there's nothing to use to compare where we are and where we were morally speaking. We can't, correctly, claim that there's progress.
\begin{center}
The idea of moral progress is called into doubt
\end{center}
This means that not only can't we compare other cultures, but we can't even compare the same culture to itself at a different time. This leads us to the following argument:
\begin{enumerate}
    \item If MCR is true, then morality is relative to culture. (This is basically the definition)
    \item If morality is relative to the culture, then there is no standard to the morality of a culture.
    \item If there is no such standard, there is no way to compare one culture to another (morally speaking)
    \item Different people at different times, if the norms changed, are, for all intents and purposes, different cultures.
    \item So (from 1, 2, 3, and 4), if MCR is true, then there is no way to compare people from one time to another (morally).
    \item If there is no way to compare people from different times, there can be no way for us to say that we are better than we were (made progress).
    \item So, if MCR is true, then there is no way to say that we have made progress.
    \item But, we have progressed (at least in some areas).
    \item Therefore, Moral Cultural Relativism is false.
\end{enumerate}
The way out for the relativist is to claim that we haven't made progress, in any way. We are no better than we ever were. 

\chapter{Part 16: So, If Moral Relativism is Wrong, What Next?}

If you are willing to say that moral cultural relativism is wrong, and, given the general responses to the various problems with moral relativism as well as what can be expected from the tests, you likely are willing to say that it's wrong, we have some problems. Namely, what is moral? Who decides who's correct in a moral conflict? Where do moral truths come from? Here I will cover some of the basic questions which former relativists tend to give and the general responses (sometimes, the response will need to be more particular).

\section{If culture's don't choose, what is morality?}

There are a few ways to go about replying to this question. Morality/ethics centers around 'should' questions of a certain kind. When we ask questions like these, for example "what should I do?", "should you call a doctor?", "should you get the assignment in on time?",  there are two different senses, which may or may not overlap, depending on context and relevance. Something is moral/ethical when it's the correct answer to the question "what should I do?", when we aren't talking about practical cases (a practical case is one where you are asking about the correct way to perform some task, like changing the oil on your car). The cultural moral relativist would claim that the correct answer to the question in non-practical cases (for example, should I flip the switch in the trolley problem?) will depend on your culture and the norms associated with it. The moral objectivist, a moral realist who is also a non-relativist, would claim that there's an actual answer to this question, which is not determined by your culture. The next module of this class covers some of the ways which philosophers have answered how to answer that question, without being relativist. Ethics is, at it's core, trying to find the correct answer to this question.

You will notice that when I give examples of cultural norms, most of the time, these are cases where the culture got the answer wrong. Ethical theories are basically hypotheses about the correct way to get the answer to the question. Some theories point to absolute, objective duties which a person has (typically, it's wrong to be irrational; acting in a certain way is irrational, therefore those ways of acting are wrong). Others point to the well-being of those affected (if an action makes the affected better off than otherwise, that's the right action). Others still point to some exemplar of morality, some person who has the perfect character and then asks "what would that person do?". You may have heard that kind of thinking before with 'WWJD'.  

\section{Who decides who's right in moral conflict?}

A moral conflict is a case where two or more people/groups disagree about the answer to the "What should I do?" question, in the relevant sense. The moral cultural relativist has a simple way to answer it, almost too simple, "if it's two groups in the same culture, the cultural norms settles the conflict, if it's two cultures, there's no real conflict." But this just does not seem right, if you recall the last page, were this true, all civil rights movements would wrong, because the cultural norms would settle the conflict in favor of the oppressing group. Very few of us would want this. So, that's out, but is there a non-cultural way to get the answer? This is the quest for the moral realist. It's not going to be based on belief (as the relativist points to).

Recall the Cultural Differences Argument. This argument relies on moral conflicts/disagreements. But, do we always settle debates like that? There may be disagreement about something, anything (for that matter), but it doesn't follow that there's no fact of the matter. If there's a conflict about math, we don't say that there's no answer, we consult the rules of mathematics (as they were discovered, not invented). When there's a conflict in science, we don't say that there's no answer, we perform experiments and discover the truth. For the moral realist, settling moral conflicts is more like settling a conflict in math or science than one in art. Realists perform experiments and consult the rules of morality to settle the debates (and the experiments either further support the rules or give examples of amendments which should be made). The quest to get the non-relativist rule to settle moral conflicts is hard, but it does not seem impossible, we do make progress in it. There's even a scientific-style method for figuring out which ethical theories should be applied and/or how they should be amended (as we will see in the next module).

In short, no one decides who's right in a moral conflict, it's just a fact that one is right and the other is wrong.

\section{Where do moral truths come from?}

Moral truths are the answers to the "what should I do?" question (in the relevant sense). The moral cultural relativist has a simple, again almost too simple, answer "from the beliefs of the culture." Yet again, how can we possibly say that a culture has the wrong answer to the question when the culture chooses the answer? Like the issues before, this just can't be right. So, the question for the realist is just that, where do they come from? 

For the realist, this is like asking "where do scientific truths come from?" or "where do mathematical truths come from?". Some moral realists point to God and say that all truths flow through Him, moral truths included, saying that God made them (an easy way to get all-knowing and all-good). Others point to the sort of creatures we are and our place in nature, these give us moral truths. Others still point to abstract notions of well-being or rationality.  And even more say that moral truths are things which just could not have been otherwise, they just always are, they didn't "come from" anything. There is some debate about where they come from.

But, don't lose heart! There's just as much debate about where mathematical and scientific truths come from. For example, in the case of math, we could claim that the truths are necessary, they could not have been otherwise. We could claim, with just as much evidence, that the truths are constructs of the definitions of the terms which we are using overlapping in consistent ways (1+2=3 because of the definitions of the terms and operations). And we could also claim that the truths came from God.  As it turns out, in philosophy, the most common stance (for where mathematical truths come from and where moral truths come from) is that they always were and always will be.

\section{What about forcing them onto another culture?}

This is a hard problem, because one of the core intuitions behind the Moral Relativist, and one of the big reasons why the stance is contradictory, is that it’s wrong to force your morals onto another. The Moral Objectivist will claim that under the right circumstances, it’s morally required that you force another to change. To some, this feels wrong. This feeling, often, gets traced back to the horrors of history, which I used as examples for the intuition behind the Cultural Imperialism Argument, where one group imposed themselves on another. At the same point, we need to look at history as a whole. There have been times when it was actually correct for one group to come in and force a change. In those cases, the imposing groups had real morality on their side and in the cases where it was wrong, we can say that they were mistaken about morality. For example, the Nazis in WW2 were just wrong about morality and the groups which came in were right. So, the Moral Objectivist just needs to be sure that their moral theory is getting the right answer in the case. This is an epistemological worry for the objectivist, not an ethical, meta-ethical, or metaphysical one. 

Interestingly, when we look at our own history, we find things which we are ashamed of, things which we condemn the previous generation for. How and when this is correct is easy for us to see. Since different people at different times are, basically, different cultures, the exact same, or a similar, thought process applies when condemning or appreciating the behaviors of another culture. Kwame Appiah in What Will Future Generations Condemn Us For?\autocite{Appiah1} gives us some compelling tests for when future generations will be ashamed of our practices. These same tests apply to the behaviors of other cultures and tell us when it may be OK to force a change. 

\section{Moral Nihilism}

Moral Relativism and Moral Objectivism (Realism) are not the only meta-ethical stances, there are two others worth noting (though this does go beyond the necessary scope of this class). Like Moral Relativism, these stances deny that there are absolute moral truths. But, unlike Relativism, these theories claim that they aren't relative. They claim that there's no truth to them at all, regardless of context. Theories which do this fall into a category called 'Moral Nihilism'. Like Relativism, Nihilism comes as a family and you can pick and choose accordingly. Skepticism is most like Nihilism in how the family tree is organized. 

Moral Nihilism, like Moral Relativism, is limited, it only makes claims about moral claims. Like Relativism, if you find one absolute, objective truth, then Global Nihilism is false, but for Limited Nihilism, you would need to find it in the context it's limited to. The two major theories which call Moral Nihilism home are 'Error Theory' and 'Expressivism'. Of the two, Error Theory is the strongest, but it can be quite counter-intuitive. 

\subsection{Error Theory}

Error Theory, applied differently, is a concept and stance which can be found in other areas of philosophy, though I have personally mostly encountered it in Metaphysics. It is a lot like Skepticism, but rather than saying that we will never know about the thing, it says that we will always be mistaken.  For Meta-Ethics, this is the stance that all of our moral judgments will be mistaken. We will always be in error, hence the name. This is true even if we were to randomly chose an answer. For the \glspl{error theory}, we will ask questions about a subject, sincerely try to answer them, but will always get the wrong answer. There is no right answer.  Applied in other fields, it claims that all of our judgments in those fields will always be in error. They come to this conclusion from three assumptions, or starting points, which get them off the ground:

First, they need to have the standard-issue feature of a moral nihilistic stance, this is that morality is not a feature of the world. In other words, there are no moral facts. Some could say that morality is a useful fiction or say that morality is self-contradictory or some other means. The second feature is a bit more precise to this stance. This is the claim that no moral judgments will ever be correct/accurate.  This follows, using some fairly simple reasoning, from the first feature. For our judgments to be accurate/correct, they must correctly/accurately depict something. However, because of the core premise to Moral Nihilism, there's nothing for the judgment to correctly depict, so it can never be accurate/correct. The third aspect follows suit, they claim that it's pretty obvious that people try to use moral judgments, we act on them, and we think about them, but those judgments will always be in error, hence the name. Since our judgments can never be correct, they will always be wrong. The error theorist is making a pretty big claim. They aren’t just trying to bash on social policies and individual actions. They are going all out and claiming that morality is a fiction.

\newglossaryentry{error theory}
{
  name=error theory,
  description={A general claim that all other claims within a field of study or context will always be mistaken, try as we might to say something truthful, we cannot because there are no truths in that area or context},
plural=Error Theorist
}


They are essentially the atheists of ethics. Atheists hold that there are no religious features in the world, the error theorist holds the same about ethical features. The atheist holds that, try as you might, all religious doctrines are mistaken because (for example) God does not exist. The error theorist holds that, try as you might, ethical judgments/beliefs are all mistaken because morality does not exist.

\subsubsection{For Error Theory to be Correct}

All error theorists hold that the basic mistake in moral thinking is that it depends on some absolute or objective truth about morality. This truth applies to all of us, regardless of who or what we are. These are what some refer to this as ‘objective morality’ and ‘categorical reasons’.

To get error theory off of the ground, they need to convince us of 2 things. First, morality really does depend on these absolute standards (they would need to come up with the correct moral theory, examples are in Module 8). Second, supposing that it does, they have to show us that at least one of these is false, which would mean that the correct theory to describe morality is fundamentally and irreparably flawed. So, for the Error Theorist to prove themselves, they need to solve all problems in Ethics and then show that the stance which did it couldn't be right. This is quasi-contradictory.

\subsection{Expressivism}

Expressivism could, in principle, be a stance in other areas of philosophy and maybe, for some of them, it could make a solid stance, but I have not encountered it so much in my time and, when I have, it has almost exclusively been in Ethics. As I mentioned before, given the core flaws in trying to prove Error Theory (in the case of Ethics, in other fields this issue is not present), 
\Gls{expressivism} is the stronger of the two. As it is a Nihilistic stance, it does come with two assumptions, but it differs from Error Theory in the last aspect.

\newglossaryentry{expressivism}
{
  name=expressivism,
  description={A general claim that all other claims within a field of study or context are not intended to and do not express anything true or false, rather those claims are expressing emotions, commands, or questions. In ethics, saying something is wrong, according to this stance, is the same as saying something like `boo' and saying that something is right is the same as saying `yay'},
plural=expressivist
}


First, as with the Error Theorist, the Expressivist claims that morality is not a feature of the world. But, it doesn't say that it's a useful fiction or anything like that. It also claims, but the same reasoning, that moral judgments are never accurate. But, it does not say that they are always wrong. I know that this last bit seems odd, so let me explain: The Moral Objectivist (Realist), Relativist, and Error Theorist all share one thing in common: When we make moral claims, we are, at least, trying to describe the world, pick out some feature in it. The Expressivist is different: They claim that when we make moral claims our intent is to express something, not describe. They think that our ethical statements/judgments aren't the sort of things which can be true/false, correct/incorrect, or accurate/inaccurate. Take these two examples: 
\noindent
\begin{tabular}{p{2.75in}|p{2.75in}}
Example A&Example B\\\hline
    Murder is wrong&Blue is a color
\end{tabular}

Although it may look like these examples are stating the same sort of thing, and, in fact, were we to  do a sentence/semantic diagram for these examples, they would look freakishly similar, according to the Expressivist, these are fundamentally different sorts of sentences. In linguistics, profanity is often claimed to not add any meaning to a sentence, but rather is an emotional or rhetorical expression; the Expressivist says, basically, that moral claims, like (A), are the same sort of thing, they are only emotionally/rhetorically meaningful. In the case of (B), it is expressing something true about the world, there's a collection of things, colors, and it's putting 'blue' in that collection (by one theory) or there's a thing in the world, 'blue', and assigning a feature to 'is a color' (by another theory, this is what I really work in). But, in the case of (A), we aren’t assigning a feature to murder, according to the Expressivist. Rather it is like we are saying something like:
\begin{earg}
    \item[]BOO murder!
    \item[]Don’t murder!
    \item[]Let’s not murder!
    \item[]Wouldn’t it be nice if we didn’t murder?
\end{earg}
Those sorts of statements aren’t the kinds of things which are true or false, they are more like actions or commands or questions (for interesting content on statements as actions, check out J.L. Austin's How To Do Things With Words). If you have ever been to a highly emotionally charged (peaceful) protest, people screaming statements which boil down to "this is wrong" will feel more like "boo this thing" rather than something based on an intellectual process, which is evidence in support of this sort of stance. 

\subsection{Three Concerns for Expressivism:}

There are three major worries for Expressivism. Although, I will say, Expressivism is the most popular, it is by far the weaker of the two stances and it will need some time in the gym, so to speak, to really hold its own. The first worry for the Expressivist concerns how we really reason about morality. Take the following arguments:

    \begin{earg}
    \item[]All cases of hurting people are immoral.
    \item[]Torture hurts people.
    \item[]Therefore, torture is immoral.
   \end{earg}

\begin{earg}
    \item[]All men are mortal.
    \item[]Socrates is a man.
    \item[]Therefore, Socrates is mortal.
\end{earg}

In fact, we use the exact same style of reasoning in both cases, agree or disagree with it, the logic is exactly the same. But, according to the Expressivist, the first argument looks like this:
\begin{earg}
    \item[]Hurting people - YUCKY
    \item[]Torture hurts people.
    \item[]Therefore, BOO torture. 
\end{earg}
We reason about morality logically, not emotionally (sometimes we do both, but when we are thinking clearly, emotions play a supporting role, they aren't the main character). According to Expressivism, however, we wouldn't do this, it would only be emotional. This observation, if correct, means that Expressivism can't be accurate. 

The second concern for Expressivism comes from Amoralism. An amoralist is a person who sincerely makes moral claims but is completely unmoved into action by them. These people might be out there, this would be like Data from Star Trek, in that they would not feel the emotions associated with ethical claims. If they do exist, this is a serious problem for the Expressivist. Expressivism claims that moral claims are emotional expressions. But, emotional expressions, most of the time, move us to action (even if we don’t actually act, we still kind of want to). How can a person sincerely make a moral claim, which is supposed to be an emotional expression, without being moved to action?  Basically, if a person can sincerely make a moral claim without being moved by it, then Expressivism is incorrect. 

The final concern for the Expressivist concerns how we actually make these moral judgments, much like the first. The absolutist (objectivist/realist), the relativist, and the error theorist disagree on just about everything, but they do agree on one thing, that moral claims are, at least, trying to describe the world. The Expressivist denies this. They have to paraphrase the moral claims to not have them attempt to describe the world. But, if this were really the case, why wouldn’t we just express our emotions and not conceal them? There are many statements, which are ethical in nature, which the Expressivist can't handle, many of which I have heard people make in the real world, for example:
\factoidbox{
    I’m not sure whether torture is moral, but I think people smarter than me would know.
    
    There’s a difference between being moral, being required, and being praise worthy.
    
    How much we punish should match how bad the crime was.
    
    If war is immoral, then generals are not as good as they seem.}

If we read these like the other stances would, then we have no problems, all are clearly understandable, but if we read them as the Expressivist thinks they should be, we are up a creek. We could all be lying or mistaken, but when we think about what we are trying to do when we make these statements, we see that we are really trying to describe the world. 