\part{Is Death Bad For The Person Who Died?}
\label{ch.modtwo}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}
\input{death}
\stepcounter{chapcount}
\chapter{Part \thechapcount: How Does Nagel Define Death?}\setcounter{seccount}{1}
In the reading, Nagel argues from the perspective that death is just the end of conscious experience, after that point, there are no feelings, no emotions, no thought for the person who died. One could think that there is some kind of after-life; if you think that, then you could think of this as a ‘what-if’ sort of mindset. This could be either a best-case or worst-case scenario. The question which Nagel is looking into is whether death, as we have so defined it, is bad for the person who died. I am not asking whether it is bad for the family or loved ones, as they will certainly be sad and that is bad, but what about the person who died? Is it bad for them?

For some practice in abstract thinking, think about how Nagel is defining death. Could there be cases where a human is dead, by Nagel's definition, but is technically still alive? (Give this a moment's thought before continuing.) One such example could be a case of a human, who was previously perfectly normal with all of the relevant faculties, entering into an irreversible coma. They would be, to use a common phrase, brain-dead. Their heart would still be beating and maybe they would still be breathing (we can easily assume that they could need a breathing machine), so they would still be alive, but without the ability to have or process experiences, they would be dead. There could be other examples like this, but you need to be careful because it needs to be the permanent end of conscious experience (so falling into a deep sleep doesn't count).  

So, if death is bad for the person who died, it can't be because of what it has (because death, as defined here, is the absence of everything). If it's bad, it must be because of what it lacks. Nagel lists four (4) different generic `things' which living has but death lacks. 

First, we have perception. Perception is the ability to take in and process stimuli. This is not necessarily the stimuli we get from the external world, like sight, sound, smell, and touch, but also includes the stimuli we get internally from us thinking. Death, since it doesn't have consciousness, can't have any perception (either internal or external). 

Second, we have thought. Thinking is essential to the ability to perceive the world, either internal or external (though, it could be argued, easily, that perception is essential to thought). In death, we can't access memories, think about our lives, engage in puzzles, or anything. There is nothing.

The third is desire. We all have things which we we want. The feeling of wanting something is desire. Though this requires thought, there are additional features to desire. Without desire, we can't love, hope, plan, go on adventures, or have fun (just to name a few things). Without desire, also, we can't experience heart-break, disappointment, frustration, or boredom. Death can't have these.

The final example of what death lacks is activity. Activity seems to require all of the previous (at least activity which you want to do requires desire). In death, we can't enact our plans, fulfill our desires, or do anything.

Next, we will move on to why Nagel thinks death is bad for the person who died. This is a common confusion for students. Nagel is arguing that death is bad for the person who died. 

\section{Part \thechapcount.\theseccount:  Is Death Bad for The Person Who Died?}\stepcounter{seccount}

As I have mentioned previously, if death is bad, then it must be because of what it lacks and according to Nagel, death lacks four (4) generic things, perception, thought, desire, and activity. So, the first question Nagel needs to grapple with is whether some state of affairs can be bad (for you) because it causes you to miss out on something (in the case of death, this would be those for things).

\thoughtex{Closed Pizzeria}{Suppose that you live in an apartment above the only pizzeria in town. At the end of the day, you get the left over pizza. But, one day, that pizzeria shuts down. When it shuts down, not only will you miss out on the good of having a place to live but you will also miss out on the good of the greasy nummy cheesy pizza.}{pizzeria.jpg}{A depressed looking man staring at an open pizza box.}

If you think that the pizza shop closing down was bad for you, then you are thinking like Nagel in this case (I could easily substitute different examples for this). The key thing is that there was a good which you were getting, and now you aren't, so you are worse off because of the lack. If you think about it, the four things which Nagel lists are the four things necessary for anything to be good for us. So if you die, you lose out on the good of the ability to experience goods. The lack of any potential good would seem to make death bad for the person who died. 

\subsection{The Problem and Nagel's Take}
The problem here is that the very same things which are required to experience good things in the world (perception, thought, desire, and activity) are the things which are required to experience bad. For example, without perception, you can't experience pain; without thought, you can't experience sadness; without desire, you can't experience disappointment; without activity, you can't experience frustration. There is no aspect of experience which can't be used to have a bad one. It would seem from this that there can be some sets of experiences, with no end in sight, which can make life not worth living any longer. For example, a person with a terminal illness slowly and painfully killing them. How much painful or bad experiences does a person have to go through before life is not worth living any longer?

Nagel has a reply to this because he wants to say that death is always bad for the person who died. He says that experience itself is always good. You can think of experience itself as a variable value which is always just enough to make the value of some experience better than not having any at all. Yes, some experiences are better than others, but any experience is better than not having any at all, according to Nagel. Some of you might have heard the phrase ``well, at least you experienced it", that's the sort of intuition Nagel is coming from. So, in response to the terminal illness, Nagel would say that it's a bad experience, but it's better than having none at all.  

\section{Part \thechapcount.\theseccount:  The Three Objections to Nagel's Stance}\stepcounter{seccount}
Nagel now turns to discussing the objections others might have to his stance. This is fairly common practice in Philosophy. You look at your position and think about how another person, on the opposing side, would combat or object to your thinking. Doing this makes you prepared for the upcoming debate. Nagel seeks to handle three possible objections to his stance. This is the bulk of the paper and the main topic for the homework this module.

\subsection{Can anything be bad for a person if it is not unpleasant to them?}

Another way of phrasing this would be to ask whether there's anything which is bad merely because we miss out on it or are ignorant of it. This line of thinking is akin to the phrase ``what you don't know can't hurt you." Nagel says that there is a problem with this objection, namely, that if it applies to death, then it must also apply to other supposed evils as well. Nagel gives a few different counter examples, and only mentions this one in passing, but it's a great example worthy of being expanded. 

\thoughtex{Bob vs Dave}{Suppose that you have two people, Bob and Dave. From their first person perspective, they can't tell the difference, but there is a difference. Bob is cheated on by their spouse, ridiculed by their friends, hated by their neighbors, and disparaged by their coworkers. Bob never is given a hint about this happening behind their back. Dave, on the other hand, has a faithful and loving spouse, their friends really do like them, their neighbors are sincere, and their coworkers actually respect them.}{bobvdave.jpg}{Two seemingly idential people, one with a depressing life while the other has a pleasant one, they cannot tell the difference.}

Who has the better life? Dave or Bob? One way to compare them is to ask yourself which life you would like to enter into, would you rather be reborn as Bob or Dave? If you think that Dave has the better life, then you are agreeing with Nagel that there are certain things which are bad for you despite you never actually experiencing them.  

\subsection{When we are dead, we don’t experience anything, there is no subject of the experience, so can we say that anything is good or bad for a person when they don’t exist?}

Another way to put this is to say that, in the Bob vs Dave case, there is a subject, a person, who is actually missing out, but when a person is dead, there is no person to miss out on it, they don't exist. There are many things out there which require us to experience them in order for them to be either good or bad to us, for example, pleasure and pain. This, however, is not true for all cases. For example, a person who cheerfully pursues a life trying to communicate with plants would be wasting their time and we could call that a worse life than someone who cheerfully spends their life pursuing a vaccine for the common cold, even if neither actually succeed. The experience of the individuals in both of these cases is relatively the same, but there seems to be a difference in which life we would like to enter.   

If we think that all goods and evils must be assigned to a person and experience, then we will have a hard time figuring out what makes things bad. Loss, betrayal, deception, and ridicule are on this view bad because people suffer when they learn of them. But it should be asked how our ideas of human value would have to be constituted to accommodate these cases directly instead. This view has the idea that we hurt when we learn of these things because they are bad, not that they are bad because we hurt when we learn of them.

\subsection{Why is there a difference between nonexistence prior to birth and nonexistence after birth?}

In this case, we say that the time we missed out on prior to our birth isn't good or bad (as in, we don't say that we missed out on that stuff) but the nonexistence after our death is something bad, according to Nagel. Some may think that this is a strange distinction. In both cases, we don't exist and there are experiences which we could have had, so what is with the difference? Nagel has two different replies to this and they work together to refute this. First, Nagel contends that prior to a person's birth, if they are born at all, they aren't missing out on anything. This is because, with the exception of some premature births, a person born at a substantially different time, even with the same parents, would not be you. This is a stance which waxes and wanes in popularity called Origin Essentialism. Essentially, you could not have had different biological parents and could only come from that particular sperm and egg. Since that particular set up could have only happened in a particular way and at a particular time, there is very little you could have missed out on prior to your birth. The second part of this is a response to the Roman philosopher and poet Titus Lucretius Carus, better known as Lucretius. Lucretius makes the following claim, put into more ordinary words: Prior to your birth, you don't exist and after your death, you don't exist; so it's irrational to fear death just as much as it is irrational to fear the time prior to your birth. Lucretius, in this piece, is claiming that there is a temporal symmetry between these two times, so we should think of them the same. Nagel, however, thinks that it's asymmetrical, they aren't the same. The time after your death is time that you could have had, experiences which you could have experienced. You aren't missing out on the time prior to your birth but you are missing out on the time after your death.